---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.path = "man/figures/README-", out.width = "80%")
me_ok <- rlang::is_installed("marginaleffects")
```

# *bartisan*: Generalized Bayesian Additive Regression Trees

<!-- badges: start -->
<!-- badges: end -->

## Overview

*bartisan* fits Bayesian additive regression trees (BART) for likelihoods
outside the conditionally conjugate Gaussian case, which standard BART
implementations cannot reach. The primary function is `bartisan()`, and its
interface follows that of `glm()`: a model is specified with a formula, a data
frame, and a family, with the forest taking the place of the linear predictor, so
nothing has to be said about which terms enter the model, which are curved, or
which interact.

Ordinary `family` objects are used unchanged, including their
links, so moving a model from `glm()` to `bartisan()` is a one-word change.
Supported families include binomial, Poisson, negative binomial, gamma, ordinal,
multinomial, zero-inflated counts, beta and ordered beta, a Tweedie compound
Poisson for a non-negative response with a point mass at zero, three accelerated
failure time models and a discrete proportional hazards model for right-censored
data, location-scale regression for a variance that varies with the predictors, a
Dirichlet process mixture for the error distribution itself, and a likelihood
written as an R function. Decision rules may be hard, as in standard BART, or
soft, as in the SoftBart model of Linero and Yang (2018), which yields smoother
fits. A fitted model is read through the packages that already do that work:
[*marginaleffects*](https://CRAN.R-project.org/package=marginaleffects) for
counterfactual estimands with posterior intervals,
[*loo*](https://CRAN.R-project.org/package=loo) for model comparison,
[*bayesplot*](https://CRAN.R-project.org/package=bayesplot) and
[*posterior*](https://CRAN.R-project.org/package=posterior) for convergence
diagnostics, and
[*performance*](https://CRAN.R-project.org/package=performance) for fit
statistics.

Check out the *bartisan* [website](https://ngreifer.github.io/bartisan/)!

## Installation

*bartisan* is not yet on CRAN. You can install the development version from
[GitHub](https://github.com/ngreifer/bartisan) with:

```r
# install.packages("pak")
pak::pak("ngreifer/bartisan")
```

Installation compiles C++ code, so it requires a C++17 toolchain along with
*Rcpp* and *RcppArmadillo*.

## Examples

### Right heart catheterization in critically ill patients

The `rhc` dataset holds 1500 patients from the SUPPORT study, with an indicator
for whether each received right heart catheterization within 24 hours of
admission to an intensive care unit, whether they died during follow-up, how long
that took, and thirteen physiological covariates recorded before catheterization
(Connors et al., 1996). Catheterization was not randomized, so the comparison is
confounded by how sick each patient was on admission, and the covariates are what
adjustment has to work with.

```{r}
library(bartisan)

data("rhc")
set.seed(123)

fit <- bartisan(death ~ . - days, data = rhc,
                family = binomial("probit"))

fit
```

`summary()` reports how often each predictor was split on, which is the forest's
account of what it used:

```{r}
summary(fit)
```

A forest has no coefficients, so the effect of catheterization is a contrast
between what the model predicts under one value of the treatment and under
another, averaged over the covariate distribution in the sample. That is
g-computation, and *marginaleffects* performs it on a fit directly, reporting a
posterior interval with the estimate:

```{r, eval = me_ok}
marginaleffects::avg_comparisons(fit, variables = "rhc")
```

The same event as a right-censored survival outcome, given as a `Surv` object on
the left-hand side of the formula:

```r
bartisan(survival::Surv(days, death) ~ . , data = rhc,
         family = lognormal_aft())
```

Prediction returns the full posterior rather than a point estimate, since every
draw of every tree is retained:

```{r}
draws <- predict(fit, newdata = rhc[1:5, ], type = "response", draws = TRUE)

apply(draws, 2, quantile, c(.025, .975))
```

`vignette("bartisan")` walks through a complete analysis, and the vignettes it
links to take the pieces on their own: `vignette("families")` for choosing a
likelihood, `vignette("survival")` for censored responses, `vignette("effects")`
for reading the fitted function, `vignette("diagnostics")` for convergence and
fit, `vignette("importance")` for which predictors matter,
`vignette("comparison")` for choosing between models, and `vignette("causal")`
for causal inference.

## Comparison With Other BART Packages

Checked against the installed versions of each package rather than from
memory: *dbarts* 0.9.34, *BART* 2.9.10, *flexBART* 2.0.3, *SoftBart*
1.0.3, *bartMachine* 1.4.2, and *stochtree* 0.4.5. A dash means the
package does not offer the feature, not that it fits it badly.

|  | bartisan | dbarts | BART | flexBART | SoftBart | bartMachine | stochtree |
|----|:--:|:--:|:--:|:--:|:--:|:--:|:--:|
| **Responses** |  |  |  |  |  |  |  |
| Gaussian | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ |
| Binary, probit | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ |
| Binary, logit | ✓ | — | ✓ | ✓ | — | — | — |
| Count: Poisson, negative binomial | ✓ | — | — | — | — | — | — |
| Gamma, beta, ordered beta, Tweedie | ✓ | — | — | — | — | — | — |
| Ordinal | ✓ 3 links | — | — | — | — | — | ✓ cloglog |
| Multinomial | ✓ logit, probit | — | ✓ | — | — | — | — |
| Zero-inflated counts | ✓ | — | — | — | — | — | — |
| Survival | ✓ 3 AFT, PH | — | ✓ **+ recurrent, competing risks** | — | — | — | — |
| Heteroskedastic | ✓ | — | — | ✓ | — | — | ✓ |
| Error distribution itself modeled | ✓ | — | — | — | — | — | — |
| A likelihood written by the user | ✓ | — | — | — | — | — | — |
| **Rules and priors** |  |  |  |  |  |  |  |
| Hard decision rules | ✓ | ✓ | ✓ | ✓ | — | ✓ | ✓ |
| Soft decision rules | ✓ 4 gates | — | — | — | ✓ | — | — |
| Dirichlet sparsity prior (DART) | ✓ | — | ✓ | ✓ | ✓ | — | — |
| Splitting weights fixed by the user | ✓ | — | — | — | ✓ groups | ✓ | — |
| Categorical splits on level subsets | ✓ | — | — | ✓ **+ nested, network** | — | — | — |
| Missing predictors, no imputation | ✓ default | — | — | — | — | ✓ | — |
| **Structure** |  |  |  |  |  |  |  |
| Random intercepts | ✓ | ✓ | — | — | — | — | ✓ |
| Varying coefficients | ✓ | — | — | ✓ | ✓ | — | ✓ |
| Treatment-effect (BCF) structure | ✓ | — | — | — | — | — | ✓ |
| Formula interface | ✓ | ✓ | — | ✓ | — | — | — |
| **Running it** |  |  |  |  |  |  |  |
| Several chains | ✓ | ✓ | ✓ | ✓ | — | ✓ | ✓ |
| Threads inside one chain | — | ✓ | ✓ | — | — | ✓ | ✓ |
| Grow-from-root warm start | — | — | — | — | — | — | ✓ |
| Cross-validation over hyperparameters | — | ✓ | — | — | — | ✓ | — |
| Save and reload a fitted model | RDS | — | — | — | — | ✓ | ✓ JSON |
| **Reading the fit** |  |  |  |  |  |  |  |
| `predict()` on new data | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ |
| Convergence diagnostics built in | ✓ | — | ✓ | — | — | ✓ | — |
| Variable importance | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ | — |
| Formal variable-selection test | ✓ inclusion | — | ✓ permutation | — | ✓ inclusion | ✓ permutation | — |
| Partial dependence | ✓ | ✓ | — | — | ✓ | ✓ | — |
| Interaction detection | ✎ | — | — | — | — | ✓ | — |
| Counterfactual estimands with intervals | ✓ | — | — | — | — | — | — |
| Cross-validated model comparison | ✓ | — | — | — | — | — | — |

**✎ means a helper package covers it, not this one.** Interaction
detection is the one row left marked, and *bartisan* does it through
*marginaleffects*, because a fit works with it and every estimand there
is computed by pushing the draws through:
`avg_comparisons(variables = "x", by = "z")` with `hypotheses(~pairwise)`
is interaction detection and needs no code here.

The variable-selection row divides into two things that both get called a
test. One is the posterior inclusion probability under the sparsity
prior, which `variable_importance()` reports as `prop_used` and
*SoftBart* reports as `posterior_probs()`; the other is a permutation
test, which *bartMachine* and *BART* offer and this package does not.
See `vignette("implementation")`.

The other columns worth reading as gaps rather than as differences are
these. There are **no threads inside a chain**, so a single chain is
single-core here where *dbarts*, *BART*, *bartMachine*, and *stochtree*
all use several; chains do run in parallel, which is the cheaper win,
but it does not help one chain. There is **no cross-validation** over
`k`, `num_trees`, and the tree prior, which `dbarts::xbart()` and
`bartMachine::bartMachineCV()` both automate. There is **no
grow-from-root warm start**, which is *stochtree*’s way of shortening
burn-in. And there is **no JSON serialization**, so a fit round-trips
through `saveRDS()` and not into another language. On survival, *BART*
is ahead: recurrent events and competing risks are there and here they
are not.

Three more packages are single-purpose rather than general, so they are
not columns above: [*bcf*](https://CRAN.R-project.org/package=bcf) fits
Bayesian causal forests only,
[*VCBART*](https://github.com/skdeshpande91/VCBART) varying-coefficient
models only, and *bartCause* wraps *dbarts* for causal estimands.

## Citing *bartisan*

*bartisan* implements the sampler of Linero (2025), and its MCMC engine is
adapted from that paper's `FlexBart` reference implementation. Please cite both
the method and the package, the latter with its version number, which
`citation("bartisan")` supplies:

Linero, A. R. (2025). Generalized Bayesian additive regression trees models:
beyond conditional conjugacy. *Journal of the American Statistical Association*,
120(549), 356–369. <https://doi.org/10.1080/01621459.2024.2337156>

Linero, A. R., & Yang, Y. (2018). Bayesian regression tree ensembles that adapt
to smoothness and sparsity. *Journal of the Royal Statistical Society Series B*,
80(5), 1087–1110. <https://doi.org/10.1111/rssb.12293>

## Questions and Bug Reports

Please file an issue at
<https://github.com/ngreifer/bartisan/issues>, with a reproducible example where
the report concerns a fit.
