---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.path = "man/figures/README-", out.width = "80%")
options(digits = 4)

surv_ok <- rlang::is_installed("survival")
```

# *bartisan*: Generalized Bayesian Additive Regression Trees

<!-- badges: start -->
<!-- badges: end -->

## Overview

*bartisan* fits Bayesian additive regression trees (BART) for responses that
standard BART cannot reach, using the Laplace-approximation reversible-jump
sampler of Linero (2025), which lifts the requirement that the leaf parameters
be integrable in closed form. The primary function is `bartisan()`, and a model
is specified the way it is in `glm()`: a formula, a data frame, and a family,
with a forest in place of the linear predictor, so nothing has to be said about
which terms are curved or which interact. The `stats::family` objects `glm()`
takes are accepted unchanged, links included. The families that have no `glm()`
counterpart are supplied in the same style: negative binomial, ordinal,
multinomial, beta and ordered beta, zero-inflated counts, a Tweedie compound
Poisson, accelerated failure time and proportional hazards models for
right-censored times, location-scale regression, a Dirichlet process mixture for
the error distribution, and a log density written as an R function. Decision
rules are soft by default, in the manner of Linero and Yang (2018), which gives
smoother fits than the step functions of standard BART, and `gate = "hard"`
gives those back. Treatment effects come from `estimate_effect()`, and a fit is
otherwise read through the packages that already do that work:
[*marginaleffects*](https://CRAN.R-project.org/package=marginaleffects) for the
wider range of counterfactual estimands,
[*loo*](https://CRAN.R-project.org/package=loo) for model comparison,
[*bayesplot*](https://CRAN.R-project.org/package=bayesplot) and
[*posterior*](https://CRAN.R-project.org/package=posterior) for convergence
diagnostics, and
[*performance*](https://CRAN.R-project.org/package=performance) for fit
statistics.

Check out the *bartisan*
[website](https://ngreifer.github.io/bartisan/)!

## Installation

*bartisan* is not yet on CRAN. You can install the development version from
[GitHub](https://github.com/ngreifer/bartisan) with:

```r
# install.packages("pak")
pak::pak("ngreifer/bartisan")
```

Installation compiles C++, so it requires a C++17 toolchain along with *Rcpp*
and *RcppArmadillo*.

## Example

### Right heart catheterization in critically ill patients

`rhc` holds 1500 patients from the SUPPORT study, with an indicator for whether
each was given right heart catheterization within 24 hours of admission to an
intensive care unit, whether they died during follow-up, and how long that took
(Connors et al., 1996). Thirteen covariates recorded before the catheter went in
come with them, and `?rhc` lists them. Catheterization was not randomized, so
sicker patients were likelier to receive it, and those covariates are what
adjustment has to work with. Nothing in the model supplies the assumption that
they suffice; what it supplies is a flexible estimate of the outcome under each
treatment given the covariates, which is what g-computation then averages.

```{r}
library(bartisan)

data("rhc")
set.seed(123)

# Death as the outcome, with every variable but the timing of the same event
# as a candidate predictor. The family is the one `glm()` would be given
fit <- bartisan(death ~ . - days, data = rhc,
                family = binomial("probit"))

fit
```

The forest finds its own interactions and nonlinearity, so what it reports in
place of a coefficient table is how it spent its splitting rules:

```{r}
head(variable_importance(fit), 5)
```

Having no coefficients also means the effect of catheterization is a contrast
between what the model predicts with `rhc` set to one and with it set to zero,
averaged over the covariate distribution in the sample. That is g-computation,
and every posterior draw goes through it, so the interval is a credible
interval rather than a delta-method approximation:

```{r}
estimate_effect(fit, treat = "rhc")
```

The same event is also a right-censored survival time, and asking for it that
way changes the left-hand side of the formula and the family and nothing else.
Here the contrast is a ratio, so the estimate is the factor by which
catheterization multiplies expected survival:

```{r, eval = surv_ok}
sfit <- bartisan(survival::Surv(days, death) ~ ., data = rhc,
                 family = lognormal_aft())

estimate_effect(sfit, treat = "rhc", comparison = "ratio")
```

`vignette("bartisan")` works through a complete analysis, from fitting to
convergence checks to the estimates, and the other vignettes take those pieces
one at a time. `vignette("implementation")` sets out how the sampler works and
compares the package's features against *dbarts*, *BART*, *flexBART*,
*SoftBart*, *bartMachine*, and *stochtree*.

## Citing *bartisan*

*bartisan* implements the sampler of Linero (2025), and its MCMC engine is
adapted from that paper's `FlexBart` reference implementation. Please cite both
the method and the package, the latter with the version number, which
`citation("bartisan")` supplies:

Linero, A. R. (2025). Generalized Bayesian additive regression trees models:
beyond conditional conjugacy. *Journal of the American Statistical
Association*, 120(549), 356–369.
<https://doi.org/10.1080/01621459.2024.2337156>

Linero, A. R., & Yang, Y. (2018). Bayesian regression tree ensembles that adapt
to smoothness and sparsity. *Journal of the Royal Statistical Society Series
B*, 80(5), 1087–1110. <https://doi.org/10.1111/rssb.12293>

## Questions and Bug Reports

Please file an issue at <https://github.com/ngreifer/bartisan/issues>, with a
reproducible example where the report concerns a fit.
