---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.path = "man/figures/README-", out.width = "80%")
me_ok <- rlang::is_installed("marginaleffects")
```

# *bartisan*: Generalized Bayesian Additive Regression Trees

<!-- badges: start -->
<!-- badges: end -->

## Overview

*bartisan* fits Bayesian additive regression trees (BART) for likelihoods
outside the conditionally conjugate Gaussian case, which standard BART
implementations cannot reach. The primary function is `bartisan()`, and its
interface follows that of `glm()`: a model is specified with a formula, a data
frame, and a family, with the forest taking the place of the linear predictor, so
nothing has to be said about which terms enter the model, which are curved, or
which interact. Ordinary `family` objects are used unchanged, including their
links, so moving a model from `glm()` to `bartisan()` is a one-word change.
Supported families include binomial, Poisson, negative binomial, gamma, ordinal,
multinomial, zero-inflated counts, beta and ordered beta, three accelerated
failure time models and a discrete proportional hazards model for right-censored
data, location-scale regression for a variance that varies with the predictors, a
Dirichlet process mixture for the error distribution itself, and a likelihood
written as an R function. Decision rules may be hard, as in standard BART, or
soft, as in the SoftBart model of Linero and Yang (2018), which yields smoother
fits. A fitted model is read through the packages that already do that work:
[*marginaleffects*](https://CRAN.R-project.org/package=marginaleffects) for
counterfactual estimands with posterior intervals,
[*loo*](https://CRAN.R-project.org/package=loo) for model comparison,
[*bayesplot*](https://CRAN.R-project.org/package=bayesplot) and
[*posterior*](https://CRAN.R-project.org/package=posterior) for convergence
diagnostics, and
[*performance*](https://CRAN.R-project.org/package=performance) for fit
statistics.

Check out the *bartisan* [website](https://ngreifer.github.io/bartisan/)!

## Installation

*bartisan* is not yet on CRAN. You can install the development version from
[GitHub](https://github.com/ngreifer/bartisan) with:

```r
# install.packages("pak")
pak::pak("ngreifer/bartisan")
```

Installation compiles C++ code, so it requires a C++17 toolchain along with
*Rcpp* and *RcppArmadillo*.

## Examples

### Right heart catheterization in critically ill patients

The `rhc` dataset holds 1500 patients from the SUPPORT study, with an indicator
for whether each received right heart catheterization within 24 hours of
admission to an intensive care unit, whether they died during follow-up, how long
that took, and thirteen physiological covariates recorded before catheterization
(Connors et al., 1996). Catheterization was not randomized, so the comparison is
confounded by how sick each patient was on admission, and the covariates are what
adjustment has to work with. Two features make the data a useful illustration:
the covariates enter through a forest, so no functional form has to be committed
to, and the same event is available both as a binary indicator and as a
right-censored survival time, so it supports two families without changing the
call's shape.

```{r}
library(bartisan)

data("rhc")
set.seed(123)

# `days` is the timing of the same event as `death`, so it is excluded
# rather than conditioned on. The family is read off the response
fit <- bartisan(death ~ . - days, data = rhc,
                num_trees = 20, num_burn = 200, num_draws = 200)

fit
```

`summary()` reports how often each predictor was split on, which is the forest's
account of what it used:

```{r}
summary(fit)
```

A forest has no coefficients, so the effect of catheterization is a contrast
between what the model predicts under one value of the treatment and under
another, averaged over the covariate distribution in the sample. That is
g-computation, and *marginaleffects* performs it on a fit directly, reporting a
posterior interval with the estimate:

```{r, eval = me_ok}
marginaleffects::avg_comparisons(fit, variables = "rhc")
```

The same event as a right-censored survival outcome, given as a `Surv` object on
the left-hand side of the formula:

```r
bartisan(survival::Surv(days, death) ~ . , data = rhc,
         family = lognormal_aft())
```

Prediction returns the full posterior rather than a point estimate, since every
draw of every tree is retained:

```{r}
draws <- predict(fit, newdata = rhc[1:5, ], type = "response", draws = TRUE)

apply(draws, 2, quantile, c(.025, .975))
```

`vignette("bartisan")` walks through a complete analysis, and the vignettes it
links to take the pieces on their own: `vignette("families")` for choosing a
likelihood, `vignette("survival")` for censored responses, `vignette("effects")`
for reading the fitted function, `vignette("diagnostics")` for convergence and
fit, `vignette("importance")` for which predictors matter,
`vignette("comparison")` for choosing between models, and `vignette("causal")`
for causal inference.

## Citing *bartisan*

*bartisan* implements the sampler of Linero (2025), and its MCMC engine is
adapted from that paper's `FlexBart` reference implementation. Please cite both
the method and the package, the latter with its version number, which
`citation("bartisan")` supplies:

Linero, A. R. (2025). Generalized Bayesian additive regression trees models:
beyond conditional conjugacy. *Journal of the American Statistical Association*,
120(549), 356–369. <https://doi.org/10.1080/01621459.2024.2337156>

Linero, A. R., & Yang, Y. (2018). Bayesian regression tree ensembles that adapt
to smoothness and sparsity. *Journal of the Royal Statistical Society Series B*,
80(5), 1087–1110. <https://doi.org/10.1111/rssb.12293>

## Questions and Bug Reports

Please file an issue at
<https://github.com/ngreifer/bartisan/issues>, with a reproducible example where
the report concerns a fit.
