---
output: github_document
bibliography: vignettes/references.bib
link-citations: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.path = "man/figures/README-", out.width = "90%",
                      fig.width = 7, fig.height = 3.4, dpi = 130,
                      fig.align = "center")
options(digits = 4)

if (rlang::is_installed("future")) {
  future::plan(future::multisession)
}

if (interactive() && rlang::is_installed("progressr")) {
  progressr::handlers(global = TRUE)
}
```

# *bartisan*: Generalized Bayesian Additive Regression Trees

<!-- badges: start -->
<!-- badges: end -->

## Overview

Bayesian additive regression trees (BART) estimate a regression function as a
sum of small trees, so nonlinearity and interactions are found rather than
specified. *bartisan* does that for response distributions that standard BART
cannot reach, using the Laplace-approximation reversible-jump sampler of
@linero2025, which relaxes the requirement that the leaf parameters be
integrable in closed form. That requirement ties standard BART to a Gaussian
likelihood, and lifting it is the sense in which the model here is
*generalized*: the same sense a generalized linear model is, in which the
response distribution is a choice the analyst makes rather than an assumption
the sampler imposes.

A model is written the way it is in `glm()`, with a formula, a data
frame, and a family, and the `stats::family` objects `glm()` takes are accepted
unchanged. Beyond them come the families that have no `glm()` counterpart:
negative binomial, ordinal, multinomial, beta and ordered beta, zero-inflated
counts, a Tweedie compound Poisson, accelerated failure time and proportional
hazards models for right-censored times, location-scale regression, a Dirichlet
process mixture for the error distribution, and a log density written as an R
function. Decision rules are smooth by default, in the manner of @linero2018,
which fits a smoother function than the step functions of standard BART.

*bartisan* supports varying-coefficient models, where the effect of a predictor receives its own forest, a special case of which is the Bayesian causal forest (BCF) model for estimating treatment effects. Along with this is an interface for estimating average treatment effects from BART and BCF models. In addition, *bartisan* supports random intercepts BART models.

Integration is provided with the [*future*](https://CRAN.R-project.org/package=future) and [*progressr*](https://CRAN.R-project.org/package=progressr) packages for parallel processing and progress bars.

## Installation

*bartisan* is not yet on CRAN. You can install the development version from
[GitHub](https://github.com/ngreifer/bartisan) with:

```r
# install.packages("pak")
pak::pak("ngreifer/bartisan")
```

Installation compiles C++, so it needs a C++17 toolchain.

## Example

`rhc` holds 1500 critically ill patients from the SUPPORT study, recording
whether each was given right heart catheterization on admission to intensive
care, whether they died during follow-up, and thirteen covariates measured
beforehand [@connors1996]. Fitting a model to it takes one call, and the
family is the one `glm()` would be given:

```{r fit}
library(bartisan)

data("rhc")
set.seed(123)

fit <- bartisan(death ~ rhc + age + sex + race + edu + pafi + 
                  paco2 + crea + surv2m + card,
                data = rhc, family = binomial())

fit
```

Calling `summary()` on the fit displays a measure of variable importance, how often predictors were used to split the trees:

```{r}
summary(fit)
```

Nothing had to be said about which predictors matter, which are curved, or which
interact. `plot()` shows what the model made of one of them, here the study's own
estimate of each patient's chance of surviving two months:

```{r pdp, fig.alt = "Fitted probability of death against the study's two-month survival estimate. The curve sits flat near .87 up to about .3, falls steeply from there to about .75, and flattens again near .48, inside a credible band that widens at both ends."}
partial_dependence(fit, ~ surv2m) |>
  plot() +
  ggplot2::labs(x = "Estimated probability of surviving two months",
                y = "Fitted probability of death")
```

A forest has no coefficients, so an effect is a contrast between what the model
predicts under one value of a predictor and under another, averaged over the
sample. Every posterior draw goes through that contrast, so what comes back is a
credible interval rather than a standard error:

```{r effect}
estimate_effect(fit, treat = "rhc")
```

To examine effect heterogeneity, one can generate a plot of conditional effect estimates for all units:

```{r}
estimate_effect(fit, treat = "rhc", estimand = "CATE") |>
  plot()
```


## Learning More

`vignette("bartisan")` is the place to start. It carries one analysis from the
formula through to the estimates, and each of its sections points at the guide
that takes that step further:

| Topic | Vignette |
| --- | --- |
| Choosing a response family | `vignette("families")` |
| Censored and survival outcomes | `vignette("survival")` |
| Varying coefficient models | `vignette("varying")` |
| Effects, curves, and interactions | `vignette("effects")` |
| Which predictors the model uses | `vignette("importance")` |
| Convergence and model fit | `vignette("diagnostics")` |
| Comparing and selecting models | `vignette("comparison")` |
| Causal inference and Bayesian causal forests | `vignette("causal")` |
| What the model is and how it is fitted | `vignette("implementation")` |
| Short answers to common questions | `vignette("faq")` |

A fitted model also works with the packages that already read Bayesian fits:
[*marginaleffects*](https://CRAN.R-project.org/package=marginaleffects) for a
wider range of estimands,
[*loo*](https://CRAN.R-project.org/package=loo) for model comparison,
[*bayesplot*](https://CRAN.R-project.org/package=bayesplot) and
[*posterior*](https://CRAN.R-project.org/package=posterior) for the draws
themselves, and
[*performance*](https://CRAN.R-project.org/package=performance) for fit
statistics. Load one and call its functions on the fit; there is nothing to set
up.

## Citing *bartisan*

*bartisan* implements the sampler of @linero2025, and its MCMC engine is
adapted from that paper's `FlexBart` reference implementation. Please cite the
method and the package both, the latter with its version number, which
`citation("bartisan")` supplies. The smooth decision rules are those of
@linero2018. Both papers are listed under References below.

## Questions and Bug Reports

Please file an issue at <https://github.com/ngreifer/bartisan/issues>, with a
reproducible example where the report concerns a fit.

## References
