---
title: "Getting Started with bartisan"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Getting Started with bartisan}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.2, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 4, marginaleffects_safe = FALSE)

has_me <- rlang::is_installed("marginaleffects")
has_loo <- rlang::is_installed("loo")
has_bp <- rlang::is_installed("bayesplot")
```

## Introduction

*bartisan* fits Bayesian additive regression trees (BART) using the same
interface as `glm()`. We supply a formula, a data frame, and a family, and the
model estimates the relationship between the predictors and the outcome without
anything being said about what shape that relationship takes. Nonlinearity and
interactions are found rather than specified.

In this guide we will work through a complete analysis. First we'll fit a model
and check both that the sampler converged and that the fit describes the data.
Next we'll see which predictors the forest uses, read the effects off it, and
predict for new observations. Finally we'll compare two models and set out what
the fit does not do. The guide assumes a reader comfortable with regression and
assumes nothing about machine learning or Bayesian methods.

The main thing to take from it is that the defaults are meant to be used. The
priors, the number of trees, and the sampler settings are chosen to work across a
wide range of problems, and tuning them is rarely where the gains are. Almost
everything below is a single function call with no arguments beyond the formula
and the data.

Most sections point to a vignette that covers the same ground in more depth.

```{r libs}
library(bartisan)
```

### Parallelization and Progress Bars

Using parallelization can speed up fitting `bartisan()` models with multiple chains. Parallelization is controlled through the [*future*](https://CRAN.R-project.org/package=future) package. To request multi-session evaluation, one can simply call the following:

```{r, eval = FALSE}
future::plan("multisession")
```

The [*future.apply*](https://CRAN.R-project.org/package=future.apply) package must also be installed (installing it also installs *future*).

Some operations in *bartisan* can be take some time, and one can request a progress bar using the [*progressr*](https://CRAN.R-project.org/package=progressr) package. One can request progress bars globally for all functions that produce them using the following:

```{r, eval = FALSE}
progressr::handlers(global = TRUE)
```

To request a progress bar for a single call, wrap the function evaluation in `progressr::with_progress()`, e.g.,

```{r, eval = FALSE}
progressr::with_progress(
  bartisan(y ~ ., data = d)
)
```

## The Data

`rhc` records a random 1500 of the 5735 patients in the SUPPORT study and whether
each received right heart catheterization, a monitoring procedure, within a day
of arriving in intensive care [@connors1996]. The question the study asked is
whether the procedure helps or harms.

```{r data}
data(rhc)

str(rhc)
```

The outcome comes in two forms. `death` is whether the patient died during
follow-up, and `days` is how long that took. This vignette uses the binary form;
`vignette("survival")` uses the other.

The remaining variables are the patient's age, sex, race and years of education,
seven physiological measurements taken on the first day, whether cardiovascular
disease was among the diagnoses, and `surv2m`, the study's own estimate of the
patient's chance of surviving two months. All were recorded before
catheterization.


## Fitting the Model

To fit the model, we simply call `bartisan::bartisan()` with the same model formula one would use for `glm()` or any other standard model-fitting function, additionally supplying the dataset to `data` and the response family to `family`. Because BART involves random processes, we must also set a seed using `set.seed()` to ensure reproducibility (though not the `kind` does not matter).

```{r fit}
set.seed(2026)

fit <- bartisan(
  death ~ rhc + age + sex + race + edu + aps + meanbp + resp + hema +
    pafi + paco2 + crea + surv2m + card,
  data = rhc, family = binomial(), chains = 4
)

fit
```

`family = binomial()` says the outcome is binary, exactly
as in `glm()`. The family may be omitted, in which case it is read off the
outcome and reported; naming it is clearer and silences the message.

`chains = 4` runs the sampler four times from different starting points. The
default is one chain, and the diagnostics in the next section are still
computed for it, with R-hat taken by splitting that single chain into
segments; running several is what lets them detect chains that have settled in
different places.

Choosing a family is the one modeling decision that usually matters more than any
sampler setting. `vignette("families")` covers the choice, and
`vignette("survival")` covers censored outcomes such as this one's `days`.

The other decision worth knowing about is the shape of the decision rules. By
default a rule is soft, so an observation near a split contributes to both sides
of it and the fitted function comes out smooth rather than piecewise constant;
`gate` in `bartisan_control()` switches to the hard rules of standard BART,
which fit faster but less accurately. `vignette("implementation")` has the
comparison.

## Checking the Model

In a Bayesian analysis, one must first determine whether the sampler converged before moving forward with a model's results. Separately, one should assess whether the model is a good fit to the data.

### Convergence

`diagnose()` computes the convergence and mixing
statistics, applies the conventional thresholds, and says what to change about
whichever of them fall short.

```{r diagnose}
diagnose(fit)
```

`rhat` compares variation between chains to variation within them; values near 1
indicate the chains have settled on the same answer. `rhat_late` is the same
statistic on the second half of the draws alone, which is what distinguishes a
warmup that ended too early from chains that have each settled somewhere
different. `ess_bulk` and `ess_tail` are effective sample sizes, and count how
many independent draws the correlated ones are worth, in the middle of the
distribution and in the tails.

Read the rows that correspond to the quantities being reported. `eta.eta` is the
fitted function and appears twice, once averaged over the observations and once
over the worst 5% of them; the second is what governs a prediction for one
observation. Forests mix slowly on their fitted values, so a figure above 1.01 on
the worst-5% row is ordinary rather than alarming; `vignette("diagnostics")`
explains what to do about it and when to worry.

No row here governs an effect, though, and the average row in particular should
not be read as if it did. An effect is a contrast, and a contrast can mix badly
where the function it is a contrast of mixes well. `diagnose()` takes the
estimand itself when there is one to take:

```{r diagnose_effect}
diagnose(estimate_effect(fit, treat = "rhc"))
```

Far fewer effective draws than the table above would suggest, and the last check
says why: the splitting prior gives `rhc` no rule in some draws, which puts the
contrast at exactly zero and can hold it there for a long run. Nothing in the
fit's own table shows it, because the other predictors keep the fitted function
moving the whole time. `vignette("causal")` covers the settings that remove the
atom when an effect is what is being reported.

The table is in `diagnose(fit)$table` when the numbers are wanted rather than
the report.

### Fit

The question worth asking of a binary outcome is whether the predicted
probabilities mean what they say: among the patients the model gave a 30% chance
of dying, did about 30% die? A calibration plot answers it.

```{r ppcheck, eval = has_bp && has_loo, fig.height = 3.0}
bayesplot::pp_check(fit, type = "loo_calibration")
```

The line should follow the diagonal, and here it does across the whole range.
Each patient is judged against a probability estimated without them, so this is
not the optimistic in-sample reading.

The default `pp_check()` compares the distribution of simulated outcomes with the
observed distribution, which is the check to reach for when the outcome is
continuous; with two values to get right it finds nothing here.
`vignette("diagnostics")` covers what a departure from the diagonal looks like
and what else to check.

## Variable Importance

A `bartisan()` fit produces no model coefficients; to see which covariates played the biggest role in the fit, we can use `variable_importance()` on its output. This produces statistics about the use of each predictor in the trees.

```{r varimp}
variable_importance(fit)
```

`splits` is the average number of splitting rules the forest spends on each
predictor per draw, and `prop_used` is the proportion of draws in which the
predictor received any rule at all.

`surv2m` takes the most rules, which is unsurprising: it is a prognostic score
built to predict survival. `age` follows. At the bottom, `sex`, `race` and
several of the physiological measurements are used in little more than half of
the draws, which says the model can often do without them.

It's important to remember that usage is not effect size: a predictor can be split on constantly
and still move the prediction very little, and the comparisons in the next
section are the better guide to that. Also, when predictors are correlated, the usage
distributes among them more or less arbitrarily.

`vignette("importance")` covers variable importance and selection, including how
to tell whether a difference in this table means anything.

## Interpreting the Fit

A forest has no table of coefficients to read, so a fit is interpreted by
putting questions to it: what happens to the prediction when a predictor is
changed, and what shape does the prediction trace as that predictor varies.
The functions below answer versions of that question, and they differ in what
they average over rather than in what they are asking.

### A Table of Average Comparisons

`marginaleffects::avg_comparisons()` is the closest thing to a table of
regression slopes. For each predictor in turn it changes that predictor, leaves
the others as they are, predicts every patient twice, and averages the
difference over the sample.

```{r comparisons, eval = has_me}
marginaleffects::avg_comparisons(fit)
```

`Contrast` says what change was made.^[By default *marginaleffects* reports the
posterior median as the point estimate, whereas `predict()` and
`estimate_effect()` report the posterior mean, so the same quantity can differ
slightly between them. Setting `options(marginaleffects_posterior_center = mean)`
before the call switches it to the mean; we do this in `vignette("causal")`.]
For a categorical predictor it is a difference between two levels, and for a
numeric one it is an increase of one unit, which is a default rather than
anything the data suggested. The comparisons are on the probability scale, so
`rhc` reads as an increase of a little over five percentage points in the
probability of death and `age` as three tenths of a point per year of age. A
logistic regression would report each of these as one slope on the log-odds
scale; these are averages over the sample of a quantity the model allows to
differ from patient to patient.

Several rows are exactly zero rather than merely small, which is worth knowing
about. The default splitting prior can leave a predictor out of the forest
altogether in a given draw, and in such a draw every comparison involving it is
exactly zero, so the posterior has a point mass there. Where that mass covers
the middle of the posterior the median falls inside it and the estimate prints
as exactly zero, which is what has happened to the predictors at the bottom of
the importance table. The same mass is why the lower bound for `rhc` is zero
rather than merely close to it.

The one-unit default deserves a second look whenever a predictor does not span a
unit. `surv2m` is a probability, so an increase of one is wider than its whole
observed range, and the model holds its prediction flat past the edge of that
range rather than continuing any trend. Asking instead for a change the data
contains gives a larger answer:

```{r comparisons_iqr, eval = has_me}
marginaleffects::avg_comparisons(fit, variables = list(surv2m = "iqr"))
```

Moving a patient from the first quartile of the prognostic score to the third
lowers the predicted probability of death by about .29, against the .18 the
one-unit contrast reported.

### The Shape of a Relationship

One number for a predictor hides the shape of the relationship behind it, and
the two `surv2m` contrasts are what that looks like when the shape matters: the
answer depends on which change is asked about. Plotting the fit against one
predictor shows the whole curve, averaging over the other predictors at each
value.

```{r pdp, fig.height = 3.0}
plot(fit, ~ surv2m) +
  ggplot2::labs(x = "Estimated probability of surviving two months",
                y = "Fitted probability of death")
```

The fitted probability of death falls from close to .9 to about .5 as the
prognostic score rises, and the fall is not a straight line, which is why the
two contrasts above disagree about its size. A logistic regression reports one
slope on the log-odds scale for the whole range. Nothing had to be specified to
find the shape.

The band is a credible interval on the *average* prediction at each value, not
on any one patient's, and it widens at the top where few patients were that
healthy. `partial_dependence()` returns the same numbers without drawing them,
and `marginaleffects::plot_predictions()` is the one to reach for when the grid
or what is conditioned on needs more control.

`vignette("effects")` covers effects, curves, and interactions.

### The Effect of a Treatment

`rhc` has a row in the table above, so in one sense its effect has already been
reported. `estimate_effect()` asks the same question in the vocabulary of a
causal analysis: it takes the treatment as an argument rather than as one
predictor among many, it names the estimand being averaged, and it reports the
two potential outcomes the contrast is a difference of.

```{r effects}
eff <- estimate_effect(fit, treat = "rhc")

eff
```

Catheterization is associated with an increase of about five percentage points
in the probability of death. The interval runs from zero to roughly eleven
points, so the direction is reasonably clear and the size is not. The estimate
differs a little from the `rhc` row of the comparisons table because it is the
posterior mean of the same draws rather than their median.

Because the outcome is binary, this is a difference in probability, which is
interpretable without reference to the model, and it is usually the number to
report. The two probabilities it is a difference of are printed below it: about
63% of patients would be expected to die without catheterization and 69% with
it, averaging over the covariates as they actually occur in this sample.

The treatment has to be categorical, since the contrast is taken between its
levels. The effect of a continuous treatment is a slope or a dose-response curve
instead, and `estimate_effect()` says so and names the functions that produce
one.

The defaults are not necessarily the best settings for estimating a causal
effect, and the zero at the end of the interval is the first sign of it: a
sparsity prior that can drop a predictor is reasonable for prediction and poor
for a treatment whose effect is being reported. More specialized methods, such
as Bayesian causal forests (BCF) and BART without a sparsity-inducing prior, are
described in `vignette("causal")`, which also covers what has to be true of the
data before any of this can be read as an effect of the procedure.

## Predicting New Observations

`predict()` returns the posterior mean prediction.

```{r pred}
new_patient <- rhc[1, ]
new_patient$rhc <- 1

predict(fit, newdata = new_patient)
```

For a prediction with an interval, use `marginaleffects::predictions()`, which
reports the posterior median rather than the mean `predict()` reports:

```{r predint, eval = has_me}
marginaleffects::predictions(fit, newdata = new_patient)
```

This describes the probability that a patient with these characteristics dies. It
is an interval for that probability, not a statement about which way any
individual patient will go: the outcome itself is either 0 or 1, and a
probability of .7 is entirely compatible with survival.

## Comparing Models

Approximate leave-one-out cross-validation estimates how well a model predicts
data it has not seen.

```{r loo, eval = has_loo}
library(loo)

loo(fit)
```

`elpd_loo` is the estimated log predictive density on held-out data, where higher
is better. `p_loo` is the effective number of parameters, which is a measure of
how much of the data the forest is actually using. The Pareto k diagnostics are
all good, meaning the approximation is trustworthy for this fit.

Two models can be compared directly. Here we ask whether everything recorded
about how sick the patient was on arrival, the seven physiological measurements
and the prognostic score among them, earns its keep over the demographics
alone:

```{r loocompare, eval = has_loo}
set.seed(2026)
demographics <- bartisan(death ~ rhc + age + sex + race + edu, data = rhc,
                         family = binomial(), chains = 4)

loo_compare(list(full = loo(fit),
                 demographics = loo(demographics)))
```

The full model predicts better by around six times the standard error of the
difference, which is what we would expect: how sick a patient is on arrival is
the main thing that predicts whether they die.

`vignette("comparison")` covers model comparison, and the cases where
leave-one-out fails.

## Limitations

The model is flexible about the shape of the relationship and nothing else.

It does not make an association causal. Patients were not randomized to
catheterization; sicker patients were more likely to receive it, which is exactly
the kind of confounding that can produce an apparent harm. Whether the estimate
above can be read as the effect of the procedure depends on whether the
covariates account for that selection, which is a question about the study and
not about the fit. `vignette("causal")` covers what is required, using this same
data.

It does not extrapolate reliably. A predictor value outside the range of the
training data is mapped to the edge of that range, so the prediction is held
flat at whatever the fitted function is at the edge rather than continuing any
trend.

It does not fix a badly chosen family. Getting the outcome distribution wrong
matters more than any sampler setting.

## Further Reading

| Topic | Vignette |
| --- | --- |
| How BART works, and the sampler | `vignette("implementation")` |
| Choosing a family | `vignette("families")` |
| Convergence and fit | `vignette("diagnostics")` |
| Variable importance and selection | `vignette("importance")` |
| Effects, Curves, and Interactions | `vignette("effects")` |
| Varying coefficients and random intercepts | `vignette("varying")` |
| Model comparison | `vignette("comparison")` |
| Causal inference | `vignette("causal")` |
| Censored and survival outcomes | `vignette("survival")` |
| Frequently asked questions | `vignette("faq")` |

`?bartisan-package` has a longer map, organized by task and pointing at
functions rather than vignettes.

## References
