---
title: "Causal inference with BART"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Causal inference with BART}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.0, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 4, marginaleffects_safe = FALSE)

has_me <- rlang::is_installed("marginaleffects")
has_cobalt <- rlang::is_installed("cobalt")
run <- has_me

if (rlang::is_installed("future")) {
  future::plan(future::multisession)
}
```

## Introduction

BART is a popular choice for causal inference because it allows one to fit the nuisance functionals required for effect estimation (the relationships among the outcome, the treatment, and the potential confounders) in a flexible manner [@hill2011]. BART has repeatedly been shown to outperform other effect estimation methods in competitions [@dorie2019]. In this vignette, we demonstrate how to use BART implemented in *bartisan* to estimate treatment effects. This includes both standard BART as well as Bayesian causal forests (BCF), which is an improvement on traditional BART that works by fitting separate BART models for the outcome under control and for the treatment effect itself.

However, it's important to remember that BART is not a "causal inference method"; it is just a method of estimating certain quantities, which often are interpretable as associations. The assumptions that turn an association into a causal effect are assumptions about the design, and no model supplies them.

The example is the one the data were collected for: whether right heart catheterization helps or harms critically ill patients [@connors1996].

```{r libs, eval = run}
library(bartisan)
library(cobalt)
library(marginaleffects)

data(rhc)
```

## Confounding is the whole problem

Catheterization was not randomized. Doctors chose it, and they chose it more
often for patients who were already doing badly. Those patients were also more
likely to die. We can use `cobalt::bal.tab()` to see how the patients differ between groups:

```{r}
bal.tab(rhc ~ age + sex + race + edu + aps + meanbp + resp +
          hema + pafi + paco2 + crea + surv2m + card,
        data = rhc, stats = c("m", "ovl"), disp = "m")
```

The `M.0.Un` column indicates the mean for each variable in control group and the `M.1.Un` column indicates the mean for each variable in treated group. `Diff.Un` and `OVL.Un` are measures of the distributional difference between the groups for each covariate; values far from 0 indicate imbalance due to differential selection into treatment. In particular, we can see that patients with higher values of `aps` and `crea` and lower values of `meanbp, hema`, `pafi`, `paco2`, and `surv2m` are overrepresented among treated units.

Given that sicker patients are both more likely to die and more likely to receive RHC, it would not be unexpected to see that patients who receive RHC are more likely to die, and we do see that:

```{r naive, eval = run}
with(rhc, tapply(death, rhc, mean))
```

However, that doesn't mean RHC causes death; to distangle the effects of RHC from the confounding effects of pateitns characteristics, we need to adjust for them.

## What has to be true

Several assumptions are required to interpret an adjusted effect estimate as causal, none of which the fit can check.

**No unmeasured confounding.** Every common cause of catheterization and death is
in the model. Here that is the crux: the covariates include a physiological
profile and the study's own prognostic score, which is a serious attempt, but a
doctor's judgment at the bedside may not be fully captured by these fourteen variables.

**Positivity.** Every kind of patient could have received the procedure or not.
This one is partly checkable and is checked below.

**Consistency.** "Catheterization" names a single well defined intervention.

**No interference.** One patient's treatment does not affect another's outcome.

The first is the one that usually fails and the one to be explicit about. State
it as an assumption in what you write, rather than letting the interval imply it
has been handled.

### Checking positivity

Positivity fails when some combination of covariates makes the treatment nearly
certain. To look for that, model the treatment and inspect the fitted
probabilities in both groups. BART is a good choice for this model too, for the
same reason it is a good choice for the outcome: the functional form is not
known.

```{r psfit, eval = run}
set.seed(2026)

ps_fit <- bartisan(
  rhc ~ age + sex + race + edu + aps + meanbp + resp + hema + pafi +
    paco2 + crea + surv2m + card,
  data = rhc, family = binomial(), chains = 4
)

prop_score <- fitted(ps_fit)
```

We can use `cobalt::bal.plot()` to examine the overlap of the propensity score distribution between the groups:

```{r balplot, eval = run && has_cobalt, fig.height = 2.8}
bal.plot(rhc ~ prop_score, data = rhc, type = "hist", mirror = TRUE)
```

What matters for positivity is that the two groups overlap over most of their
range, and they do. Distributions pushed against zero and one with little overlap
are what failure looks like, and the honest response then is to restrict the
analysis to the region of overlap rather than let the model extrapolate into a
part of the covariate space where one treatment was never observed.

This check belongs before the outcome model, not after. A flexible outcome model
will happily produce an estimate in a region with no data, and the interval will
not tell you that is what happened.

## The estimator: g-computation

To estimate the average treatment effect $\tau_{\text{ATE}} = E[Y(1)] - E[Y(0)]$, which is a function of the unoabsevred potential outcoem $Y(1)$ and $Y(0)$, we can use the assumptions above to express it as a function of observed quantities:

$$
E[Y(1)] - E[Y(0)] = E \left[ E[Y | X, A = 1] \right] - E \left[ E[Y | X, A = 0] \right]
$$
To estimate $E \left[ E[Y | X, A = a] \right] = \theta_a$, we use g-computation:

$$
\hat{\theta}_a = \frac{1}{n}\sum_{i=1}^n {\mu(a, x_i)}
$$

where $\mu(a, x_i)$ is the predicted value from a regression of $Y$ on $A$ and $X$ for a unit with covariate profile $X=x_i$ and treatment $A$ set to $a$. This estimator is sometimes also known as the "regression estimator" or the "plug-in estimator".

Here, we use traditional BART and BCF to model $\mu(A, X)$. The rest of the analysis comes from the definitions above.

### One setting to change first

Before we fit the outcome model, we need to change one setting to make traditional BART suitable for estimating the ATE. The default splitting prior is a variable-selection prior. It can drop a
predictor from every tree in the forest at once, which is what makes it worth
having when the goal is prediction, and exactly what you do not want when the
estimand is a contrast on one particular predictor. Every outcome model below is fitted with `sparsity = FALSE`, which weights the
predictors equally and cannot drop any of them.

The propensity score model above keeps the default, and should: predicting who
was treated is a prediction problem, and no contrast is read off it.
`?bartisan_control` has the measurements behind both halves of this, and
`split_prior` is the alternative when there are enough covariates that weighting
them all alike is wasteful.

## The outcome model

We can fit the outcome model as follows:

```{r po, eval = run}
set.seed(2026)

fit <- bartisan(death ~ rhc + age + sex + race + edu + aps + meanbp + resp +
                  hema + pafi + paco2 + crea + surv2m + card,
                data = rhc, family = binomial(), chains = 4, sparsity = FALSE)
```

We include both the treatment and covariate in the formula's right-hand side, set `family = binomial()` to model the binary outcome with logistic regression, and `sparsity = FALSE` to remove the sparsity-inducing prior. Normally, we would examine convergence diagnostics for this model to make sure it was fit correctly; see `vignette("diagnostics")` for more information on how to do that.

## Potential outcomes

The quantities underlying every effect below are the two average potential
outcomes: the proportion who would die if every patient were catheterized, and if
none were. `marginaleffects::avg_predictions()` reports them directly.

```{r}
avg_predictions(fit, variables = "rhc")
```

The two rows are the estimates of $E[Y(0)]$ and $E[Y(1)]$, each averaged over the observed
covariate distribution. Reporting both is often more informative than reporting
their difference alone, because a difference of six percentage points means
something different against a baseline of 63% than it would against 5%.

## The average treatment effect

The effect is the contrast between those two quantities, which can be requested using `marginaleffects::avg_comparisons()`^[Here we report the risk difference, the default quantity returned by `avg_comparisons()` with a binary outcome. You can also set `comparison = "lnratioavg", transform = "exp"` to report the risk ratio and its credible interval.]:

```{r ate, eval = run}
ate <- avg_comparisons(fit, variables = "rhc")

ate
```

Under the assumptions above this is the average treatment effect: catheterization
raises the probability of death by about `r round(100 * abs(ate$estimate[1]))` percentage points, with an interval
running from roughly `r round(100 * abs(ate$conf.low[1]))` to `r round(100 * abs(ate$conf.high[1]))`.

## Using Bayesian causal forests

Traditional BART shrinks the fitted function toward a constant; that shrinkage applies to
everything the forest fits, including the part of the outcome that depends on the
exposure. When the exposure is strongly predicted by the covariates, the forest
can explain the outcome using the covariates alone, leave little for the exposure
to explain, and shrink the estimated effect toward zero. The interval shrinks with
it, so the result is a confident estimate biased toward no effect.

@hahn2020 identified this mechanism. Their remedy is to give the treatment effect
its own forest with its own prior, so that shrinking the confounding part does
not shrink the effect, which is the Bayesian causal forest (BCF) model, a special case of the varying coefficients BART model. `bcf()` fits it.

One specifies the control function in the model formula and identifies the treatment in the `treatment` argument. `bcf()` then fits a varying coefficient BART model, the BCF. 

```{r bcffit, eval = run}
set.seed(2026)

fit_bcf <- bcf(death ~ age + sex + race + edu + aps + meanbp + resp + hema +
                 pafi + paco2 + crea + surv2m + card,
               treatment = ~ rhc, data = rhc,
               family = binomial(), chains = 4)

fit_bcf
```

Using `bcf()` rather than writing the varying coefficient BART model by hand sets five settings to improve effect estimation: the effect gets a forest of its own, the estimated propensity score goes into the control function and not into the effect
forest, the effect forest gets fewer trees because effect heterogeneity is
usually simpler than a prognostic surface, `sparsity = FALSE` for the reason in
the section above, and a binary treatment's coding is drawn rather than fixed, so
the answer does not depend on which arm was written as 1.

The output is just like a regular BART model, so we can use `avg_comparisons()` to extract the average treatment effect estimate.

```{r}
avg_comparisons(fit_bcf, variables = "rhc")
```

Because the effect is a parameter of the model rather than a difference of
predictions, it can be read off directly, one value per patient:

```{r bcfcoef, eval = run}
coef(fit_bcf)[, "rhc"] |>
  quantile(probs = c(0, .25, .5, .75, 1)) |>
  round(3)
```

That column is the conditional effect for each patient, **on the link scale** --
here a difference in log odds, because the outcome is binary. It is not the same
number as the contrast above, which is on the probability scale, and for a binary
outcome the probability scale is the one to report. `coef()` is for looking at how
the effect varies from patient to patient, not for reporting its size.

For a family whose link is the identity, the two coincide exactly: the
coefficient *is* the contrast, and averaging it gives the average treatment
effect with nothing lost in between.

None of this changes what has to be true for the number to be causal. It changes
how the prior treats the effect, not whether the covariates account for
selection.

## A second example: a continuous outcome, and the ATT

The catheterization question is about a whole population, so the ATE is the
estimand. Many questions are not. When a program is offered to a particular group
and the question is whether it helped *them*, the ATT is what to report, and the
outcome is often continuous rather than binary.

`lalonde`, from *cobalt*, is the standard example: a job training program, with
earnings in 1978 as the outcome.

```{r lalonde, eval = run && has_cobalt}
data("lalonde", package = "cobalt")

with(lalonde, tapply(re78, treat, mean))
```

Participants earned less than non-participants. Taken at face value the program
looks harmful, and it is not: participants were selected for being out of work,
so they would have earned less anyway. This is confounding in the opposite
direction to the catheterization example, and it is why the raw comparison is
worth showing before the adjusted one.

```{r lalondefit, eval = run && has_cobalt}
set.seed(2026)

fit_earn <- bartisan(
  re78 ~ treat + age + educ + race + married + nodegree + re74 + re75,
  data = lalonde, family = dpm(), chains = 4, sparsity = FALSE
)

avg_comparisons(fit_earn, variables = "treat",
                newdata = subset(treat == 1))
```

`newdata = subset(treat == 1)` is what makes this the ATT: the individual
contrasts are averaged over the participants alone rather than over everyone. The
adjusted estimate is positive where the raw difference was negative, though its
interval is wide enough that the size is not settled.

`family = dpm()` is the default for a numeric outcome and the right choice here.
Earnings are heavily skewed with a spike at zero, which a single normal describes
badly; the Dirichlet process mixture estimates the shape instead of assuming it,
and costs almost nothing when a normal would have done. See
`vignette("families")`.

The two potential outcomes are worth reporting alongside the difference, since a
few hundred dollars means something different against a baseline of six thousand
than it would against six hundred:

```{r lalondepo, eval = run && has_cobalt}
avg_predictions(fit_earn, variables = "treat",
                newdata = subset(treat == 1))
```

Everything said above about identification applies here unchanged. The
observational Lalonde sample is famous precisely because its covariates are *not*
enough to recover the experimental benchmark, which is a caution rather than a
counterexample: it is what the assumption failing looks like.

## Moderation

A forest is free to fit a different effect for every covariate pattern, so asking
whether the effect differs across groups needs no interaction term and no refit.
`by` splits the average.

```{r moderation, eval = run}
avg_comparisons(fit, variables = "rhc", by = "card")
```

The estimates differ a little between patients with and without cardiovascular
disease. It is tempting to read a difference like that as moderation, and more
tempting still when the two intervals do not overlap equally or when one covers
zero and the other does not. Neither reading is a test. Comparing two intervals
says nothing about whether the quantities they cover differ, because the
comparison that matters has an uncertainty of its own that neither interval
reports.

The question is whether they differ from each other, which needs the difference
itself:

```{r pairwise, eval = run}
avg_comparisons(fit, variables = "rhc", by = "card",
                hypothesis = ~pairwise)
```

The difference is small with an interval covering zero. There is no evidence of
moderation by cardiovascular disease here, and the subgroup estimates should be
read as two noisy estimates of one common effect.

Detecting moderation needs much more data than estimating a main effect, and a
flexible model does not change that. What it changes is that you are not required
to guess the form of the interaction in advance.



## What the interval means

The posterior interval is a credible interval for the estimand under the model
and under the identification assumptions. It covers uncertainty from having a
finite sample and from not knowing the shape of the outcome surface. It does not
cover uncertainty about whether the assumptions hold.

An unmeasured confounder does not widen the interval. It moves the estimate and
leaves the interval where it was. This is why a sensitivity analysis, asking how
strong a confounder would have to be to explain the result away, is a more
informative addition than any refinement of the model. The `sensemakr` and
`EValue` packages do this and work from the estimate and its standard error,
which `avg_comparisons()` provides.

## Where to go next

`vignette("effects")` covers the estimand machinery in more detail, including
subgroup effects and testing whether they differ. `vignette("diagnostics")`
covers checking the fit, which should happen before any of this is interpreted.

## References
