---
title: "Causal inference with BART"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Causal inference with BART}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.0, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 4, marginaleffects_safe = FALSE)

has_me <- rlang::is_installed("marginaleffects")
has_cobalt <- rlang::is_installed("cobalt")
run <- has_me
```

## Introduction

BART is a popular choice for causal inference [@hill2011], for a good reason: the
step a flexible model helps with is the one people get wrong. Estimating a
treatment effect from observational data requires modelling how the outcome
depends on the covariates, and a linear model imposes a functional form that is
rarely justified and can bias the answer when it is wrong.

What BART does not do is deliver identification. The assumptions that turn an
association into a causal effect are assumptions about the design, and no model
supplies them. This vignette covers the part BART does help with, the checks that
come before it, and one specific way a regularized model can go wrong in this
setting that does not arise elsewhere.

The example is the one the data were collected for: whether right heart
catheterization helps or harms critically ill patients [@connors1996].

```{r libs, eval = run}
library(bartisan)
library(marginaleffects)

data(rhc)

model <- death ~ rhc + age + sex + race + edu + aps + meanbp + resp +
  hema + pafi + paco2 + crea + surv2m + card
```

## The confounding is the whole problem

Catheterization was not randomized. Doctors chose it, and they chose it more
often for patients who were already doing badly. Those patients were also more
likely to die. So the raw comparison is confounded in a specific and predictable
direction:

```{r naive, eval = run}
with(rhc, tapply(death, rhc, mean))
```

Patients who were catheterized died more often. Some of that difference is the
procedure and some of it is that they were sicker to begin with, and nothing in
this table separates the two. Everything below is an attempt to.

## What has to be true

Four assumptions, none of which the fit can check.

**No unmeasured confounding.** Every common cause of catheterization and death is
in the model. Here that is the crux: the covariates include a physiological
profile and the study's own prognostic score, which is a serious attempt, but a
doctor's judgment at the bedside is not fully captured by fourteen variables.

**Positivity.** Every kind of patient could have received the procedure or not.
This one is partly checkable and is checked below.

**Consistency.** "Catheterization" names a single well defined intervention.

**No interference.** One patient's treatment does not affect another's outcome.

The first is the one that usually fails and the one to be explicit about. State
it as an assumption in what you write, rather than letting the interval imply it
has been handled.

## Checking positivity

Positivity fails when some combination of covariates makes the treatment nearly
certain. To look for that, model the treatment and inspect the fitted
probabilities in both groups. BART is a good choice for this model too, for the
same reason it is a good choice for the outcome: the functional form is not
known.

```{r psfit, eval = run}
set.seed(2026)

ps_fit <- bartisan(
  rhc ~ age + sex + race + edu + aps + meanbp + resp + hema + pafi +
    paco2 + crea + surv2m + card,
  data = rhc, family = binomial(), chains = 4
)

ps <- fitted(ps_fit)

round(do.call(rbind, tapply(ps, rhc$rhc, range)), 3)
```

*cobalt* draws the standard picture:

```{r balplot, eval = run && has_cobalt, fig.height = 2.8}
library(cobalt)

bal.plot(rhc ~ ps, data = rhc, type = "hist", mirror = TRUE)
```

The treated distribution sits to the right of the untreated one, which is the
confounding made visible: patients more likely to be catheterized were different
patients. What matters for positivity is that the two overlap over most of their
range, and they do. Distributions pushed against zero and one with little overlap
are what failure looks like, and the honest response then is to restrict the
analysis to the region of overlap rather than let the model extrapolate into a
part of the covariate space where one treatment was never observed.

This check belongs before the outcome model, not after. A flexible outcome model
will happily produce an estimate in a region with no data, and the interval will
not tell you that is what happened.

## One setting to change first

The default splitting prior is a variable-selection prior. It can drop a
predictor from every tree in the forest at once, which is what makes it worth
having when the goal is prediction, and exactly what you do not want when the
estimand is a contrast on one particular predictor.

The reason is arithmetic. In a draw where the treatment is in no tree, the fitted
outcome does not depend on it, so the contrast in that draw is not small but
*exactly* zero. The posterior of the effect is then a mixture: a point mass at
zero holding whatever share of draws dropped the treatment, and the rest spread
over nonzero values. If that share exceeds 2.5%, the lower bound of a 95%
interval is exactly zero however strong the evidence is, and the point estimate is
pulled toward zero along with it. On a treatment effect of 0.2 against residual
noise of 1, the prior halved the estimate and its interval covered the truth 60%
of the time rather than 95%.

So every outcome model below is fitted with `sparsity = FALSE`, which weights the
predictors equally and cannot drop any of them.

```{r sparsitynote, eval = FALSE}
bartisan(model, data = rhc, family = binomial(), chains = 4, sparsity = FALSE)
```

The propensity score model above keeps the default, and should: predicting who
was treated is a prediction problem, and no contrast is read off it.
`?bartisan_control` has the measurements behind both halves of this, and
`split_prior` is the alternative when there are enough covariates that weighting
them all alike is wasteful.

## Potential outcomes

The quantities underlying every effect below are the two average potential
outcomes: the proportion who would die if every patient were catheterized, and if
none were. `avg_predictions()` reports them directly.

```{r po, eval = run}
set.seed(2026)

fit <- bartisan(model, data = rhc, family = binomial(), chains = 4,
                sparsity = FALSE)

avg_predictions(fit, variables = "rhc")
```

The two rows are $E[Y(0)]$ and $E[Y(1)]$, each averaged over the observed
covariate distribution. Reporting both is often more informative than reporting
their difference alone, because a difference of six percentage points means
something different against a baseline of 63% than it would against 5%.

## The average treatment effect

The effect is the contrast between those two quantities, which is what
g-computation computes: predict for everyone as if treated, predict again as if
untreated, and average the difference.

```{r ate, eval = run}
ate <- avg_comparisons(fit, variables = "rhc")

ate
```

Under the assumptions above this is the average treatment effect: catheterization
raises the probability of death by about `r round(100 * abs(ate$estimate[1]))` percentage points, with an interval
running from roughly `r round(100 * abs(ate$conf.low[1]))` to `r round(100 * abs(ate$conf.high[1]))`.

Compare that with the raw difference in the first table, which was over nine
points. Adjustment has moved the estimate toward zero, which is what you expect
when the treated group was sicker. It has not moved it to zero, and it has not
changed its sign. That was the finding that made this study influential.

## A second example: a continuous outcome, and the ATT

The catheterization question is about a whole population, so the ATE is the
estimand. Many questions are not. When a program is offered to a particular group
and the question is whether it helped *them*, the ATT is what to report, and the
outcome is often continuous rather than binary.

`lalonde`, from *cobalt*, is the standard example: a job training program, with
earnings in 1978 as the outcome.

```{r lalonde, eval = run && has_cobalt}
data("lalonde", package = "cobalt")

with(lalonde, tapply(re78, treat, mean))
```

Participants earned less than non-participants. Taken at face value the program
looks harmful, and it is not: participants were selected for being out of work,
so they would have earned less anyway. This is confounding in the opposite
direction to the catheterization example, and it is why the raw comparison is
worth showing before the adjusted one.

```{r lalondefit, eval = run && has_cobalt}
set.seed(2026)

fit_earn <- bartisan(
  re78 ~ treat + age + educ + race + married + nodegree + re74 + re75,
  data = lalonde, family = dpm(), chains = 4, sparsity = FALSE
)

avg_comparisons(fit_earn, variables = "treat",
                newdata = subset(treat == 1))
```

`newdata = subset(treat == 1)` is what makes this the ATT: the individual
contrasts are averaged over the participants alone rather than over everyone. The
adjusted estimate is positive where the raw difference was negative, though its
interval is wide enough that the size is not settled.

`family = dpm()` is the default for a numeric outcome and the right choice here.
Earnings are heavily skewed with a spike at zero, which a single normal describes
badly; the Dirichlet process mixture estimates the shape instead of assuming it,
and costs almost nothing when a normal would have done. See
`vignette("families")`.

The two potential outcomes are worth reporting alongside the difference, since a
few hundred dollars means something different against a baseline of six thousand
than it would against six hundred:

```{r lalondepo, eval = run && has_cobalt}
avg_predictions(fit_earn, variables = "treat",
                newdata = subset(treat == 1))
```

Everything said above about identification applies here unchanged. The
observational Lalonde sample is famous precisely because its covariates are *not*
enough to recover the experimental benchmark, which is a caution rather than a
counterexample: it is what the assumption failing looks like.

## Moderation

A forest is free to fit a different effect for every covariate pattern, so asking
whether the effect differs across groups needs no interaction term and no refit.
`by` splits the average.

```{r moderation, eval = run}
avg_comparisons(fit, variables = "rhc", by = "card")
```

The estimates differ a little between patients with and without cardiovascular
disease. It is tempting to read a difference like that as moderation, and more
tempting still when the two intervals do not overlap equally or when one covers
zero and the other does not. Neither reading is a test. Comparing two intervals
says nothing about whether the quantities they cover differ, because the
comparison that matters has an uncertainty of its own that neither interval
reports.

The question is whether they differ from each other, which needs the difference
itself:

```{r pairwise, eval = run}
avg_comparisons(fit, variables = "rhc", by = "card",
                hypothesis = ~pairwise)
```

The difference is small with an interval covering zero. There is no evidence of
moderation by cardiovascular disease here, and the subgroup estimates should be
read as two noisy estimates of one common effect.

Detecting moderation needs much more data than estimating a main effect, and a
flexible model does not change that. What it changes is that you are not required
to guess the form of the interaction in advance.

## Regularization-induced confounding

This is the failure specific to using a regularized model for causal inference,
and it is worth understanding rather than filing away.

BART shrinks the fitted function toward a constant. That shrinkage applies to
everything the forest fits, including the part of the outcome that depends on the
exposure. When the exposure is strongly predicted by the covariates, the forest
can explain the outcome using the covariates alone, leave little for the exposure
to explain, and shrink the estimated effect toward zero. The interval shrinks with
it, so the result is a confident estimate biased toward no effect.

@hahn2020 identified this mechanism. Their remedy is to give the treatment effect
its own forest with its own prior, so that shrinking the confounding part does
not shrink the effect, which is the Bayesian causal forest model. `bcf()` fits
it.

```{r bcffit, eval = run}
set.seed(2026)

fit_bcf <- bcf(death ~ age + sex + race + edu + aps + meanbp + resp + hema +
                 pafi + paco2 + crea + surv2m + card,
               treatment = ~ rhc, data = rhc, family = binomial(), chains = 4)

avg_comparisons(fit_bcf, variables = "rhc")
```

The model is a *control function* for the probability of death with no
catheterization, and a separate forest for the effect. Writing it as `bcf()`
rather than by hand sets four things: the effect gets a forest of its own, the
estimated propensity score goes into the control function and not into the effect
forest, the effect forest gets fewer trees because effect heterogeneity is
usually simpler than a prognostic surface, and `sparsity = FALSE` for the reason
in the section above.

Because the effect is a parameter of the model rather than a difference of
predictions, it can be read off directly, one value per patient:

```{r bcfcoef, eval = run}
summary(coef(fit_bcf)[, "rhc"])
```

That column is the conditional effect for each patient, **on the link scale** --
here a difference in log odds, because the outcome is binary. It is not the same
number as the contrast above, which is on the probability scale, and for a binary
outcome the probability scale is the one to report. `coef()` is for looking at how
the effect varies from patient to patient, not for reporting its size.

For a family whose link is the identity, the two coincide exactly: the
coefficient *is* the contrast, and averaging it gives the average treatment
effect with nothing lost in between.

None of this changes what has to be true for the number to be causal. It changes
how the prior treats the effect, not whether the covariates account for
selection.

A cheaper mitigation is to include the estimated propensity score as an
additional covariate in the outcome model, and there is direct evidence on how
much it is worth. The 2016 Atlantic Causal Inference Competition compared a large
number of methods across 77 data generating processes [@dorie2019], and
@carnegie2019 used those data to decompose which features of a BART fit
contributed to its performance.

Two findings are worth carrying over. Bias was small for every BART variant, and
including the propensity score reduced average absolute bias by about a tenth,
which on that scale is a refinement rather than a rescue. Interval coverage is
where the differences showed: base BART covered at 83%, and the best combination
without targeted learning, using ten chains, symmetric intervals and the
propensity score as a covariate, reached about 92%. Both running several chains
and including the score contributed.

Carnegie notes a limitation that matters for how far to carry this: ignorability
and overlap held by construction in every competition dataset, so the value of
modelling treatment assignment may well be larger when they are strained, which
is the situation this section is about.

So, in order:

- Check how strongly the exposure is predicted by the covariates. The overlap
  plot above is the diagnostic. Densities concentrated at the ends mean strong
  prediction and more scope for this bias; here they are spread out and heavily
  overlapping, which is the benign case.
- Run several chains, which the fits here already do.
- Include the propensity score as a covariate.

```{r pscovariate, eval = run}
rhc_ps <- rhc
rhc_ps$ps <- ps

set.seed(2026)
fit_ps <- bartisan(update(model, . ~ . + ps), data = rhc_ps,
                   family = binomial(), chains = 4, sparsity = FALSE)

avg_comparisons(fit_ps, variables = "rhc")
```

The estimate barely moves, which is what to expect when overlap is good and the
exposure is not strongly predicted by the covariates.

## What the interval means

The posterior interval is a credible interval for the estimand under the model
and under the identification assumptions. It covers uncertainty from having a
finite sample and from not knowing the shape of the outcome surface. It does not
cover uncertainty about whether the assumptions hold.

An unmeasured confounder does not widen the interval. It moves the estimate and
leaves the interval where it was. This is why a sensitivity analysis, asking how
strong a confounder would have to be to explain the result away, is a more
informative addition than any refinement of the model. The `sensemakr` and
`EValue` packages do this and work from the estimate and its standard error,
which `avg_comparisons()` provides.

## Where to go next

`vignette("effects")` covers the estimand machinery in more detail, including
subgroup effects and testing whether they differ. `vignette("diagnostics")`
covers checking the fit, which should happen before any of this is interpreted.

## References
