---
title: "Causal Inference with BART"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Causal Inference with BART}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.0, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 4, marginaleffects_safe = FALSE)

has_me <- rlang::is_installed("marginaleffects")
has_cobalt <- rlang::is_installed("cobalt")
has_ggplot2 <- rlang::is_installed("ggplot2")
run <- TRUE

if (rlang::is_installed("future")) {
  future::plan(future::multisession)
}

if (interactive() && rlang::is_installed("progressr")) {
  progressr::handlers(global = TRUE)
}
```

## Introduction

BART is a popular choice for causal inference because it allows one to fit the nuisance functionals required for effect estimation (the relationships among the outcome, the treatment, and the potential confounders) in a flexible manner [@hill2011]. BART has repeatedly been shown to outperform other effect estimation methods in competitions [@dorie2019]. In this vignette, we demonstrate how to use BART implemented in *bartisan* to estimate treatment effects under the assumption of no unmeasured confounding. This includes both standard BART as well as Bayesian causal forests (BCF), which is an improvement on traditional BART that works by fitting separate BART models for the outcome under control and for the treatment effect itself as a varying coefficient model.

However, it's important to remember that BART is not a "causal inference method"; it is just a method of estimating certain quantities, which often are interpretable as associations. The assumptions that turn an association into a causal effect are assumptions about the design, and no model supplies them.

In addition to its use here, BART can be used with instrumental variables analysis [@mccullochCausalInferenceInstrumental2021], regression discontinuity [@alcantaraModifiedBARTLearning2024], and difference-in-differences [@soutoForestsDifferencesRobust2025].

The example we use here is a dataset used to answer whether right heart catheterization helps or harms critically ill patients [@connors1996].

```{r libs, eval = run}
library(bartisan)

data(rhc)

set.seed(2026)
```

```{r libs2, eval = run, echo = FALSE}
if (has_cobalt) library(cobalt)
if (has_me) library(marginaleffects)
if (has_ggplot2) library(ggplot2)
```

## Confounding

Catheterization was not randomized. Doctors chose it, and they chose it more often for patients who were already doing badly. Those patients were also more likely to die. We can use `cobalt::bal.tab()` to see how the patients differ between groups:

```{r}
bal.tab(rhc ~ age + sex + race + edu + aps + meanbp + resp +
          hema + pafi + paco2 + crea + surv2m + card,
        data = rhc, stats = c("m", "ovl"), disp = "m")
```

The `M.0.Un` column indicates the mean for each variable in control group and the `M.1.Un` column indicates the mean for each variable in treated group. `Diff.Un` and `OVL.Un` are measures of the distributional difference between the groups for each covariate; values far from 0 indicate imbalance due to differential selection into treatment. In particular, we can see that patients with higher values of `aps` and `crea` and lower values of `meanbp, hema`, `pafi`, `paco2`, and `surv2m` are overrepresented among treated units.

Given that sicker patients are both more likely to die and more likely to receive RHC, it would not be unexpected to see that patients who receive RHC are more likely to die, and we do see that:

```{r naive, eval = run}
with(rhc, tapply(death, rhc, mean))
```

However, that doesn't mean RHC causes death; to disentangle the effects of RHC from the confounding effects of patients' characteristics, we need to adjust for these characteristics.

## Assumptions for Causal Inference

Several assumptions are required to interpret an adjusted effect estimate as causal, none of which the fit can check:

1.  **No unmeasured confounding.** Every common cause of catheterization and death is in the model. Here that is the crux: the covariates include a physiological profile and the study's own prognostic score, which is a serious attempt, but a doctor's judgment at the bedside may not be fully captured by these thirteen variables.

2.  **Positivity.** Every kind of patient could have received the procedure or not. This one is partly checkable and is checked below.

3.  **Consistency.** "Catheterization" names a single well defined intervention.

4.  **No interference.** One patient's treatment does not affect another's outcome.

The first is the one that usually fails and the one to be explicit about. State it as an assumption in what you write, rather than letting the interval imply it has been handled.

### Checking positivity

Positivity fails when some combination of covariates makes the treatment nearly certain. To look for that, model the treatment and inspect the fitted probabilities in both groups. BART is a good choice for this model too, for the same reason it is a good choice for the outcome: the functional form is not known.

```{r psfit, eval = run}
ps_fit <- bartisan(
  rhc ~ age + sex + race + edu + aps + meanbp + resp + hema + pafi +
    paco2 + crea + surv2m + card,
  data = rhc, family = binomial(), chains = 4
)

prop_score <- fitted(ps_fit)
```

We can use `cobalt::bal.plot()` to examine the overlap of the propensity score distribution between the groups:

```{r balplot, eval = run && has_cobalt}
bal.plot(rhc ~ prop_score, data = rhc, type = "hist", mirror = TRUE)
```

What matters for positivity is that the two groups overlap over most of their range, and they do. Distributions pushed against zero and one with little overlap are what failure looks like, and the honest response then is to restrict the analysis to the region of overlap rather than let the model extrapolate into a part of the covariate space where one treatment was never observed.

This check belongs before the outcome model, not after. A flexible outcome model will happily produce an estimate in a region with no data, and the interval will not tell you that is what happened.

## The Estimator: G-computation

To estimate the average treatment effect $\tau_{\text{ATE}} = E[Y(1)] - E[Y(0)]$, which is a function of the unobserved potential outcomes $Y(1)$ and $Y(0)$, we can use the assumptions above to express it as a function of observed quantities:

$$E[Y(1)] - E[Y(0)] = E \left[ E[Y | X, A = 1] \right] - E \left[ E[Y | X, A = 0] \right]$$ To estimate $E \left[ E \left[ Y | X, A = a \right] \right] = \theta_a$, we use g-computation [@snowdenImplementationGComputationSimulated2011]:

$$\hat{\theta}_a = \frac{1}{n}\sum_{i=1}^n {\mu(a, x_i)}$$

where $\mu(a, x_i)$ is the predicted value from a regression of $Y$ on $A$ and $X$ for a unit with covariate profile $X=x_i$ and treatment $A$ set to $a$. This estimator is sometimes also known as the "regression estimator" or the "plug-in estimator".

Here, we use traditional BART and BCF to model $\mu(A, X)$. The rest of the analysis comes from the definitions above.

### Controlling Sparsity

Before we fit the outcome model, we need to change one setting to make traditional BART suitable for estimating the ATE. The default splitting prior is a variable-selection prior. It can drop a predictor from every tree in the forest at once, which is what makes it worth having when the goal is prediction, and exactly what you do not want when the estimand is a contrast on one particular predictor. Every outcome model below is fitted with `sparsity = FALSE`, which weights the predictors equally and cannot drop any of them.

The propensity score model above keeps the default, and should: predicting who was treated is a prediction problem, and no contrast is read off it. `?bartisan_control` explains both halves of this, and `split_prior` is the alternative when there are enough covariates that weighting them all alike is wasteful.

## The outcome model

We can fit the outcome model as a simple binary BART regression of the outcome on the treatment and covariates, as in @hill2011:

```{r po, eval = run}
fit <- bartisan(death ~ rhc + age + sex + race + edu + aps + meanbp + resp +
                  hema + pafi + paco2 + crea + surv2m + card,
                data = rhc, family = binomial(),
                chains = 4, sparsity = FALSE)
```

We include both the treatment and covariate in the formula's right-hand side, set `family = binomial()` to model the binary outcome with logistic regression, and `sparsity = FALSE` to remove the sparsity-inducing prior. We could also have included the propensity score as a covariate, which is recommended by @carnegie2019 to slightly improve performance (in this case it doesn't affect the result, which has also been reported by @soutoAblationStudiesNovel2024). Normally, we would examine convergence diagnostics for this model to make sure it was fit correctly; see `vignette("diagnostics")` for more information on how to do that.

## Potential outcomes

The quantities underlying the effect estimate are the two average potential
outcomes: the proportion who would die if every patient were catheterized, and
if none were. `estimate_effect()` computes them on the way to the
effect and keeps them, and `summary()` on a `bcf()` fit prints them beside it. On
a fit from `bartisan()` the treatment has to be named, since nothing in the
formula marks one predictor as the treatment, and `summary()` there reports the
forests rather than an effect:

```{r ate, eval = run}
ate <- estimate_effect(fit, treatment = "rhc")

ate
```

Under the assumptions above this is the average treatment effect: catheterization
raises the probability of death by about
`r if (run) round(100 * abs(ate$estimate[1])) else "3"` percentage points, with
an interval running from roughly
`r if (run) round(100 * ate$lower[1], 1) else "0"` to
`r if (run) round(100 * ate$upper[1], 1) else "8"`.

The potential outcomes come with it, in the `"potential_outcomes"` attribute:

```{r po_attr, eval = run}
attr(ate, "potential_outcomes")
```

The two rows are the estimates of $E[Y(0)]$ and $E[Y(1)]$, each averaged over the
observed covariate distribution. Reporting both is often more informative than
reporting their difference alone, because a difference of a few percentage points
means something different against a baseline of 63% than it would against 5%.

### How the estimate is computed, and on which scale

`estimate_effect()` does the g-computation above literally. Every unit is
predicted twice, once with the treatment set to each of its levels; the two sets
of predictions are averaged over the units the estimand asks for, *within each
posterior draw*; and only then are the two averages contrasted. What comes back
is a posterior for the estimand, summarized by its mean and a credible interval.

That order is what makes a ratio here the marginal ratio rather than the average
of the conditional ones, which is a different quantity. For a difference the two
orders agree, so it only shows up once a ratio is asked for.

The scale matters, and it is the reason the default is `type = "response"`. On
that scale the average of the unit-level differences *is* the marginal effect. A
contrast read off the link scale is not: on a logistic fit the average of the
conditional log odds ratios is not the marginal log odds ratio, and the two can
differ by a good deal. `comparison` asks for the contrast rather than the scale,
so a risk ratio or an odds ratio is available without leaving the response scale:

```{r ate_or, eval = run}
estimate_effect(fit, treatment = "rhc", comparison = "lnor")
```

### The same answer through *marginaleffects*

A fit also works with *marginaleffects*, and for the average effect the two
routes compute the same thing from the same draws. The one thing to set is the
posterior summary: *marginaleffects* reports the median by default and
`estimate_effect()` reports the mean, so an unadjusted comparison shows a
difference that is not there.

```{r me_equiv, eval = run && has_me}
options(marginaleffects_posterior_center = "mean")

avg_comparisons(fit, variables = "rhc")
```

The point estimate and the interval match `estimate_effect()` above to machine
precision. Which to reach for is a question of what else is wanted:
`estimate_effect()` covers the estimands a treatment question asks for and needs
no extra package, while *marginaleffects* covers a much wider class of
quantities. `vignette("effects")` is about the second.

## Using Bayesian Causal Forests

Traditional BART shrinks the fitted function toward a constant; that shrinkage
applies to everything the forest fits, including the part of the outcome that
depends on the exposure. When the exposure is strongly predicted by the
covariates, the forest can explain the outcome using the covariates alone,
leaving little for the exposure to explain, and shrink the estimated effect
toward zero. The interval shrinks with it, so the result is a confident estimate
biased toward no effect.

@hahn2020 identified this mechanism; their remedy is to give the treatment effect
its own forest with its own prior, so that shrinking the confounding part does
not shrink the effect. This is the Bayesian causal forest (BCF) model, a special
case of the varying coefficients BART model. `bcf()` fits it.

One specifies the control function in the model formula and identifies the
treatment in the `treatment` argument. `bcf()` then fits a varying coefficient
BART model, the BCF. By default, `bcf()` estimates a propensity score using a
logistic BART model and includes that as a covariate in the control function, as
recommended by @hahn2020.

```{r bcffit, eval = run}
fit_bcf <- bcf(death ~ age + sex + race + edu + aps + meanbp + resp + hema +
                 pafi + paco2 + crea + surv2m + card,
               treatment = ~ rhc, data = rhc,
               family = binomial(), chains = 4)

fit_bcf
```

Using `bcf()` rather than writing the varying coefficient BART model by hand sets
four settings to improve effect estimation:

1.  the effect gets a forest of its own
2.  the estimated propensity score goes into the control function and not into the effect forest
3.  the effect forest gets fewer trees because effect heterogeneity is usually simpler than a prognostic surface
4.  a binary treatment's coding is drawn rather than fixed, so the answer does not depend on which arm was written as 1.

Note that `sparsity = FALSE` is not among them: the reason for it in the section
above is that the prior can drop the predictor whose contrast we want, and here
the treatment is the coefficient rather than a predictor the forest splits on, so
nothing can drop it.

Because the treatment is named in the call, the fit knows what its own effect is
and `summary()` reports it. The treatment does not have to be named again:

```{r bcfsummary, eval = run}
summary(fit_bcf)
```

Three blocks: the average effect, the two potential outcomes it is a difference
of, and the quartiles of the per-unit effects, which say how much the effect
varies rather than how well any one patient is estimated.

### Conditional effects

The effect forest gives one value per patient, and `estimate_effect()` returns
them with `estimand = "CATE"`. Here we ask for them as odds ratios, which for a
conditional effect is a conditional odds ratio:

```{r bcfcate, eval = run}
cate <- estimate_effect(fit_bcf, estimand = "CATE", comparison = "or")

quantile(cate$estimate, probs = c(0, .25, .5, .75, 1))
```

`plot()` draws them: patients ordered by their estimate, with a credible interval
each and the marginal effect as a band behind them, which is what makes the
spread readable as heterogeneity rather than as a list of numbers.

```{r bcfplot, eval = run && has_ggplot2, fig.height = 3.4}
plot(fit_bcf)
```

`plot()` on the fit is the conditional effects on the response scale; passing a
`<bartisan_effect>` object to `plot()` draws whatever that object holds, so
`plot(estimate_effect(fit_bcf, by = ~ sex))` is a subgroup forest plot instead.

Note that a conditional odds ratio is not the marginal one, and their average is
not it either. That is the scale caution from earlier in a second place: an
average of conditional contrasts equals the marginal contrast only when the
contrast is a difference. For an identity link, which the next example uses, the
ATE *is* the average of the conditional effects.

## A second example: a continuous outcome, and the ATT

The catheterization question is about a whole population, so the ATE is the
estimand. Many questions are not. When a program is offered to a particular group
and the question is whether it helped *them*, the ATT is what to report, and the
outcome is often continuous rather than binary.

`lalonde`, from *cobalt*, is the standard example: a job training program, with
earnings in 1978 as the outcome.

```{r lalonde, eval = run && has_cobalt}
data("lalonde", package = "cobalt")

with(lalonde, tapply(re78, treat, mean))
```

Participants earned less than non-participants. Taken at face value, the program
looks harmful, but that doesn't mean it actually is: participants were selected
for being out of work, so they would have earned less anyway. This is confounding
in the opposite direction to the catheterization example, and it is why the raw
comparison is worth showing before the adjusted one.

Below, we fit the BCF model, this time setting `family = dpm()`, which is the
default for a numeric outcome and makes sense here. Earnings are heavily skewed
with a spike at zero, which a single normal describes badly; the Dirichlet process
mixture estimates the shape instead of assuming it, and costs almost nothing when
a normal would have done. Alternatively, `family = tweedie()` can also work well
for this type of outcome. See `vignette("families")`.

```{r lalondefit, eval = run && has_cobalt}
fit_earn_bcf <- bcf(
  re78 ~ age + educ + race + married + nodegree + re74 + re75,
  treatment = ~ treat,
  data = lalonde, family = dpm(),
  chains = 4
)

summary(fit_earn_bcf, estimand = "ATT")
```

`estimand = "ATT"` averages over the treated rather than over everyone, and
`focal` is not needed because a 0/1 treatment settles which level is the treated
one. Two cases do need it, and they differ. When the treatment has more than two
levels, `focal` is required and naming it is the whole of the choice, since
`"ATT"` and `"ATC"` then mean the same thing: the effect among the units in the
level named. When it has two levels whose labels say nothing about which is which,
the level order is assumed and a message says so, which `focal` silences.

The three blocks read as before, and the potential outcomes earn their place
here: a few hundred dollars means something different against a baseline of six
thousand than it would against six hundred.

Because the link is the identity, the ATT is exactly the average of the
conditional effects among the treated, which is worth checking once to see that
the two agree:

```{r lalondecate, eval = run && has_cobalt}
att <- estimate_effect(fit_earn_bcf, estimand = "ATT")
cate_att <- estimate_effect(fit_earn_bcf, estimand = "CATE",
                            newdata = subset(lalonde, treat == 1))

c(ATT = att$estimate, mean_CATE = mean(cate_att$estimate))
```

And the conditional effects, drawn:

```{r lalondeplot, eval = run && has_cobalt && has_ggplot2, fig.height = 3.4}
plot(cate_att)
```

Most of the conditional effects are positive, as the ATT is, and none of their
intervals excludes zero. That is the usual picture: a per-unit effect is estimated
from far less information than an average, so the intervals are wide even where
the average is clear.

## What the credible interval means

The posterior interval is a credible interval for the estimand under the model and under the identification assumptions. It covers uncertainty from having a finite sample and from not knowing the shape of the outcome surface. It does not cover uncertainty about whether the assumptions hold.

An unmeasured confounder does not widen the interval. It moves the estimate and leaves the interval where it was. This is why a sensitivity analysis, asking how strong a confounder would have to be to explain the result away, is a more informative addition than any refinement of the model.

## Where to go next

`?estimate_effect` is the reference for the estimands used here, including the subgroup form and the contrast types. `vignette("effects")` covers what *marginaleffects* adds beyond them, which is a much wider class of quantities and the route to take when the question is not a treatment contrast. `vignette("diagnostics")` covers checking the fit, which should happen before any of this is interpreted.

## References
