---
title: "Causal inference with BART"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Causal inference with BART}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.0, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 4, marginaleffects_safe = FALSE)

has_me <- rlang::is_installed("marginaleffects")
has_cobalt <- rlang::is_installed("cobalt")
run <- rlang::is_installed("MASS") && has_me
```

## Introduction

BART is a popular choice for causal inference [@hill2011], for a good reason: the
step a flexible model helps with is the one people get wrong. Estimating a
treatment effect from observational data requires modelling how the outcome
depends on the covariates, and a linear model imposes a functional form that is
rarely justified and can bias the answer when it is wrong.

What BART does not do is deliver identification. The assumptions that turn an
association into a causal effect are assumptions about the design, and no model
supplies them. This vignette covers the part BART does help with, the checks that
come before it, and one specific way a regularized model can go wrong in this
setting that does not arise elsewhere.

The running example treats maternal smoking as the exposure and birth weight as
the outcome. It is used because it is familiar and the covariates are plausible,
not because these data support a causal claim.

```{r libs, eval = run}
library(bartisan)
library(marginaleffects)

data("birthwt", package = "MASS")

d <- birthwt
d$race  <- factor(d$race,  levels = 1:3, labels = c("white", "black", "other"))
d$smoke <- factor(d$smoke, levels = 0:1, labels = c("no", "yes"))
d$ht    <- factor(d$ht,    levels = 0:1, labels = c("no", "yes"))
d$ui    <- factor(d$ui,    levels = 0:1, labels = c("no", "yes"))
d$low   <- NULL
```

## What has to be true

Four assumptions, none of which the fit can check.

**No unmeasured confounding.** Every common cause of smoking and birth weight is
in the model. Here that is implausible: socioeconomic status, nutrition and
partner smoking are absent and plainly relevant.

**Positivity.** Every kind of mother could have smoked or not. This one is partly
checkable and is checked below.

**Consistency.** "Smoking" names a single well defined exposure. Smoking two
cigarettes a day and twenty are treated as the same condition here, which they
are not.

**No interference.** One mother's smoking does not affect another's outcome.
Reasonable here.

The first is the one that usually fails and the one to be explicit about. State
it as an assumption in what you write, rather than letting the interval imply it
has been handled.

## Checking positivity

Positivity fails when some combination of covariates makes the exposure nearly
certain. To look for that, model the exposure and inspect the fitted
probabilities in both groups. BART is a good choice for this model too, for the
same reason it is a good choice for the outcome: the functional form is not
known.

```{r psfit, eval = run}
set.seed(2026)

ps_fit <- bartisan(smoke ~ . - bwt, data = d,
                   family = binomial(), chains = 4)

ps <- fitted(ps_fit)

round(do.call(rbind, tapply(ps, d$smoke, range)), 3)
```

Both groups span a similar range and neither approaches zero or one, so there is
no region where one exposure is effectively impossible.

\pkg{cobalt} draws the standard picture:

```{r balplot, eval = run && has_cobalt, fig.height = 2.8}
library(cobalt)

bal.plot(smoke ~ ps, data = d, type = "histogram",
         mirror = TRUE)
```

The two densities overlap over most of their range, which is what positivity
looks like when it holds. Distributions pushed against zero and one, with little
overlap, are what it looks like when it does not, and the honest response then is
to restrict the analysis to the region of overlap rather than let the model
extrapolate into a part of the covariate space where one exposure was never
observed.

A BART propensity score is more regularized than a logistic one and so is less
prone to the very extreme fitted probabilities a saturated logistic model
produces when a covariate pattern happens to be perfectly predictive. Those
extreme values are usually artefacts, and they are what makes a positivity
assessment look worse than it is.

This check belongs before the outcome model, not after. A flexible outcome model
will happily produce an estimate in a region with no data, and the interval will
not tell you that is what happened.

## Potential outcomes

The quantities underlying every effect below are the two average potential
outcomes: what birth weight would average if every mother smoked, and if none
did. `avg_predictions()` reports them directly.

```{r po, eval = run}
fit <- bartisan(bwt ~ ., data = d,
                family = dpm(), chains = 4)

avg_predictions(fit, variables = "smoke")
```

The two rows are \eqn{E[Y(0)]} and \eqn{E[Y(1)]}, each averaged over the observed
covariate distribution. Reporting both is often more informative than reporting
their difference alone, because a difference of 300 grams means something
different at 3000 grams than it would at 1500.

## The average treatment effect

The effect is the contrast between those two quantities, which is what
g-computation computes: predict for everyone as if exposed, predict again as if
unexposed, and average the difference.

```{r ate, eval = run}
ate <- avg_comparisons(fit, variables = "smoke")

ate
```

Under the assumptions above this is the average treatment effect: about
`r round(abs(ate$estimate))` grams lower, with an interval from roughly
`r round(abs(ate$conf.low))` to `r round(abs(ate$conf.high))`. It is the
difference of the two rows in the previous table, with an interval computed from
the posterior of the difference rather than from the two intervals separately.

The average treatment effect on the treated restricts the averaging to the mothers who actually smoked:

```{r att, eval = run}
avg_comparisons(fit, variables = "smoke",
                newdata = subset(smoke == "yes"))
```

These are close here. They differ when the effect varies with the covariates and
the exposed group is unlike the rest, which is exactly when the distinction
matters. Decide which one answers your question before fitting.

## Moderation

A forest is free to fit a different effect for every covariate pattern, so asking
whether the effect differs across groups needs no interaction term and no refit.
`by` splits the average.

```{r moderation, eval = run}
avg_comparisons(fit, variables = "smoke", by = "race")
```

The three estimates differ, and so does how close each interval comes to zero. It
is tempting to read that as the effect being present in some groups and absent in
others. That reading is not supported, and the reasoning behind it is not a test:
whether one interval covers zero and another does not says nothing about whether
the two effects differ. The question is whether they differ from each other,
which needs the difference itself, with its own interval.

```{r pairwise, eval = run}
avg_comparisons(fit, variables = "smoke", by = "race",
                hypothesis = ~pairwise)
```

Every pairwise difference has an interval comfortably covering zero, and the
intervals are wide relative to the differences. There is no evidence of
moderation by race here, and the subgroup estimates above should be read as three
noisy estimates of one common effect rather than as three different effects.

This is the honest reading and it is the usual one. Detecting moderation needs
much more data than estimating a main effect, and a flexible model does not
change that. What it changes is that you are not required to guess the form of
the interaction in advance.

## Regularization-induced confounding

This is the failure specific to using a regularized model for causal inference,
and it is worth understanding rather than filing away.

BART shrinks the fitted function toward a constant. That shrinkage applies to
everything the forest fits, including the part of the outcome that depends on the
exposure. When the exposure is strongly predicted by the covariates, the forest
can explain the outcome using the covariates alone, leave little for the exposure
to explain, and shrink the estimated effect toward zero. The interval shrinks with
it, so the result is a confident estimate biased toward no effect.

@hahn2020 identified this mechanism. Their remedy is to give the treatment effect
its own forest with its own prior, so that shrinking the confounding part does
not shrink the effect, which is the Bayesian causal forest model.

A cheaper mitigation is to include the estimated propensity score as an
additional covariate in the outcome model, and there is direct evidence on how
much it is worth. The 2016 Atlantic Causal Inference Competition compared a large
number of methods across 77 data generating processes [@dorie2019], and
@carnegie2019 used those data to decompose which features of a BART fit
contributed to its performance.

Two findings are worth carrying over. Bias was small for every BART variant, and
including the propensity score reduced average absolute bias by about a tenth,
which on that scale is a refinement rather than a rescue. Interval coverage is
where the differences showed: base BART covered at 83%, and the best combination
without targeted learning, using ten chains, symmetric intervals and the
propensity score as a covariate, reached about 92%. Both running several chains
and including the score contributed.

Carnegie notes a limitation that matters for how far to carry this: ignorability
and overlap held by construction in every competition dataset, so the value of
modelling treatment assignment may well be larger when they are strained, which
is the situation this section is about.

So, in order:

- Check how strongly the exposure is predicted by the covariates. The overlap
  plot above is the diagnostic. Densities concentrated at the ends mean strong
  prediction and more scope for this bias; here they are spread out and heavily
  overlapping, which is the benign case.
- Run several chains, which the fits here already do.
- Include the propensity score as a covariate.

```{r pscovariate, eval = run}
d$ps <- ps

fit_ps <- bartisan(bwt ~ ., data = d,
                   family = dpm(), chains = 4)

avg_comparisons(fit_ps, variables = "smoke")
```

The estimate barely moves, which is what to expect when overlap is good and the
exposure is not strongly predicted by the covariates.

## What the interval means

The posterior interval is a credible interval for the estimand under the model
and under the identification assumptions. It covers uncertainty from having a
finite sample and from not knowing the shape of the outcome surface. It does not
cover uncertainty about whether the assumptions hold.

An unmeasured confounder does not widen the interval. It moves the estimate and
leaves the interval where it was. This is why a sensitivity analysis, asking how
strong a confounder would have to be to explain the result away, is a more
informative addition than any refinement of the model. The `sensemakr` and
`EValue` packages do this and work from the estimate and its standard error,
which `avg_comparisons()` provides.

## Where to go next

`vignette("effects")` covers the estimand machinery in more detail, including
subgroup effects and testing whether they differ. `vignette("diagnostics")`
covers checking the fit, which should happen before any of this is interpreted.

## References
