---
title: "Choosing between models"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Choosing between models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.2, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 4)

has_loo <- rlang::is_installed("loo")
run <- has_loo
```

## Introduction

Model selection means something different here than it does for a linear model,
and the difference is worth stating before any code.

With `lm()` you choose which terms enter, whether to add a squared term, whether
to include an interaction. Those choices are the model. With BART the forest
makes them, so they are not yours to make. What remains is a shorter list:

1. Which variables the model is allowed to see.
2. Which likelihood, meaning which family and link.
3. Occasionally, a sampler setting such as the number of trees.

This vignette covers how to compare those choices.

```{r libs, eval = run}
library(bartisan)
library(loo)

data(rhc)

model <- death ~ rhc + age + sex + race + edu + aps + meanbp + resp +
  hema + pafi + paco2 + crea + surv2m + card

set.seed(2026)
full <- bartisan(model, data = rhc, family = binomial())
```

The fits here use the default single chain rather than the four used elsewhere.
Leave-one-out needs draws, not chains, and four times the fits would make this
vignette slow to build for no gain. Check convergence separately, as in
`vignette("diagnostics")`, before trusting any of these comparisons.

## Leave-one-out cross-validation

```{r loo, eval = run}
loo(full)
```

`elpd_loo` estimates the log predictive density on data the model has not seen,
with higher being better. It is computed by importance sampling from the fitted
posterior rather than by refitting, which is why it is fast [@vehtari2017].

`p_loo` is the effective number of parameters, about 32 here. For a forest with
hundreds of leaves across its trees, that number is small because the prior
shrinks most of them toward zero. It is a useful measure of how much of the data the
model is actually using.

The Pareto $k$ diagnostics are the thing to check. Leave-one-out by
importance sampling is trustworthy only when the importance weights are well
behaved, and $k$ above 0.7 says they are not for that observation. Here they
are all good.

### When it fails

A forest is a flexible function, so a single observation can have a lot of
influence on the leaves it falls into, and high $k$ values are more common
than for a parametric model. If many observations are flagged, the estimate is
unreliable and the answer is held-out data rather than a different diagnostic:

```{r holdout, eval = FALSE}
train <- d[1:4000, ]
held  <- d[4001:nrow(d), ]

fit <- bartisan(death ~ ., data = train, family = binomial())

sum(predict(fit, newdata = held, type = "density", log = TRUE))
```

This is the same quantity `elpd_loo` approximates, computed directly.

## Comparing two models

```{r compare, eval = run}
set.seed(2026)
demographics <- bartisan(death ~ rhc + age + sex + race + edu, data = rhc,
                         family = binomial())

loo_compare(list(full = loo(full), demographics = loo(demographics)))
```

The full model predicts better by around six times the standard error of the
difference. The standard error is the important half: a difference of many
standard errors is clear, and a difference smaller than its own standard error is
not evidence of anything.

This is the right way to ask whether a set of variables earns its place, and it
is a better question than the one variable importance answers, because it is
about prediction rather than about how the forest happened to spend its splits.
Here the answer is not in doubt: how sick a patient is on arrival predicts
whether they die, and demographics alone do not.

## Comparing links

The other choice is the likelihood. For a binary outcome the family is settled
and what remains is the link, which decides how the forest's output is mapped to
a probability.

```{r links, eval = run}
set.seed(2026)
probit <- bartisan(model, data = rhc, family = binomial("probit"))

set.seed(2026)
cloglog <- bartisan(model, data = rhc, family = binomial("cloglog"))

loo_compare(list(logit = loo(full), probit = loo(probit),
                 cloglog = loo(cloglog)))
```

The three are within a point or two of each other, and the differences are
smaller than their standard errors. *loo* flags this directly. The reading is
that the link does not matter here.

That is a useful negative result and worth reporting as one. It is also the usual
outcome: with a flexible function on the inside, the link has little left to do,
because the forest can absorb the difference between one link and another. This
is not true of a linear model, where the link carries the whole shape of the
relationship.

The one link worth thinking about separately is `cloglog`, which is asymmetric
and is the right choice when the outcome is a discretized survival time. See
`vignette("survival")`.

## A trap: the densities must be on the same scale

`loo()` is built on the pointwise log density of each observation, and comparing
two models by it assumes both densities are with respect to the same measure.
This is normally automatic and there is one case in this package where it is not.

The accelerated failure time families report the density of $\log T$ while
`ph()` reports the density of $T$. The two differ by a Jacobian, so a
comparison across that boundary is off by $\sum \log t$, which can be
hundreds of points and can reverse the ordering. `vignette("survival")` gives the
correction.

The general rule: `loo()` compares models fitted to the same outcome on the same
scale. If you transform the outcome, refit, and compare, the comparison is
invalid unless you account for the Jacobian. Comparing a model of `y` to a model
of `log(y)` is the everyday version of this mistake.

## Comparing against a logistic regression

There is no obstacle to this, and it is worth doing. If a logistic regression
predicts as well as the forest, that is evidence the relationship is close to
linear on the log-odds scale, and the simpler model is easier to report.

The comparison has to be like for like. Fit both in a framework that produces the
same kind of pointwise log density, for example by fitting the logistic
regression with *rstanarm* or *brms* and comparing with `loo_compare()`, or by comparing
held-out log scores computed the same way for both.

Comparing `elpd_loo` from this package against `AIC` from `glm()` is not a
comparison and should not be reported as one.

## What not to select on

Do not choose the number of trees, or `k`, or the gate, by cross-validation as a
matter of routine. The priors are chosen so that the defaults work across a wide
range of problems, and tuning them typically produces small gains and an
optimistically biased estimate of performance if the same data chose them. If you
do tune, hold out data for the final assessment.

Do not select variables by fitting many models and keeping the best. With a
handful of candidate specifications chosen in advance, `loo_compare()` is
informative. With a search over all subsets it is not, for the same reason
stepwise regression is not.

## Where to go next

`vignette("diagnostics")` covers checking that each model fits before comparing
them, which is the step people skip. `vignette("families")` covers what the
families assume, which is what a comparison between them is really about.

## References
