---
title: "Choosing between models"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Choosing between models}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{=html}
<style>
pre {overflow-x: auto;}
pre code {word-wrap: normal; white-space: pre;}
</style>
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.2, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 4)

has_loo <- rlang::is_installed("loo")
has_rstanarm <- rlang::is_installed("rstanarm")
has_surv <- rlang::is_installed("survival")

elpd <- function(x) unname(x$estimates["elpd_loo", "Estimate"])
has_cobalt <- rlang::is_installed("cobalt")
run <- has_loo
```

## Introduction

Model selection means something different for a BART model than it does for a linear model, and the difference is worth stating before any code.

With `lm()` we choose which terms enter, whether to add a squared term, and whether to include an interaction; those choices are the model. With BART the forest makes them, so they are not ours to make. What remains is a shorter list:

1.  Which variables the model is allowed to see.
2.  Which likelihood, meaning which family and link.
3.  Occasionally, a sampler setting (e.g., the number of trees).

In this guide we will compare those choices using leave-one-out cross-validation. First we'll fit a model to the `rhc` data and read its `loo()` output, including the diagnostics that say when the approximation cannot be trusted and what to do when it cannot. Next we'll compare two sets of variables and then three links, work through the two ways two log densities end up on different scales and repair both, and put the forest up against a Bayesian logistic regression. Finally we'll cover how to tune a setting and how to select variables, both of which have a right way and a more tempting wrong one.

```{r libs, eval = run}
library(bartisan)
library(loo)

data("rhc")

model <- death ~ rhc + age + sex + race + edu + aps + meanbp + resp +
  hema + pafi + paco2 + crea + surv2m + card

set.seed(2026)

full <- bartisan(model, data = rhc, family = binomial())
```

The fits here use the default single chain rather than the four used elsewhere; leave-one-out needs draws rather than chains, and four times as many fits would make this vignette slow to build for no gain. Convergence should still be checked separately, as in `vignette("diagnostics")`, before we trust any of these comparisons.

## Leave-One-Out Cross-Validation (`loo()`)

```{r loo, eval = run}
loo(full)
```

`elpd_loo` estimates the log predictive density on data the model has not seen, with higher values being better. It is computed by importance sampling from the fitted posterior rather than by refitting, which is why it is fast [@vehtari2017].

`p_loo` is the effective number of parameters, about 32 here; for a forest with hundreds of leaves across its trees that number is small, because the prior shrinks most of them toward zero. It is a useful measure of how much of the data the model is actually using.

The Pareto $k$ diagnostics are the thing to check: leave-one-out by importance sampling is trustworthy only when the importance weights are well behaved, and a $k$ above .7 says that for that observation they are not. Here they are all good.

### When the Approximation Fails

A forest is a flexible function, so a single observation can have a good deal of influence on the leaves it falls into, and high $k$ values are therefore more common than they are for a parametric model. If many observations are flagged, the estimate is unreliable, and the remedy is held-out data rather than a different diagnostic: fit to part of the sample and score the part the model was not shown.

```{r holdout, eval = run}
set.seed(2026)
train_id <- sample.int(nrow(rhc), 1200)

train <- rhc[train_id, ]
held  <- rhc[-train_id, ]

fit_train <- bartisan(model, data = train, family = binomial())

score <- predict(fit_train, newdata = held, type = "density", log = TRUE)

sum(score)
```

`type = "density"` evaluates the outcome under each posterior draw and averages over the draws before taking the log, which is the same form as one pointwise `elpd_loo` contribution. The two are therefore estimates of the same thing, which the two numbers do not make obvious:

```{r holdoutscale, eval = run}
elpd_total <- loo(full)$estimates["elpd_loo", "Estimate"]

rbind(loo     = c(total = elpd_total, n = nrow(rhc),     per_obs = elpd_total / nrow(rhc)),
      heldout = c(total = sum(score), n = length(score), per_obs = mean(score)))
```

The totals differ by a factor of five because they sum different numbers of terms: `loo()` scores every observation and the split scores only the ones held out. It is the per-observation column that the two can be read against each other, and there they nearly agree. The gap that remains is in the direction to expect, since leave-one-out trains on 1499 observations and the split trains on 1200.

That is the whole of what a held-out score gives us on its own, which is not much: a log score has no scale, and the only thing to do with one is put it beside another. The next section does that, first through `loo()` and then through this split.

## Comparing Two Models (`loo_compare()`)

We can compare the fit of two models by supplying their `loo()` output to `loo::loo_compare()`. Below, we fit a model that only includes demographic variables as predictors to compare to our full model.

```{r compare, eval = run}
set.seed(2026)
demographics <- bartisan(death ~ rhc + age + sex + race + edu,
                         data = rhc, family = binomial())

loo_compare(list(full = loo(full),
                 demographics = loo(demographics)))
```

The full model predicts better by around six times the standard error of the difference. The standard error is the important half: a difference of many standard errors is clear, and a difference smaller than its own standard error is not evidence of anything.

`p_worse` reports that reading as a probability. It is `pnorm(0, elpd_diff, se_diff)`, the chance that the model on that row is really the worse of the two given how far apart they came out and how precisely the difference is known. The best-ranked model has nothing to be compared against and gets `NA`. Two properties are worth knowing before reading one. Because the models are sorted by `elpd_loo` before it is computed, every reported value is at least .5 by construction, so .5 does not mean "even odds after weighing the evidence" but "the ranking is arbitrary and another sample could reverse it"; and because it comes from the normal approximation behind `se_diff`, it inherits that approximation's failures, which is why *loo* flags them in `diag_diff` and `diag_elpd` when it detects them. Here it is 1.00, and the reading is that a sample like this one would essentially never put the demographic model ahead.

### Totals, Averages, and Why the Convention Is a Total

`elpd_diff` is a difference of *totals*, and the reason is that a log score is additive. Summing it over observations gives a log predictive likelihood, so the difference between two models is a log likelihood ratio: the evidence the sample carries about which model predicts better, in nats. On that reading the total is the quantity with meaning, and it should grow with the sample, because more data is more evidence. It also puts `elpd` on the scale AIC, WAIC and DIC are quoted on, which are likewise totals.

Nothing is lost by preferring the average, because the decision does not depend on the choice. The mean difference and the total differ by a factor of $n$, its standard error differs by the same factor, and the ratio of the two, which is what says whether the difference is real, is identical either way.

What the average is better for is a comparison whose two halves do not sum over the same observations. The section above ran into exactly that: `elpd_loo` totals 1500 terms and the held-out score totals 300, so the totals are not comparable and the per-observation averages are. The same is true of a score quoted across datasets, or across subgroups of different sizes. Within one `loo_compare()` call it never arises, since every model there is scored on the same rows, which is why the convention can afford to be a total.

### The Same Comparison Through the Split

When the Pareto diagnostics say leave-one-out cannot be trusted, this comparison is the one to rebuild on held-out data. Both candidates are fitted to the training half and scored on the observations neither of them saw:

```{r holdoutcompare, eval = run}
demographics_train <- bartisan(death ~ rhc + age + sex + race + edu,
                               data = train, family = binomial())

score_demographics <- predict(demographics_train, newdata = held,
                              type = "density", log = TRUE)

d <- score - score_demographics

c(mean_diff = mean(d), se = sd(d) / sqrt(length(d)),
  ratio = mean(d) / (sd(d) / sqrt(length(d))))
```

That is `loo_compare()`'s arithmetic done by hand and quoted per observation: the average difference in log score, its standard error, and the ratio that decides whether to believe it. The full model is ahead by a little over two standard errors rather than six, which is what the split costs. It sees 300 observations where `loo()` sees 1500, and the precision of a difference goes as the square root of the count.

This is the right way to ask whether a set of variables earns its place, and it is a better question than the one variable importance answers, because it is about prediction rather than about how the forest happened to spend its splits. Here the answer is not in doubt by either route: how sick a patient is on arrival predicts whether they die, and demographics alone do not.

## Comparing Links (`family`)

The other choice is the likelihood. For a binary outcome the family is settled, and what remains is the link (i.e., the function that maps the forest's output onto a probability). Below we compare probit and complimentary log-log BART models to our original logistic BART model.

```{r links, eval = run}
set.seed(2026)
probit <- bartisan(model, data = rhc, family = binomial("probit"))

set.seed(2026)
cloglog <- bartisan(model, data = rhc, family = binomial("cloglog"))

loo_compare(list(logit = loo(full),
                 probit = loo(probit),
                 cloglog = loo(cloglog)))
```

The three are within a point or two of each other, and the differences are smaller than their standard errors, which *loo* flags directly. The reading is that the link does not matter here.

That is a useful negative result and worth reporting as one. It is also the usual outcome: with a flexible function on the inside, the link has little left to do, because the forest can absorb the difference between one link and another. This is not true of a generalized linear model, where the link carries the whole shape of the relationship.

## The Scale of the Log Density

`loo()` is built on the pointwise log density of each observation, so comparing two models by it assumes both densities are taken with respect to the same measure. That is usually automatic. Two different things break it, and each is repaired by a constant added to every pointwise contribution.

### A Change of Variable (`ph()` Against an AFT Family)

The accelerated failure time families report the density of $\log T$, while `ph()` reports the density of $T$. Below we fit one of each to the survival outcome in these data and compare them as they come out:

```{r survfits, eval = run && has_surv}
library(survival)

surv_model <- Surv(days, death) ~ age + sex + race + edu + aps + meanbp + surv2m

set.seed(2026)
aft <- bartisan(surv_model, data = rhc, family = weibull_aft())

set.seed(2026)
prop_haz <- bartisan(surv_model, data = rhc, family = ph())

loo_compare(list(aft = loo(aft), ph = loo(prop_haz)))
```

The accelerated failure time model appears to win by thousands of points, which is implausible on its face: no choice of baseline hazard is worth that much. What the comparison is measuring is the change of variable.

Writing $Y = \log T$, the density of $T$ is $f_T(t) = f_Y(\log t) / t$, so the log density on the $T$ scale is $\log f_Y(\log t) - \log t$. Subtracting $\log t$ from each of the accelerated failure time model's pointwise contributions is the whole of the correction, and it applies to events only: a censored observation contributes a survival probability, which is a probability on either scale and carries no measure to change.

`scale` asks `loo()` for it. Naming one measure for every model in the comparison is enough, because a fit already on the scale named is returned untouched:

```{r survjacobian, eval = run && has_surv}
loo_compare(list(aft = loo(aft, scale = "time"),
                 ph = loo(prop_haz, scale = "time")))
```

The ordering reverses. Read on the scale they share, the two are a hundred-odd points apart rather than three thousand, and it is the proportional hazards model that predicts these times better.

The correction is not applied without being asked for, because `loo()` would then stop reporting the model's own predictive density: it would no longer agree with `log_lik()`, and a comparison against a proportional hazards fit from some other package would quietly take on the error the correction exists to remove.

Which of the two scales is named does not matter, since a constant per observation cancels from a difference. Asking for `"log_time"` moves both totals and leaves the gap where it was:

```{r survboth, eval = run && has_surv}
c(on_time = elpd(loo(prop_haz, scale = "time")) - elpd(loo(aft, scale = "time")),
  on_log_time = elpd(loo(prop_haz, scale = "log_time")) -
    elpd(loo(aft, scale = "log_time")))
```

The everyday version of this mistake takes the same repair by hand. A model of `log(y)` reports $\log f_Y(\log y)$, so subtracting $\log y$ from each of its pointwise contributions, with `loo(sweep(log_lik(fit), 2, log(y), "-"))`, puts it on the scale of a model of `y`.

There is also a way to sidestep the correction rather than apply it. Survival at a horizon is a probability under every family, so comparing fits on `predict(type = "survival")` needs no Jacobian at all; `vignette("survival")` gives that route, along with the same correction written for a held-out log score instead of for `loo()`.

### A Point Mass Against a Density (`gaussian()` Against `tweedie()`)

The second break is not a transformation. `tweedie()` places a point mass at zero and a density on the positive half-line; `gaussian()` places a density on the whole line. Earnings in the `lalonde` data are the case this comes up in, since a quarter of the sample earned exactly nothing.

```{r tweediefits, eval = run && has_cobalt}
data("lalonde", package = "cobalt")

earnings <- re78 ~ age + educ + race + married + nodegree + re74 + re75

set.seed(2026)
normal <- bartisan(earnings, data = lalonde, family = gaussian())

set.seed(2026)
compound <- bartisan(earnings, data = lalonde, family = tweedie())

loo_compare(list(gaussian = loo(normal), tweedie = loo(compound)))
```

The tweedie is ahead by more than a thousand points. Splitting the pointwise contributions at the zeros shows where every one of them comes from:

```{r tweediesplit, eval = run && has_cobalt}
zero <- lalonde$re78 == 0

rbind(gaussian = c(at_zero = sum(loo(normal)$pointwise[zero, "elpd_loo"]),
                   positive = sum(loo(normal)$pointwise[!zero, "elpd_loo"])),
      tweedie  = c(at_zero = sum(loo(compound)$pointwise[zero, "elpd_loo"]),
                   positive = sum(loo(compound)$pointwise[!zero, "elpd_loo"])))
```

At the positive outcomes, where both models report a density with respect to the same measure, the two are within a couple of points of each other. The whole difference sits at the exact zeros, and the two quantities being added up there are not the same kind of thing:

```{r tweediezero, eval = run && has_cobalt}
c(gaussian = predict(normal, newdata = lalonde[zero, ][1, ], type = "density"),
  tweedie  = predict(compound, newdata = lalonde[zero, ][1, ], type = "density"))
```

The tweedie's value is a probability, bounded by one. The gaussian's is a density, in units of one over dollars. Adding either to log densities from the positive observations gives a total in no units at all, and the sign that something is wrong is that the total then depends on what the outcome is measured in. Refitting both on earnings in thousands moves the gap by hundreds of points without changing a single fact about the data:

```{r tweedieunits, eval = run && has_cobalt}
thousands <- transform(lalonde, re78 = re78 / 1000,
                       re74 = re74 / 1000, re75 = re75 / 1000)

set.seed(2026)
normal_k <- bartisan(earnings, data = thousands, family = gaussian())

set.seed(2026)
compound_k <- bartisan(earnings, data = thousands, family = tweedie())

c(dollars   = elpd(loo(compound))   - elpd(loo(normal)),
  thousands = elpd(loo(compound_k)) - elpd(loo(normal_k)))
```

A density has units of one over the outcome, so dividing earnings by a thousand multiplies every gaussian density by a thousand, and all 614 of them move together. The tweedie's positive observations move with them, but its 143 atoms do not, because a probability has no units to change. The gap shifts by the difference between them, $143 \log 1000 \approx 988$, up to the noise in refitting.

That also says how to repair it. The atom needs to be a density before it can be added to densities, and a probability becomes a density when it is spread over the width the outcome is recorded to: earnings to the nearest dollar put $P(Y = 0)$ over an interval one dollar wide, so the density there is $P(Y = 0) / \delta$ and the log density is $\log P(Y = 0) - \log \delta$. That is a constant per observation, exactly like the Jacobian above, and it makes the comparison well posed:

```{r tweediefix, eval = run && has_cobalt}
on_grid <- function(fit, resolution) {
  elpd(loo(fit)) - sum(zero) * log(resolution)
}

c(dollars   = on_grid(compound, 1)     - elpd(loo(normal)),
  thousands = on_grid(compound_k, .001) - elpd(loo(normal_k)))
```

The two now agree, and the tweedie is ahead by the same amount whichever units the outcome was recorded in. In dollars the correction happens to be zero, since $\log 1 = 0$, which is why the raw comparison looked reasonable and was right by coincidence rather than by construction.

Note what $\delta$ is and is not. It is a fact about the data, the resolution the outcome was recorded at, and it has to be supplied because no fitted model knows it; it is not a tuning knob, and choosing it to favor a model is choosing an answer. Where the outcome really is continuous and the zeros are exact rather than rounded, there is no $\delta$ to use and the comparison stays ill posed, which is the case the restricted comparison above covers instead.

Even with the correction in hand, an `elpd` is a blunt way to settle whether the outcome needs a point mass, because it answers with one number where the question is about the shape of the predictive distribution. A posterior predictive check answers it directly:

```{r tweediecheck, eval = run && has_cobalt}
zero_share <- function(y) mean(y == 0)

c(observed = zero_share(lalonde$re78),
  gaussian = mean(apply(rstantools::posterior_predict(normal), 1, zero_share)),
  tweedie  = mean(apply(rstantools::posterior_predict(compound), 1, zero_share)))
```

A gaussian fit never produces an exact zero, and about a quarter of these outcomes are exactly zero. `vignette("families")` covers the families that carry a point mass and `vignette("diagnostics")` covers the checks.

The rule behind both halves of this section is that `loo()` compares models whose densities are taken with respect to the same measure. A transformation of the outcome is undone with its Jacobian, and a point mass is reconciled with a density by the resolution the outcome is recorded at. Both are a constant added to each pointwise contribution, and in both cases the tell that one is needed is that the comparison moves when something about the outcome that should not matter changes.

## Comparing Against a Bayesian GLM

Often it is a good idea to compare a flexible model to a more easily interpretable (generalized) linear model to assess whether the flexibility buys us anything. There is no obstacle to this, and it is worth doing. If a logistic regression predicts as well as the forest, that is evidence the relationship is close to linear on the log-odds scale, and the simpler model is the easier one to report.

The comparison has to be like for like, which means both models have to produce a pointwise log density of the same outcome on the same scale. Fitting the regression in a Bayesian framework is what arranges that: *rstanarm* fits it with `stan_glm()` and gives it a `loo()` method, and the resulting object goes into `loo_compare()` beside ours.

```{r stanglm, eval = run && has_rstanarm}
logistic <- rstanarm::stan_glm(model, data = rhc, family = binomial(),
                               chains = 4, refresh = 0, seed = 2026)

loo(logistic)
```

```{r stancompare, eval = run && has_rstanarm}
loo_compare(list(bart = loo(full), logistic = loo(logistic)))
```

The forest is behind by roughly half a standard error of the difference, which is to say the two predict this outcome equally well[^1]. Nothing was lost by fitting the forest and nothing was gained, and the honest report of that is the one above: the flexible model was tried and did not find anything the linear one missed.

[^1]: `loo_compare()` warns that the models do not have the same `y` variable, which is not what has happened here. It compares a hash of the response that *rstanarm* and *brms* attach to their `loo()` output and this package does not, so it is holding a hash up against nothing. Attaching one would not help: the hash is taken of the response as each package stores it, and an outcome held as `integer` on one side and `double` on the other hashes differently, so the warning would start firing on identical data rather than stop firing. What we should check instead, and directly, is that both models were fitted to the same rows.

`p_loo` says where that came from. The regression has 16 coefficients and an effective number of parameters to match; the forest's is about twice that, which is the flexibility it spent looking for curvature and interaction that turned out not to be there. A forest that predicts no better while using twice the parameters is a forest reporting that the log-odds are close to linear here, which is a finding rather than a disappointment.

Comparing `elpd_loo` from this package against `AIC()` from `glm()` is not a comparison and should not be reported as one. The two differ by a factor of $-2$ before anything else, and `AIC()`'s penalty is a count of parameters, which a forest has no fixed number of; the `p_loo` above is an estimate rather than a count. Fitting the regression the Bayesian way is what lets us compare `loo()` against `loo()`, as above.

## Tuning

The number of trees, `k`, and the gate should not be chosen by cross-validation as a matter of routine. The priors are chosen so that the defaults work across a wide range of problems, and tuning them typically produces small gains together with an optimistically biased estimate of performance when the same data chose the setting.

When there is a reason to tune, the shape of it is a grid fixed in advance, chosen on one set of observations and assessed on another. The split from the earlier section is already in hand, so the grid goes through `loo()` on the training set:

```{r tunegrid, eval = run}
trees <- c(20, 50, 200)

tuned <- lapply(trees, function(n) {
  set.seed(2026)
  bartisan(model, data = train, family = binomial(), num_trees = n)
})

names(tuned) <- paste0("trees_", trees)

loo_compare(lapply(tuned, loo))
```

Nothing separates them: every difference is smaller than its own standard error, and *loo* flags both comparisons as too small to read. Scoring the same three fits on the held-out observations says it again, and is the assessment the selection is not allowed to see:

```{r tunedheld, eval = run}
sapply(tuned, function(f) sum(predict(f, newdata = held, type = "density",
                                      log = TRUE)))
```

The two orderings disagree, which is the practical content of the warning rather than a contradiction: when the spread across a grid is smaller than the noise in the estimate, whichever setting comes out top is the one the noise favored. Reporting its score as the model's performance is exactly the optimism this section opened with, and it is why the assessment has to come from observations the grid never touched. Here the honest summary is that `num_trees` does not matter on these data, and the default is what to keep.

That is the usual outcome, and it is worth knowing before spending a grid on it. When tuning does pay, it tends to be for cost rather than accuracy, and `vignette("faq")` lists the settings that buy speed; `bartisan_control()` documents what each one changes.

## Variable Selection

Variables should not be selected by fitting many models and keeping the best. With a search over all subsets, the winner is chosen partly for fitting the noise, for the same reason stepwise regression is not to be trusted.

Two things work instead, and which one to use depends on the question. When the question is whether a *set* of variables earns its place, the comparison is the one from earlier in this vignette: name the specifications in advance and put them through `loo_compare()`, as with the full and demographic models above. A handful of comparisons decided beforehand is informative in a way a search is not.

When the question is which *individual* predictor matters, dropping one at a time and reading `loo()` is a weak instrument. It works for a dominant predictor and fails for the rest:

```{r dropone, eval = run}
drop_one <- function(v) {
  set.seed(2026)
  reduced <- bartisan(update(model, paste(". ~ . -", v)), data = rhc,
                      family = binomial())

  d <- loo(full)$pointwise[, "elpd_loo"] - loo(reduced)$pointwise[, "elpd_loo"]

  c(elpd_diff = sum(d), se_diff = sqrt(length(d)) * sd(d))
}

rbind(surv2m = drop_one("surv2m"), rhc = drop_one("rhc"))
```

Losing `surv2m`, the prognostic score, costs about forty points and is plainly detectable. Losing `rhc` costs about two, which is smaller than its own standard error, and yet `rhc` is real enough that `vignette("causal")` puts its effect at around six percentage points on the probability of death. The remaining predictors carry that information too, so the forest reroutes around the one that was removed, and a comparison of predictive density cannot see a variable whose job something else can do. Running one such comparison per predictor would also be the search this section began by warning against.

So that question belongs to `variable_importance()` and to the sparsity prior, which does the selection inside the model rather than across refits. `vignette("importance")` covers reading it, including the noise-predictor calibration that says which end of the table carries information, and `vignette("effects")` covers the step after: asking what a predictor does to the outcome rather than whether it is used.

## Further Reading

`vignette("diagnostics")` covers checking that each model fits before comparing them, which is the step most often skipped. `vignette("families")` covers what the families assume, which is what a comparison between them is really about.

## References
