---
title: "Choosing between models"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Choosing between models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{=html}
<style>
pre {overflow-x: auto;}
pre code {word-wrap: normal; white-space: pre;}
</style>
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.2, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 4)

has_loo <- rlang::is_installed("loo")
run <- has_loo
```

## Introduction

Model selection means something different for a BART model than it does for a
linear model, and the difference is worth stating before any code.

With `lm()` we choose which terms enter, whether to add a squared term, and
whether to include an interaction; those choices are the model. With BART the
forest makes them, so they are not ours to make. What remains is a shorter list:

1. Which variables the model is allowed to see.
2. Which likelihood, meaning which family and link.
3. Occasionally, a sampler setting (e.g., the number of trees).

In this guide we will compare those choices using leave-one-out
cross-validation. First we'll fit a model to the `rhc` data and read its `loo()`
output, including the diagnostics that say when the approximation cannot be
trusted. Next we'll compare two sets of variables and then three links, and
we'll work through the one case in this package where two log densities are not
on the same scale. Finally we'll cover the choices that should not be made by
cross-validation at all.

```{r libs, eval = run}
library(bartisan)
library(loo)

data("rhc")

model <- death ~ rhc + age + sex + race + edu + aps + meanbp + resp +
  hema + pafi + paco2 + crea + surv2m + card

set.seed(2026)

full <- bartisan(model, data = rhc, family = binomial())
```

The fits here use the default single chain rather than the four used elsewhere;
leave-one-out needs draws rather than chains, and four times as many fits would
make this vignette slow to build for no gain. Convergence should still be
checked separately, as in `vignette("diagnostics")`, before we trust any of
these comparisons.

## Leave-One-Out Cross-Validation (`loo()`)

```{r loo, eval = run}
loo(full)
```

`elpd_loo` estimates the log predictive density on data the model has not seen,
with higher values being better. It is computed by importance sampling from the
fitted posterior rather than by refitting, which is why it is fast
[@vehtari2017].

`p_loo` is the effective number of parameters, about 32 here; for a forest with
hundreds of leaves across its trees that number is small, because the prior
shrinks most of them toward zero. It is a useful measure of how much of the data
the model is actually using.

The Pareto $k$ diagnostics are the thing to check: leave-one-out by importance
sampling is trustworthy only when the importance weights are well behaved, and a
$k$ above .7 says that for that observation they are not. Here they are all
good.

### When the Approximation Fails

A forest is a flexible function, so a single observation can have a good deal of
influence on the leaves it falls into, and high $k$ values are therefore more
common than they are for a parametric model. If many observations are flagged,
the estimate is unreliable, and the remedy is held-out data rather than a
different diagnostic:

```{r holdout, eval = FALSE}
train_id <- sample.int(nrow(rhc), 1200)

train <- rhc[train_id, ]
held  <- rhc[-train_id, ]

fit <- bartisan(model, data = train, family = binomial())

sum(predict(fit, newdata = held, type = "density", log = TRUE))
```

This is the same quantity `elpd_loo` approximates, computed directly.

## Comparing Two Models (`loo_compare()`)

```{r compare, eval = run}
set.seed(2026)
demographics <- bartisan(death ~ rhc + age + sex + race + edu,
                         data = rhc, family = binomial())

loo_compare(list(full = loo(full),
                 demographics = loo(demographics)))
```

The full model predicts better by around six times the standard error of the
difference. The standard error is the important half: a difference of many
standard errors is clear, and a difference smaller than its own standard error
is not evidence of anything.

This is the right way to ask whether a set of variables earns its place, and it
is a better question than the one variable importance answers, because it is
about prediction rather than about how the forest happened to spend its splits.
Here the answer is not in doubt: how sick a patient is on arrival predicts
whether they die, and demographics alone do not.

## Comparing Links (`family`)

The other choice is the likelihood. For a binary outcome the family is settled,
and what remains is the link (i.e., the function that maps the forest's output
onto a probability).

```{r links, eval = run}
set.seed(2026)
probit <- bartisan(model, data = rhc, family = binomial("probit"))

set.seed(2026)
cloglog <- bartisan(model, data = rhc, family = binomial("cloglog"))

loo_compare(list(logit = loo(full),
                 probit = loo(probit),
                 cloglog = loo(cloglog)))
```

The three are within a point or two of each other, and the differences are
smaller than their standard errors, which *loo* flags directly. The reading is
that the link does not matter here.

That is a useful negative result and worth reporting as one. It is also the
usual outcome: with a flexible function on the inside, the link has little left
to do, because the forest can absorb the difference between one link and
another. This is not true of a generalized linear model, where the link carries the whole
shape of the relationship.

## The Scale of the Log Density

`loo()` is built on the pointwise log density of each observation, so comparing
two models by it assumes that both densities are taken with respect to the same
measure. That is normally automatic, but there is one case in this package where
it is not.

The accelerated failure time families report the density of $\log T$, while
`ph()` reports the density of $T$. The two differ by a Jacobian, so a comparison
across that boundary is off by $\sum \log t$, which can amount to hundreds of
points and can reverse the ordering. `vignette("survival")` gives the
correction.

The general rule is that `loo()` compares models fitted to the same outcome on
the same scale. If we transform the outcome, refit, and compare, the comparison
is invalid unless the Jacobian is accounted for; comparing a model of `y` to a
model of `log(y)` is the everyday version of this mistake.

## Comparing Against a Logistic Regression

There is no obstacle to this, and it is worth doing. If a logistic regression
predicts as well as the forest, that is evidence the relationship is close to
linear on the log-odds scale, and the simpler model is easier to report.

The comparison has to be like for like, which means fitting both models in a
framework that produces the same kind of pointwise log density: we can fit the
logistic regression with *rstanarm* or *brms* and compare the two with
`loo_compare()`, or compare held-out log scores computed the same way for both.

Comparing `elpd_loo` from this package against `AIC()` from `glm()` is not a
comparison and should not be reported as one.

## Tuning and Variable Selection

The number of trees, `k`, and the gate should not be chosen by cross-validation
as a matter of routine. The priors are chosen so that the defaults work across a
wide range of problems, and tuning them typically produces small gains together
with an optimistically biased estimate of performance when the same data chose
the setting. When we do tune, data should be held out for the final assessment.

Nor should variables be selected by fitting many models and keeping the best.
With a handful of candidate specifications chosen in advance, `loo_compare()` is
informative; with a search over all subsets it is not, for the same reason
stepwise regression is not.

## Further Reading

`vignette("diagnostics")` covers checking that each model fits before comparing
them, which is the step most often skipped. `vignette("families")` covers what
the families assume, which is what a comparison between them is really about.

## References
