---
title: "Has it converged, and does it fit?"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Has it converged, and does it fit?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.0, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 3)

has_post <- rlang::is_installed("posterior")
has_bp <- rlang::is_installed("bayesplot")
has_gg <- rlang::is_installed("ggplot2")
run <- TRUE
```

## Introduction

Two questions get asked together and are worth separating. Has the sampler
converged, meaning has it explored the posterior properly? And does the model
fit, meaning does it describe the data?

The first is about the algorithm and the second is about the model. A fit can
converge beautifully on a badly chosen family, and a well chosen family can be
fitted by a chain that has not run long enough.

```{r libs, eval = run}
library(bartisan)

data(rhc)

model <- death ~ rhc + age + sex + race + edu + aps + meanbp + resp +
  hema + pafi + paco2 + crea + surv2m + card

set.seed(2026)

fit <- bartisan(model, data = rhc, family = binomial(), chains = 4)
```

## Convergence

### One call for all of it

`diagnose()` computes every convergence and mixing statistic in this section,
says which of them fall short, and says what to do about each. Everything it
reports comes from the stored draws, so it needs no other package.

```{r diagnose, eval = run}
diagnose(fit)
```

The rest of this section is what it is reporting and why, which is worth reading
once. Skip to `vignette("bartisan")` if the summary is enough.

### Run more than one chain

`chains = 4` above is doing the work. The default is one chain, which produces
estimates but no way to check them: `fit$rhat` is `NULL` for a single chain,
because the statistic compares chains to each other. Running four costs four
times as much, which for most fits is a few seconds.

```{r rhat, eval = run}
fit$rhat
```

`rhat` compares the variance between chains to the variance within them
[@vehtari2021]. If the chains have found the same posterior the two agree and the
ratio is near 1. `ess_bulk` and `ess_tail` are effective sample sizes: how many
independent draws the correlated ones are worth, in the middle of the
distribution and in the tails.

The conventional thresholds are `rhat` below 1.01 and effective sample sizes above
about 400 for a quantity you intend to report precisely. Those come from the
general MCMC literature and are a reasonable default here, with one exception
described below. `diagnose()` applies them, and takes both as arguments if you
want to move them.

`diagnose()` adds two things this table does not have. It repeats `rhat` on the
second half of the retained draws alone, which is what tells a warmup that ended
too early from chains that have each settled somewhere different: throwing away
the early draws is exactly what more `num_burn` would have done, so if that fixes
`rhat`, warmup was the problem. And it reports the total number of splitting
rules in the forest at each draw, which is the one quantity here that is about
the trees rather than about the fitted values, and which no general-purpose MCMC
diagnostic would think to look at.

### What the rows are

`loglik` is the log likelihood of the whole dataset at each draw. It is a useful
scalar summary of the fit and mixes reasonably.

`aux.*` are the nuisance parameters of the family, and are usually the best
behaved rows in the table. A binomial likelihood has none, so they do not appear
here; a Gaussian fit would show `aux.sigma`, and a survival fit its scale or
baseline hazard.

`eta.*` is the additive predictor, summarized over the observation that mixes
worst. This is the fitted function, so it is the row that matters when
predictions or effects are the output.

One quantity is deliberately not in the table. The leaf prior scale, in
`fit$sigma_mu`, mixes badly in every BART implementation: on one dataset its
counterparts in *dbarts* and in *stochtree* come out at `rhat` 1.12 and 1.16
with effective sample sizes of 22 and 17 out of 3000 draws, and *stochtree*
draws it exactly from its full conditional, so no sampler can do better. The
*BART* package avoids the question by never drawing it. It is a hyperparameter
whose disagreement between chains does not reach the fitted function, which on
those same fits has `rhat` 1.00 and thousands of effective draws, so reporting it
beside `eta` only produced an alarming row with nothing to act on. It is out of
this table and only out of this table: the draws are in `fit$sigma_mu` and reach
`as_draws()`, so it can still be diagnosed by anyone who wants to.

### The one exception

The `eta` row is a maximum over every observation, so it is conservative by
construction, and forests mix slowly on their fitted values. Two chains can visit
quite different collections of trees that imply nearly identical predictions,
which inflates a between-chain statistic without meaning the two chains disagree
about anything you would report.

On harder problems this shows up clearly. The following fits the Friedman
function, first with a chain far too short and then at the defaults.

```{r friedmanbad}
set.seed(3)
fr <- as.data.frame(matrix(runif(400 * 10), 400, 10))
names(fr) <- paste0("x", 1:10)
fr$y <- 10 * sin(pi * fr$x1 * fr$x2) + 20 * (fr$x3 - 0.5)^2 +
  10 * fr$x4 + 5 * fr$x5 + rnorm(400)

too_short <- bartisan(y ~ ., fr, family = gaussian(), chains = 4,
                      control = bartisan_control(num_trees = 20, num_burn = 50,
                                                 num_draws = 50))
too_short$rhat
```

Everything is bad: `rhat` above 2 for the log likelihood, the residual standard
deviation and the fitted values, with effective sample sizes under 10. This chain
has not converged and nothing from it should be used.

```{r friedmangood}
long_enough <- bartisan(y ~ ., fr, family = gaussian(), chains = 4)

long_enough$rhat
```

Now the log likelihood and the residual standard deviation have converged, and
`eta` has improved but is still above 1.01. That residual is the phenomenon
described above, and adding draws does not reliably remove it.

When the `eta` row is elevated, look at the distribution behind it rather than
the maximum:

```{r perobs, eval = run}
per_observation_rhat <- function(fit) {
  draws <- fit$eta$eta
  per <- nrow(draws) / fit$chains
  index <- matrix(seq_len(per * fit$chains), per, fit$chains)
  apply(draws, 2, function(column) {
    posterior::rhat(matrix(column[index], per, fit$chains))
  })
}

round(quantile(per_observation_rhat(fit), c(0.5, 0.9, 0.99, 1)), 3)
```

The median is a little above 1 and so is most of the distribution, so the
maximum in the table is not one badly behaved observation: the fitted function as
a whole mixes slowly here. That is the phenomenon described above rather than a
broken chain, and the two are told apart by what happens to the quantities you
report. If the log likelihood has converged, along with any nuisance parameters the
family has, and an `avg_comparisons()` estimate is stable across chains and
across a longer run, the fit is usable.

If instead the log likelihood were badly elevated too, or the estimate moved when
the chain was lengthened, that would be a chain that has not converged, and the
answer is more draws or a simpler model.

### Looking at the chains

`as_draws()` hands the fit to *posterior* and *bayesplot*. It carries the
scalar parameters and a representative set of `eta` columns.

```{r summarise, eval = run && has_post}
library(posterior)

summarise_draws(as_draws(fit))
```

```{r trace, eval = run && has_bp, fig.height = 2.6}
library(bayesplot)

mcmc_trace(as_draws(fit, eta = 1), pars = c("loglik", "eta[1]"))
```

What you want to see is four chains overlapping, wandering around the same level,
with no drift and no long excursions. `eta = 1` selects the first observation;
`eta = TRUE` gives a spread of ten and `eta = FALSE` gives none.

### What to do when it has not converged

In order of what usually helps:

1. Increase `num_burn` and `num_draws`. This is the first thing to try and it
   fixes most cases.
2. Reduce `num_trees`. A smaller forest has fewer ways to represent the same
   function and mixes faster.
3. Check the family. A likelihood that fits the data badly can produce a
   posterior that is hard to explore.

## Fit

Convergence says the sampler did its job. It says nothing about whether the model
is right.

### Posterior predictive checks

Simulate outcomes from the fitted model and compare their distribution to the
observed one.

```{r ppc, eval = run && has_bp, fig.height = 3.0}
pp_check(fit)
```

For a continuous outcome this is the workhorse check, and systematic differences
are what to look for: replicates that are too narrow, that miss a second mode, or
that put mass where the outcome cannot go. Simulating negative values for an
outcome that cannot be negative says the family is wrong, and
`vignette("families")` covers the alternatives.

For a **binary** outcome it is a weak check, which is worth knowing before
reading too much into it. There are only two values the replicates can take, so
they will match the observed proportion unless the model has gone badly wrong.
Passing this tells you almost nothing.

### Calibration

The useful question for a binary outcome is whether the predicted probabilities
mean what they say. Group the patients by their fitted probability and compare
the average prediction in each group with the proportion who actually died.

```{r calibration, eval = run && has_gg, fig.height = 3.4}
library(ggplot2)

p <- fitted(fit)
bins <- cut(p, quantile(p, seq(0, 1, 0.1)), include.lowest = TRUE)

calibration <- data.frame(
  predicted = tapply(p, bins, mean),
  observed  = tapply(rhc$death, bins, mean)
)

ggplot(calibration, aes(predicted, observed)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60") +
  geom_point(size = 2) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "mean predicted probability", y = "observed proportion") +
  theme_bw(base_size = 9)
```

Points on the diagonal mean the probabilities are calibrated: among patients the
model gave a 30% chance of dying, about 30% died. These sit close to it across
the whole range.

Points systematically above the line at the left and below it at the right would
mean the predictions are too extreme, which is the usual failure of an
overfitted model. The opposite pattern means they are too timid, which is what
heavy shrinkage produces.

This is an in-sample check and so is optimistic. For an honest version, fit on
part of the data and calibrate on the rest.

### Residuals

For a continuous outcome, plot residuals against fitted values and read it the
way you would for a linear model, with one difference. A forest shrinks its
predictions toward the overall mean, so a mild negative trend is expected even
when the model is correct: the highest fitted values are pulled down and the
lowest pulled up. A strong slope is not expected, and a fan shape means the
spread of the outcome depends on the predictors, which `location_scale()` models
directly.

For a binary outcome the residuals take two values for any given fitted
probability and the plot is not informative. Use the calibration check instead.

### How much is the model explaining

```{r r2, eval = run && rlang::is_installed("performance")}
performance::r2(fit)
```

This is the Bayesian $R^2$ of @gelman2019, computed from the posterior rather
than from a single fit, so it comes with an interval and cannot exceed 1. For a
binary outcome it is bounded well below 1 by the outcome's own randomness, so
read it as a relative measure rather than against any absolute standard.

## A checklist

Before reporting anything from a fit:

1. Fit with `chains = 4`.
2. Run `diagnose(fit)` and read what it says to change. It covers steps 2 and 3
   of the older advice: the thresholds, which rows are exempt, and whether a
   short warmup or genuinely disagreeing chains is the problem.
3. Run `pp_check()` and look for systematic differences.
4. If the outcome is bounded, check that the replicates respect the bound.

## Where to go next

`vignette("comparison")` covers choosing between models once each of them fits.
`vignette("families")` covers what to do when the posterior predictive check says
the family is wrong.

## References
