---
title: "Has it converged, and does it fit?"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Has it converged, and does it fit?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.0, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 3)

has_post <- rlang::is_installed("posterior")
has_bp <- rlang::is_installed("bayesplot")
has_gg <- rlang::is_installed("ggplot2")
run <- rlang::is_installed("MASS")
```

## Introduction

Two questions get asked together and are worth separating. Has the sampler
converged, meaning has it explored the posterior properly? And does the model
fit, meaning does it describe the data?

The first is about the algorithm and the second is about the model. A fit can
converge beautifully on a badly chosen family, and a well chosen family can be
fitted by a chain that has not run long enough.

```{r libs, eval = run}
library(bartisan)

data("birthwt", package = "MASS")

d <- birthwt
d$race  <- factor(d$race,  levels = 1:3, labels = c("white", "black", "other"))
d$smoke <- factor(d$smoke, levels = 0:1, labels = c("no", "yes"))
d$ht    <- factor(d$ht,    levels = 0:1, labels = c("no", "yes"))
d$ui    <- factor(d$ui,    levels = 0:1, labels = c("no", "yes"))
d$low   <- NULL

set.seed(2026)

fit <- bartisan(bwt ~ ., data = d, family = gaussian(), chains = 4)
```

## Convergence

### Run more than one chain

`chains = 4` above is doing the work. The default is one chain, which produces
estimates but no way to check them: `fit$rhat` is `NULL` for a single chain,
because the statistic compares chains to each other. Running four costs four
times as much, which for most fits is a few seconds.

```{r rhat, eval = run}
fit$rhat
```

`rhat` compares the variance between chains to the variance within them
[@vehtari2021]. If the chains have found the same posterior the two agree and the
ratio is near 1. `ess_bulk` and `ess_tail` are effective sample sizes: how many
independent draws the correlated ones are worth, in the middle of the
distribution and in the tails.

The conventional thresholds are `rhat` below 1.01 and effective sample sizes above
about 400 for a quantity you intend to report precisely. Those come from the
general MCMC literature and are a reasonable default here, with one exception
described below.

### What the rows are

`loglik` is the log likelihood of the whole dataset at each draw. It is a useful
scalar summary of the fit and mixes reasonably.

`aux.*` are the nuisance parameters, here the residual standard deviation. These
are usually the best behaved and are the ones a reported `sigma` depends on.

`sigma_mu.*` is a prior scale internal to the forest. It mixes slowly in almost
every BART fit and does not improve much with longer chains. Nothing you report
depends on it directly. It is in the table for completeness rather than as
something to act on.

`eta.*` is the additive predictor, summarized over the observation that mixes
worst. This is the fitted function, so it is the row that matters when
predictions or effects are the output.

### The one exception

The `eta` row is a maximum over every observation, so it is conservative by
construction, and forests mix slowly on their fitted values. Two chains can visit
quite different collections of trees that imply nearly identical predictions,
which inflates a between-chain statistic without meaning the two chains disagree
about anything you would report.

On harder problems this shows up clearly. The following fits the Friedman
function, first with a chain far too short and then at the defaults.

```{r friedmanbad}
set.seed(3)
fr <- as.data.frame(matrix(runif(400 * 10), 400, 10))
names(fr) <- paste0("x", 1:10)
fr$y <- 10 * sin(pi * fr$x1 * fr$x2) + 20 * (fr$x3 - 0.5)^2 +
  10 * fr$x4 + 5 * fr$x5 + rnorm(400)

too_short <- bartisan(y ~ ., fr, family = gaussian(), chains = 4,
                      control = bartisan_control(num_trees = 20, num_burn = 50,
                                                 num_save = 50))
too_short$rhat
```

Everything is bad: `rhat` above 2 for the log likelihood, the residual standard
deviation and the fitted values, with effective sample sizes under 10. This chain
has not converged and nothing from it should be used.

```{r friedmangood}
long_enough <- bartisan(y ~ ., fr, family = gaussian(), chains = 4)

long_enough$rhat
```

Now the log likelihood and the residual standard deviation have converged, and
`eta` has improved but is still above 1.01. That residual is the phenomenon
described above, and adding draws does not reliably remove it.

When the `eta` row is the only one left elevated, look at the distribution behind
it rather than the maximum:

```{r perobs, eval = run}
per_observation_rhat <- function(fit) {
  draws <- fit$eta$eta
  per <- nrow(draws) / fit$chains
  index <- matrix(seq_len(per * fit$chains), per, fit$chains)
  apply(draws, 2, function(column) {
    posterior::rhat(matrix(column[index], per, fit$chains))
  })
}

round(quantile(per_observation_rhat(fit), c(0.5, 0.9, 0.99, 1)), 3)
```

For the birth weight fit the whole distribution is below 1.05, so the maximum was
not hiding a problem. If instead the median were elevated, the fitted function
genuinely has not converged and the answer is more draws or a simpler model.

### Looking at the chains

`as_draws()` hands the fit to *posterior* and *bayesplot*. It carries the
scalar parameters and a representative set of `eta` columns.

```{r summarise, eval = run && has_post}
library(posterior)

summarise_draws(as_draws(fit))
```

```{r trace, eval = run && has_bp, fig.height = 2.6}
library(bayesplot)

mcmc_trace(as_draws(fit, eta = 1), pars = c("aux.sigma", "eta[1]"))
```

What you want to see is four chains overlapping, wandering around the same level,
with no drift and no long excursions. `eta = 1` selects the first observation;
`eta = TRUE` gives a spread of ten and `eta = FALSE` gives none.

### What to do when it has not converged

In order of what usually helps:

1. Increase `num_burn` and `num_save`. This is the first thing to try and it
   fixes most cases.
2. Reduce `num_trees`. A smaller forest has fewer ways to represent the same
   function and mixes faster.
3. Check the family. A likelihood that fits the data badly can produce a
   posterior that is hard to explore.

## Fit

Convergence says the sampler did its job. It says nothing about whether the model
is right.

### Posterior predictive checks

Simulate outcomes from the fitted model and compare their distribution to the
observed one.

```{r ppc, eval = run && has_bp, fig.height = 3.0}
pp_check(fit)
```

The simulated densities should look like plausible replicates of the observed
one. Here they do, with the observed data slightly heavier in the lower tail than
the replicates. Systematic differences are what to look for: replicates that are
too narrow, that miss a second mode, or that put mass where the outcome cannot go.

A common and informative failure is simulating negative values for an outcome
that cannot be negative, which says the Gaussian family is wrong and a positive
family is needed. `vignette("families")` covers the alternatives.

### Residuals

```{r resid, eval = run && has_gg, fig.height = 2.8}
library(ggplot2)

check <- data.frame(fitted = fitted(fit), residual = residuals(fit))

ggplot(check, aes(fitted, residual)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_point(alpha = 0.5, size = 1) +
  labs(x = "fitted value", y = "residual") +
  theme_bw(base_size = 9)
```

Read this the way you would for a linear model, with one difference. A forest
shrinks its predictions toward the overall mean, so a plot of residuals against
fitted values will often show a mild negative trend even when the model is
correct: the highest fitted values are pulled down and the lowest pulled up. A
gentle slope is expected. A strong one, or a fan shape, is not.

A fan shape means the spread of the outcome depends on the predictors, which
`location_scale()` models directly.

### How much is the model explaining

```{r r2, eval = run && rlang::is_installed("performance")}
performance::r2(fit)
```

This is the Bayesian $R^2$ of @gelman2019, computed from the posterior rather
than from a single fit, so it comes with an interval and cannot exceed 1.

## A checklist

Before reporting anything from a fit:

1. Fit with `chains = 4`.
2. Check `fit$rhat`. The `loglik` and `aux` rows should be near 1. If `eta` is
   the only elevated row, look at the per-observation distribution.
3. Run `pp_check()` and look for systematic differences.
4. If the outcome is bounded, check that the replicates respect the bound.

## Where to go next

`vignette("comparison")` covers choosing between models once each of them fits.
`vignette("families")` covers what to do when the posterior predictive check says
the family is wrong.

## References
