---
title: "Effects, Curves, and Interactions"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Effects, Curves, and Interactions}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{=html}
<style>
pre {overflow-x: auto;}
pre code {word-wrap: normal; white-space: pre;}
</style>
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.2, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 4, marginaleffects_safe = FALSE)

has_me <- rlang::is_installed("marginaleffects")
run <- has_me
```

## Introduction

A forest has no coefficients: there is no table of slopes to read, and no standard error to put beside one. This is the part of the workflow that changes most when moving from parametric models to BART. At the same time, this is the very reason to switch to BART: the absence of coefficients accompanies to the extreme flexibility BART has in modeling relationships without the analyst having to specify their form.

Though this may seem like a loss, modern approaches to characterizing variable relationships can support a model-agnostic workflow [@rohrerModelsPredictionMachines2026]. The replacement for coefficients is to ask the fitted model questions about predictions. What does it predict for these people? What would it predict if this variable were different? How much does the answer differ between groups? The [*marginaleffects*](https://marginaleffects.com/) package [@arelbundock2024] answers all of them, and returns posterior intervals with those answers.

Two of those questions *bartisan* answers itself: `estimate_effect()` considers one binary or factor treatment and reports the contrast as the average treatment effect or as one effect per unit, optionally within subgroups; `partial_dependence()` averages the fitted surface over the sample at each value of one or two predictors. `vignette("causal")` contains the worked version of the first. Everything else in this vignette uses *marginaleffects*: slopes, contrasts between covariate values, a continuous predictor, hypotheses comparing one estimate to another, and any grid that fixes the other covariates at chosen values rather than averaging over them.

This vignette covers the questions worth asking and how to phrase them. `vignette("bartisan")` is the shorter tour, and this expands its section on interpreting the fit.

In this guide, we will start from the three kinds of question the package answers and then work through them in turn. First, we'll consider average effects of categorical and numeric predictors, along with the prior setting (`sparsity`) that can quietly attenuate a contrast. Next we'll split those effects by subgroup and test a moderation with the difference of the two, then plot the shape of a fitted relationship and choose the scale on which an effect is reported, and close with what these estimates can and cannot be taken to mean.

Below, we load in the `rhc` dataset (see `vignette("bartisan")` or `help("rhc", package = "bartisan")` for details) and fit a BART logistic regression model predicting `death` from the available predictors. (Normally we would use multiple chains, perform diagnostics on mixing, and possibly modify the fitting parameters; we skip those here for brevity.)

```{r libs, eval = run}
library(bartisan)
library(marginaleffects)

data("rhc")

set.seed(2026)

# Fit a BART logistic regression model
fit <- bartisan(death ~ rhc + age + sex + race + edu + aps + meanbp + resp +
                  hema + pafi + paco2 + crea + surv2m + card,
                data = rhc, family = binomial())

fit
```

## Predictions, Comparisons, and Slopes

Everything below is one of the following:

A **prediction** is the outcome the model expects for a set of covariate values. `predictions()` gives one per unit or covariate profile, `avg_predictions()` averages them.

A **comparison** is the difference between two predictions that differ in one variable. This is the closest thing to a regression coefficient, and we often want to report it.

A **slope** is the derivative of the prediction with respect to a numeric variable. These are often also used to report but come with some limitations due to some peculiarities of using a tree-based model.

## Average Effects (`avg_comparisons()`)

`avg_comparisons()` with no `variables` argument gives every predictor at once, which for this model is a long table. A few at a time is easier to read, and often only a few effects are of primary interest anyway [@keeleCausalInterpretationEstimated2020]:

```{r avgcomp, eval = run}
comp <- avg_comparisons(fit, variables = c("rhc", "age", "card"))

comp
```

This reads as a coefficient table. Each estimate is an average difference in predicted probability, holding everything else at each patient's own values: for a factor, between the levels named in the `Contrast` column, and for a numeric predictor, for an increase of one unit. `rhc` is coded `0` and `1`, so its one-unit contrast is the treatment effect.

The comparison for `age`, for example, means that increasing all units' age by 1, keeping all other predictors at their observed values, would yield an increase in the overall probability of death of `r round(100*comp$estimate[1], 3)` percentage points (a very small effect on this scale), with a 95% credible interval excluding 0. This isn't a causal estimate; it's just what the model predicts would happen.

### The Splitting Prior and a Contrast (`sparsity`)

An estimate here can come back as exactly zero, and an interval bound with it; that is not a rounding artifact. The default splitting prior is a variable-selection prior (i.e., one that can leave a predictor out of the forest altogether), so in a draw where it uses the predictor in no tree, the prediction does not depend on it, and the contrast is exactly zero; the posterior of the contrast is a mixture with a point mass there, holding whatever share of draws dropped the predictor. *marginaleffects* centers a posterior at its median by default, so once that point mass holds half of it the reported estimate is exactly zero however far the rest of the posterior sits from zero. Setting `options(marginaleffects_posterior_center = mean)` asks for the mean instead, which is the how `predict()` and `estimate_effect()` report predictions.

This can matter more than it sounds: on a weak signal, the prior can attenuate the estimate substantially, and its interval can cover the truth well below its nominal rate. A strong effect is untouched, because the prior never has reason to drop a predictor that is earning its splits, so this is a weak-signal problem rather than a general one.

If we are reporting a contrast, we can fit the model with `sparsity = FALSE`, or by supplying `split_prior`, which fixes the weights (i.e., the probability that each predictor is chosen for a split) and so cannot drop anything. `?bartisan_control` explains both options.

### Choosing the Step (`variables`)

One unit is the default step size for a contrast but is often the wrong scale. One point of an illness score is a small change; ten points is a difference someone would notice. For example, the variable `aps` ranges from 4 to 147 in the sample, so we might request a comparison corresponding to increasing its value by 10.

```{r step, eval = run}
avg_comparisons(fit, variables = list(aps = 10))
```

A numeric effect should always be reported together with the step it was computed at. Unlike a linear model, the answer here is not ten times the one-unit effect because the relationship is not assumed to be a straight line. We'll see below other ways to characterize the relationship graphically.

## Effects for Subgroups (`by`)

Supplying `by` to `avg_comparisons()` and other functions in *marginaleffects* splits the average by a grouping variable.

```{r bycard, eval = run}
avg_comparisons(fit, variables = "rhc", by = "card")
```

The two subgroup estimates are close, and both intervals reach zero. A common mistake is to stop here and conclude that the effect differs between groups; that comparison is not a test. The question is whether the two effects differ from each other, which needs the difference of the two (i.e., a difference of differences) with an interval of its own. This can be requested by specifying `hypothesis = ~pairwise`.

```{r hypo, eval = run}
avg_comparisons(fit, variables = "rhc", by = "card",
                hypothesis = ~pairwise)
```

The difference is small with an interval covering zero, so there is no evidence here that the effect of catheterization varies by cardiovascular disease status. This is how to test an interaction in a model that never had an interaction term to test, and the interval on that difference is the only thing that separates a real interaction from two subgroup estimates that merely look different.

`avg_predictions()` does the same thing for predictions rather than differences, which is useful for describing groups:

```{r bypred, eval = run}
avg_predictions(fit, by = "card")
```

These correspond to the fitted predictions among those in each cardiovascular disease status group. That answers a different question from whether the model predicts that changing units' disease status changes the probability of death, which corresponds to running the following:

```{r}
avg_predictions(fit, variables = "card")
```

In the latter, `avg_predictions()` computes predictions for all units, first setting `card` to `"no"` and then to `"yes"`.

## The Shape of a Relationship (`partial_dependence()`)

Averages can hide the shape of the relationship between a predictor and the outcome. To see the fitted function we plot predictions against one predictor while the others are averaged over. `partial_dependence()` does this: it sets the predictor to each value of a grid in turn, predicts for every unit, and averages within each posterior draw, so the interval that comes back is on the average prediction rather than on any one patient's. `partial_dependence(fit, ~ x)` builds the grid, and `plot()` draws it.

```{r pdp, eval = run, fig.height = 3.0}
library(ggplot2)

pd <- partial_dependence(fit, ~aps)

pd

plot(pd) +
  labs(x = "APACHE III score on day 1", y = "Fitted probability of death")

# Same thing:
## plot(fit, ~aps)
```

The probability of death rises with the illness score, and the rise is not a straight line on any scale the model was told about; nothing was specified to find the shape. `plot()` returns a *ggplot2* object, so the labels above are added the way they would be to any other plot.

The band is a credible interval and is wide at the top, where few patients were that sick, so the ends of a curve deserve more caution than the middle: with little data out there the forest shrinks its predictions toward the overall mean, which flattens the curve at both edges of the predictor's range.

Naming a second predictor shows how the shape differs across groups.

```{r pdp2, eval = run, fig.height = 3.0}
pd2 <- partial_dependence(fit, ~ aps + rhc)

pd2

plot(pd2) +
  labs(x = "APACHE III score on day 1", y = "Fitted probability of death",
       color = "Catheterized", fill = "Catheterized")
```

The two curves run close together; if they diverged, that would be a moderation worth reporting, and the difference of differences above is how to put a number on it.

### Values of the Second Predictor

A second predictor with a few values gives one curve for each of them, as happened above. With a continuous predictor, it is held at three values near its quartiles and a message reports them.

```{r pdp3, eval = run, message = TRUE, fig.height = 3.0}
plot(fit, ~ meanbp + aps) +
  labs(x = "Mean arterial blood pressure", y = "Fitted probability of death",
       color = "APACHE III", fill = "APACHE III")
```

The `values` argument chooses others, and an entry of it may be a function of the predictor rather than the values themselves. Setting `values = list(z = unique)` asks for every value a predictor takes, which usually suits a numeric predictor with four or five of them; the illness score has 114, so here we write the summary we want instead.

```{r pdp4, eval = run, fig.height = 3.0}
plot(fit, ~ meanbp + aps,
     values = list(aps = function(x) quantile(x, c(.05, .5, .95)))) +
  labs(x = "Mean arterial blood pressure", y = "Fitted probability of death",
       color = "APACHE III", fill = "APACHE III")
```

### The Same Curve Through *marginaleffects*

`marginaleffects::avg_predictions()` produces the same estimates as `partial_dependence()`. It has additional options that may be helpful, but otherwise the syntax is similar.

```{r pdp5, eval = run}
at <- quantile(rhc$aps, c(.1, .5, .9))

# Uses posterior mean
partial_dependence(fit, ~ aps, values = list(aps = at))

# Uses posterior median by default
avg_predictions(fit, variables = list(aps = at))
```

The two summarize one posterior two ways. `partial_dependence()` reports its mean and `avg_predictions()` its median, which is the whole of the difference between the estimates; the intervals are quantiles of the same draws and agree to the digit.

What *marginaleffects* adds is control over what the other predictors do. `partial_dependence()` always averages them over the sample, while `predictions()` and `plot_predictions()` in *marginaleffects* allows one to hold them fixed instead, and prints the profile it chose beside the estimates:

```{r pdp6, eval = run}
plot_predictions(fit, condition = list(aps = at), draw = FALSE)
```

Those columns are why the numbers move: each row is a prediction for one synthetic patient who is average or modal in every other respect, where the curve above is an average over the patients in the data. The average should be reported for a population, and the profile for a described kind of patient. `plot_predictions()` also takes an arbitrary grid and will draw slopes and comparisons across a condition, which is where to go when the question outgrows a partial dependence plot.

## Slopes and the Predictor Transform (`avg_slopes()`)

`marginaleffects::avg_slopes()` reports the derivative of the regression function with respect to a predictor; this is closest quantity to a regression slope. However, whether that derivative means anything depends on `gate` and `x_transform` in `bartisan_control()`. Under a soft gate, the fit is a smooth function of the (transformed) predictor; with `gate = "hard"`, the fit is piecewise constant under any transform and has no derivative worth taking.

Because a numeric predictor is transformed to the interval $[0, 1]$ according to `x_transform`, how its effect translates into a slope depends on the value supplied. Writing the fit as $f(T(x))$, a slope is $f'(T(x))\,T'(x)$, where $T(\cdot)$ is the transform function specified in `x_transform`: an affine $T(\cdot)$ has a known constant derivative, so only $f'$ is estimated.

The default, `"smoothcdf"`, is differentiable, so `avg_slopes()` works, but any error in estimating the density of $X$ yields additional error in the estimation of the slope. Under `"range"`, which is an affine map from the covariate's observed range to $[0, 1]$, slopes are more accurate. `"quantile"` maps each predictor through its empirical distribution function, which is a step function, so the fitted function is a step function of the original predictor and the difference quotient grows without bound as the step shrinks. There is no derivative there to estimate, and the number `avg_slopes()` returns is a property of the step size rather than of the fit.

So, when a slope is the quantity being reported, it is best to fit the model with `x_transform = "range"`, though the default can be okay as well, especially with a large sample (which yields less error in the estimation of the predictor distribution). For everything else, it makes most sense to use `avg_comparisons()` with a step we can interpret, as with `aps = 10` above, which evaluates the fit at two points a substantive distance apart rather than dividing by a vanishing one. This is documented at `?bartisan-marginaleffects`.

## Descriptions of the Fitted Model

Everything here is a description of the fitted model. `avg_comparisons()` reports what the model predicts would differ between two versions of the data, which is a causal quantity only if the model contains enough covariates to account for confounding. Patients were not randomized to catheterization, so for this fit that is a strong assumption. `vignette("causal")` covers what is needed.

The intervals are posterior credible intervals under the model: they cover the uncertainty in the fitted function, and they do not cover the possibility that the model is missing a confounder, that the outcome is measured with bias, or that the sample is not the population of interest.

## Further Reading

`vignette("varying")` covers `vc()`, which writes an effect into the model as a parameter with a prior of its own rather than reading it off as a contrast, and is where to go when the effect is the point of the model. `vignette("importance")` covers which predictors the forest uses, which is a different question from how much they move the outcome. `vignette("diagnostics")` covers whether the fit can be trusted before any of this is read. `?bartisan-marginaleffects` documents which *marginaleffects* functions are supported and the arguments that are specific to this package, and `?estimate_effect` and `?partial_dependence` document the two questions answered without it.

## References
