---
title: "Effects, curves and interactions"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Effects, curves and interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.4, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 4)

has_me <- rlang::is_installed("marginaleffects")
has_gg <- rlang::is_installed("ggplot2")
run <- rlang::is_installed("MASS") && has_me
```

## Introduction

A forest has no coefficients. There is no table of slopes to read, and no
standard error to put beside one. This is the part of the workflow that changes
most when moving from `lm()` to BART, and it is the part where the change is an
improvement rather than a cost.

The replacement is to ask the fitted model questions about predictions. What does
it predict for these people? What would it predict if this variable were
different? How much does the answer differ between groups? The *marginaleffects*
package [@arelbundock2024] asks all of them, and returns posterior intervals with
the answers.

This vignette covers the questions worth asking and how to phrase them.
`vignette("workflow")` is the shorter tour, and this expands its section on
interpreting the fit.

```{r libs, eval = run}
library(bartisan)
library(marginaleffects)

data("birthwt", package = "MASS")

d <- birthwt
d$race  <- factor(d$race,  levels = 1:3, labels = c("white", "black", "other"))
d$smoke <- factor(d$smoke, levels = 0:1, labels = c("no", "yes"))
d$ht    <- factor(d$ht,    levels = 0:1, labels = c("no", "yes"))
d$ui    <- factor(d$ui,    levels = 0:1, labels = c("no", "yes"))
d$low   <- NULL

set.seed(2026)

fit <- bartisan(bwt ~ ., data = d, family = gaussian(), chains = 4)
```

## Three questions

Everything below is one of three things.

A **prediction** is what the model expects for a set of covariate values.
`predictions()` gives one per row, `avg_predictions()` averages them.

A **comparison** is the difference between two predictions that differ in one
variable. This is the closest thing to a regression coefficient and it is what
you usually want.

A **slope** is the derivative of the prediction with respect to a numeric
variable. It is the least useful of the three here, for a reason given below.

## Average effects

`avg_comparisons()` with no other arguments gives every predictor at once.

```{r avgcomp, eval = run}
avg_comparisons(fit)
```

Read this as a coefficient table. For a categorical predictor the estimate is the
average difference in predicted birth weight between the named levels, holding
everything else at each person's own values. For a numeric predictor it is the
average difference associated with an increase of one unit.

Infants of mothers who smoked weigh about 290 grams less. Uterine irritability is
associated with about 445 grams less. Age has an interval covering zero.

The averaging matters and is worth being explicit about. The model's effect of
smoking is not a single number: it can differ for every mother, because the
forest is free to fit an interaction between smoking and anything else. What
`avg_comparisons()` reports is the average of those individual differences over
the sample, which is a well defined quantity whether or not the effect is
constant.

### Choosing the step for a numeric predictor

One unit is the default and is often the wrong scale. A pound of maternal weight
is a small change; ten pounds is a quantity someone might act on.

```{r step, eval = run}
avg_comparisons(fit, variables = list(lwt = 10))
```

Always say which step you used when reporting a numeric effect. Unlike a linear
model, the answer here is not simply ten times the one-unit effect, because the
relationship is not assumed to be a straight line.

## Effects for subgroups

`by` splits the average by a grouping variable.

```{r byui, eval = run}
avg_comparisons(fit, variables = "smoke", by = "ui")
```

The smoking difference is about 300 grams among mothers without uterine
irritability and about 245 grams among those with it. The intervals overlap
heavily.

A common mistake is to stop here and conclude that the effect differs between
groups, because one interval excludes zero and the other does not. That
comparison is not a test. The question is whether the two effects differ from
each other, which needs the difference of the two, with its own interval.

```{r hypo, eval = run}
avg_comparisons(fit, variables = "smoke", by = "ui",
                hypothesis = ~pairwise)
```

The difference is about 27 grams with an interval from roughly -110 to 350. There
is no evidence here that the effect of smoking depends on uterine irritability.
This is how to test an interaction in a model that never had an interaction term
to test.

`avg_predictions()` does the same thing for predictions rather than differences,
which is useful for describing groups:

```{r bypred, eval = run}
avg_predictions(fit, by = "race")
```

## The shape of a relationship

Averages hide shape. To see the fitted function, plot predictions against one
predictor with everything else held fixed.

```{r pdp, eval = run && has_gg}
plot_predictions(fit, condition = "lwt")
```

Predicted birth weight rises with maternal weight to about 125 pounds and is flat
above that. This is the plot a linear model cannot produce, because it would
report a single slope for the whole range. Nothing was specified to find the
bend.

The band is a credible interval and is wide at both ends, where few mothers were
observed. Treat the ends with more caution than the middle.

Adding a second variable shows how the shape differs across groups.

```{r pdp2, eval = run && has_gg}
plot_predictions(fit, condition = c("lwt", "smoke"))
```

The curves are close to parallel, which is the picture that goes with the
non-significant interaction above. If they converged, crossed, or had different
shapes, that would be a moderation worth reporting, and the difference of
differences is how to put a number on it.

## Why slopes are unreliable here

`avg_slopes()` reports a derivative. It is available and it will return a number,
but that number should not be trusted for a fit made with the default settings.

The reason is the predictor transform. By default numeric predictors are mapped
through their empirical distribution function before the trees see them, which
makes the fitted function a step function of the original predictor. Between two
observed values of `lwt` the prediction does not change at all, so the difference
quotient is either exactly zero or a whole step divided by a very small number,
depending on where the step lands. On this fit `avg_slopes()` reports several
hundred grams per pound where a one-unit comparison gives about 2.5.

Use `comparisons()` with a step you can interpret. If a genuine derivative is
needed, refit with `x_transform = "range"` in `bartisan_control()`, which is
linear and does have one. This is documented at `?bartisan-marginaleffects`.

## Other scales

`type` chooses the scale on which predictions are made and therefore the scale on
which effects are reported. For a binomial fit, `type = "response"` gives a
difference in probability and `type = "link"` a difference on the linear
predictor. The first is almost always the one to report, since a difference in
probability is interpretable without reference to the model.

For survival families, `type = "survival"` with a `times` argument gives a
difference in survival probability at a horizon. See `vignette("survival")`.

## What these are not

Everything here is a description of the fitted model. `avg_comparisons()` reports
what the model predicts would differ between two versions of the data, which is a
causal quantity only if the model contains enough covariates to account for
confounding. That is a question about the study design and not about the fit.
`vignette("causal")` covers what is needed.

The intervals are posterior credible intervals under the model. They cover the
uncertainty in the fitted function, and they do not cover the possibility that
the model is missing a confounder, that the outcome is measured with bias, or
that the sample is not the population of interest.

## Where to go next

`vignette("importance")` covers which predictors the forest uses, which is a
different question from how much they move the outcome. `vignette("diagnostics")`
covers whether the fit can be trusted before any of this is read.
`?bartisan-marginaleffects` documents which *marginaleffects* functions are
supported and the arguments that are specific to this package.

## References
