---
title: "Effects, curves and interactions"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Effects, curves and interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{=html}
<style>
pre {overflow-x: auto;}
pre code {word-wrap: normal; white-space: pre;}
</style>
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.2, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 4, marginaleffects_safe = FALSE)

has_me <- rlang::is_installed("marginaleffects")
has_gg <- rlang::is_installed("ggplot2")
run <- has_me
```

## Introduction

A forest has no coefficients: there is no table of slopes to read, and no
standard error to put beside one. This is the part of the workflow that changes
most when moving from `glm()` to BART, and it is the part where the change is an
improvement rather than a cost.

The replacement is to ask the fitted model questions about predictions. What does
it predict for these people? What would it predict if this variable were
different? How much does the answer differ between groups? The *marginaleffects*
package [@arelbundock2024] asks all of them, and returns posterior intervals with
the answers.

This vignette covers the questions worth asking and how to phrase them.
`vignette("bartisan")` is the shorter tour, and this expands its section on
interpreting the fit.

In this guide, we will start from the three kinds of question the package
answers and then work through them in turn. First we'll take average effects
and the choice of step for a numeric predictor, along with the one prior setting
(`sparsity`) that can quietly attenuate a contrast. Next we'll split those
effects by subgroup and test a moderation with the difference of the two, then
plot the shape of a fitted relationship and choose the scale on which an effect
is reported. Finally we'll write the effect as a parameter rather than a
contrast with `vc()`, and close with what none of these estimates can be taken
to mean.

```{r libs, eval = run}
library(bartisan)
library(marginaleffects)

data("rhc")

model <- death ~ rhc + age + sex + race + edu + aps + meanbp + resp +
  hema + pafi + paco2 + crea + surv2m + card

set.seed(2026)

fit <- bartisan(model, data = rhc, family = binomial(), chains = 4)
```

## Three Questions

Everything below is one of three things.

A **prediction** is what the model expects for a set of covariate values.
`predictions()` gives one per row, `avg_predictions()` averages them.

A **comparison** is the difference between two predictions that differ in one
variable. This is the closest thing to a regression coefficient, and it is
usually what we want.

A **slope** is the derivative of the prediction with respect to a numeric
variable. It is the least useful of the three here, for a reason given below.

## Average Effects (`avg_comparisons()`)

`avg_comparisons()` with no `variables` argument gives every predictor at once,
which for this model is a long table. A few at a time is easier to read:

```{r avgcomp, eval = run}
avg_comparisons(fit, variables = c("rhc", "age", "card"))
```

This reads as a coefficient table. Each estimate is an average difference in
predicted probability, holding everything else at each patient's own values: for
a factor, between the levels named in the `Contrast` column, and for a numeric
predictor, for an increase of one unit. `rhc` is coded `0` and `1`, so its
one-unit contrast is the treatment effect.

### The Splitting Prior and a Contrast (`sparsity`)

An estimate here can come back as exactly zero, and an interval bound with it;
that is not a rounding artifact. The default splitting prior is a
variable-selection prior (i.e., one that can leave a predictor out of the forest
altogether), so in a draw where it uses the predictor in no tree the prediction
does not depend on it and the contrast is exactly zero; the posterior of the
contrast is a mixture with a point mass there, holding whatever share of draws
dropped the predictor.

It matters more than it sounds: on a simulated effect of .2 against residual
noise of 1, the prior halved the estimate and its 95% interval covered the truth
only 60% of the time. A strong effect is untouched, because the prior never has reason
to drop a predictor that is earning its splits, so this is a weak-signal problem
rather than a general one.

If a contrast is what we are reporting, we fit with `sparsity = FALSE`, or with
`split_prior`, which fixes the weights (i.e., the probability that each
predictor is chosen for a split) and so cannot drop anything.
`?bartisan_control` has the measurements for both.

### The Step for a Numeric Predictor (`variables`)

One unit is the default and is often the wrong scale. One point of an illness
score is a small change; ten points is a difference someone would notice.

```{r step, eval = run}
avg_comparisons(fit, variables = list(aps = 10))
```

A numeric effect should always be reported together with the step it was
computed at. Unlike a linear model, the answer here is not ten times the
one-unit effect, because the relationship is not assumed to be a straight line.

## Effects for Subgroups (`by`)

`by` splits the average by a grouping variable.

```{r bycard, eval = run}
avg_comparisons(fit, variables = "rhc", by = "card")
```

The two subgroup estimates are close, and both intervals reach zero.

A common mistake is to stop here and conclude that the effect differs between
groups; that comparison is not a test. The question is whether the two effects
differ from each other, which needs the difference of the two (i.e., a
difference of differences) with an interval of its own.

```{r hypo, eval = run}
avg_comparisons(fit, variables = "rhc", by = "card",
                hypothesis = ~pairwise)
```

The difference is small with an interval covering zero, so there is no evidence
here that the effect of catheterization depends on cardiovascular disease. This
is how to test an interaction in a model that never had an interaction term to
test, and the interval on that difference is the only thing that separates a
real interaction from two subgroup estimates that merely look different.

`avg_predictions()` does the same thing for predictions rather than differences,
which is useful for describing groups:

```{r bypred, eval = run}
avg_predictions(fit, by = "card")
```

## The Shape of a Relationship (`plot_predictions()`)

Averages hide shape. The fitted function is seen by plotting predictions against
one predictor with everything else held fixed; below we ask for the numbers
rather than the plot, which gives more control over how it is drawn:

```{r pdp, eval = run && has_gg, fig.height = 3.0}
library(ggplot2)

curve <- plot_predictions(fit, condition = "aps", draw = FALSE)

ggplot(curve, aes(aps, estimate)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  geom_line() +
  labs(x = "APACHE III score on day 1", y = "fitted probability of death") +
  theme_bw(base_size = 9)
```

The probability of death rises with the illness score, and the rise is not a
straight line on any scale the model was told about; nothing was specified to
find the shape.

The band is a credible interval and is wide at the top, where few patients were
that sick, so the ends of a curve deserve more caution than the middle: with
little data out there the forest shrinks its predictions toward the overall
mean, which flattens the curve at both edges of the predictor's range.

Adding a second variable shows how the shape differs across groups.

```{r pdp2, eval = run && has_gg, fig.height = 3.0}
curve2 <- plot_predictions(fit, condition = c("aps", "rhc"), draw = FALSE)

ggplot(curve2, aes(aps, estimate, colour = factor(rhc))) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = factor(rhc)),
              alpha = 0.15, colour = NA) +
  geom_line() +
  labs(x = "APACHE III score on day 1", y = "fitted probability of death",
       colour = "catheterized", fill = "catheterized") +
  theme_bw(base_size = 9)
```

The two curves run close together; if they diverged, that would be a moderation
worth reporting, and the difference of differences above is how to put a number
on it.

## Scales (`type`)

`type` chooses the scale on which predictions are made and therefore the scale on
which effects are reported.

```{r link, eval = run}
avg_comparisons(fit, variables = "rhc", type = "link")
```

On the link scale this is a difference in log-odds, which is what a logistic
regression coefficient is. It is the less useful of the two here: a difference in
probability is interpretable without reference to the model and is the number a
reader can act on, whereas a difference in log-odds needs a baseline before it
means anything.

For survival families, `type = "survival"` with a `times` argument gives a
difference in survival probability at a horizon. See `vignette("survival")`.

## Unreliable Slopes (`avg_slopes()`)

`avg_slopes()` reports a derivative; it is available and it will return a
number, but that number should not be trusted for a fit made with the default
settings.

The reason is the predictor transform: by default numeric predictors are mapped
through their empirical distribution function before the trees see them, which
makes the fitted function a step function of the original predictor. Between two
observed values the prediction does not change at all, so the difference quotient
is either exactly zero or a whole step divided by a very small number, depending
on where the step lands.

Instead we use `comparisons()` with a step we can interpret, as with `aps = 10`
above. If a genuine derivative is needed, we refit with `x_transform = "range"`
in `bartisan_control()`, which is linear and so does have one. This is
documented at `?bartisan-marginaleffects`.

## Varying Coefficients (`vc()`)

Everything above reads an effect out of a fitted surface by asking the model what
it predicts under two versions of the data. There is another way to write the
model, in which the effect is a parameter rather than a contrast:

$$f_0(x) + z\,f_1(x)$$

Here $f_1$ is a forest of its own, and it *is* the effect of `z`: how much the
prediction moves per unit of `z`, as a function of the other predictors. `vc()`
asks for it. This is the varying-coefficient model of @deshpande2026, of which
@hahn2020 is the case of one binary covariate and @woody2020 the case of one
continuous one.

```{r vc, eval = run}
fit_vc <- bartisan(death ~ age + sex + race + edu + aps + meanbp + resp +
                     hema + pafi + paco2 + crea + surv2m + card + vc(rhc),
                   data = rhc, family = binomial(), chains = 4,
                   sparsity = FALSE)

head(coef(fit_vc))
```

`coef()` returns one value per patient, which is what a coefficient becomes when
it is allowed to vary. It is on the link scale, so for this binary outcome it is
a difference in log odds rather than in probability; `avg_comparisons()` is still
what reports an effect on the scale a reader can act on.

What the reparameterization buys is a prior on the effect itself. The forest for
$f_1$ is regularized separately from the forest for the rest of the outcome, so
shrinking the prognostic part does not shrink the effect. `vignette("causal")`
covers why that matters and `bcf()` sets it up for the causal case.

Two things worth knowing before reaching for it.

The effect is **linear in the covariate** unless we say otherwise. For a binary
treatment that is no assumption at all, since there are only two values. For a
continuous predictor it says the effect is proportional to it, which is a real
restriction. Letting the coefficient's forest split on the covariate itself
removes it, and then the effect varies across the covariate's own range:

```{r vcnonlinear, eval = FALSE}
# The effect of `aps` may itself change across `aps`.
y ~ age + vc(aps, ~ aps + age)
```

And a covariate whose coefficient varies should not also be a predictor of the
control function. With it in both, the two are not separately identified: any
function of it can move between them. Writing the covariate only inside `vc()`
is what keeps them apart, and `bartisan()` warns if the formula does otherwise.

For a family with several additive predictors, each parameter's formula carries
its own `vc()` terms, so a covariate can have a coefficient on more than one of
them:

```{r vcmulti, eval = FALSE}
# The effect of `z` on the mean, and separately on the spread.
bartisan(list(mean = y ~ x1 + x2 + vc(z), log_sd = ~ x1 + x2 + vc(z)),
         data = d, family = gaussian_ls())
```

`coef()` then returns one column per coefficient, named `mean:z` and `log_sd:z`
for the forests they come from, which is also how per-forest settings like
`num_trees` are keyed. `?vc` covers the rest, including the one family that
refuses this.

## What These Estimates Are Not

Everything here is a description of the fitted model. `avg_comparisons()` reports
what the model predicts would differ between two versions of the data, which is a
causal quantity only if the model contains enough covariates to account for
confounding. Patients were not randomized to catheterization, so for this fit
that is a strong assumption. `vignette("causal")` covers what is needed.

The intervals are posterior credible intervals under the model: they cover the
uncertainty in the fitted function, and they do not cover the possibility that
the model is missing a confounder, that the outcome is measured with bias, or
that the sample is not the population of interest.

## Where to Go Next

`vignette("importance")` covers which predictors the forest uses, which is a
different question from how much they move the outcome. `vignette("diagnostics")`
covers whether the fit can be trusted before any of this is read.
`?bartisan-marginaleffects` documents which *marginaleffects* functions are
supported and the arguments that are specific to this package.

## References
