---
title: "Generalized BART with genbart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalized BART with genbart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.6, dpi = 110)
options(digits = 3)
```

```{r libs}
library(genbart)
library(ggplot2)

set.seed(2026)
```

Every fit below uses a deliberately small chain so that this vignette builds
quickly: 20 trees and 300 draws after 300 warmup iterations, on 400
observations. The defaults, which are what you would actually use, are 50 trees
and 1000 draws after 1000 warmup iterations. Expect the numbers here to be
noisier than a real fit would give.

```{r control}
ctrl <- genbart_control(num_trees = 20, num_burn = 300, num_save = 300)

n <- 400
```

## What this package is for

Bayesian additive regression trees are usually tied to a Gaussian likelihood,
because the Gibbs sampler needs to integrate the leaf parameters out of the tree
likelihood in closed form. Reaching anything else normally means finding a
data-augmentation scheme that turns the problem back into a Gaussian one, and
for many likelihoods no such scheme exists.

*genbart* implements the sampler of Linero (2025), which removes that
restriction by proposing leaf values from a Gaussian approximation to their
conditional posterior and correcting with a Metropolis step. The approximation
only has to be good enough to be accepted often; the chain targets the exact
posterior either way. The practical consequence is that a family needs only the
log density of one observation and its first two derivatives, which is why the
list of available families includes models that are hard to find in a BART
package at all.

The interface is deliberately the one you already use for `glm()`.

## A first model

We use the Friedman function throughout, since it mixes an interaction, a
quadratic term, two linear terms, and three predictors that do nothing.

```{r sim}
friedman <- function(x) {
  10 * sin(pi * x[, 1] * x[, 2]) + 20 * (x[, 3] - 0.5)^2 +
    10 * x[, 4] + 5 * x[, 5]
}

make_data <- function(n, p = 8) {
  x <- matrix(runif(n * p), nrow = n)
  colnames(x) <- paste0("x", seq_len(p))
  list(x = x, eta = (friedman(x) - 14) / 5, d = as.data.frame(x))
}

train <- make_data(n)
test <- make_data(n)
```

A binary outcome, fit exactly as it would be with `glm()`:

```{r logit}
d <- train$d
d$y <- rbinom(n, 1, plogis(train$eta))

fit <- genbart(y ~ ., data = d, family = binomial("logit"), control = ctrl)

fit
```

Predictions come with full posterior uncertainty, because every draw of every
tree is retained:

```{r logit-pred}
p_hat <- predict(fit, newdata = test$d, type = "response")
head(p_hat)

draws <- predict(fit, newdata = test$d, type = "response", draws = TRUE)
dim(draws)
```

The fit recovers the truth on the logit scale, and the credible bands widen
where the data are thin:

```{r logit-plot}
eta_draws <- predict(fit, newdata = test$d, type = "link", draws = TRUE)

plot_df <- data.frame(
  truth = test$eta,
  estimate = colMeans(eta_draws),
  lower = apply(eta_draws, 2, quantile, 0.025),
  upper = apply(eta_draws, 2, quantile, 0.975)
)

ggplot(plot_df, aes(truth, estimate)) +
  geom_linerange(aes(ymin = lower, ymax = upper), colour = "grey70",
                 linewidth = 0.3) +
  geom_point(size = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  labs(x = "true linear predictor", y = "posterior mean") +
  theme_bw()
```

## Changing family is changing one word

Everything below is the same call with a different `family`. Ordinary
`stats::family` objects work unchanged; the rest are supplied by this package
in the same style.

```{r families}
run <- function(formula, data, family, truth, label) {
  fit <- genbart(formula, data = data, family = family, control = ctrl)
  eta <- predict(fit, type = "link")
  if (is.matrix(eta)) eta <- eta[, 1]
  data.frame(family = label, correlation = cor(eta, truth))
}

d_pois <- train$d
d_pois$y <- rpois(n, exp(train$eta + 1))

d_gam <- train$d
d_gam$y <- rgamma(n, shape = 4, rate = 4 / exp(train$eta + 1))

d_nb <- train$d
d_nb$y <- rnbinom(n, mu = exp(train$eta + 1), size = 2)

results <- rbind(
  run(y ~ ., d_pois, poisson(), train$eta + 1, "poisson"),
  run(y ~ ., d_nb, negbin(), train$eta + 1, "negbin"),
  run(y ~ ., d_gam, Gamma("log"), train$eta + 1, "Gamma")
)

results
```

Nuisance parameters are drawn alongside the trees and reported in `aux`:

```{r aux}
fit_nb <- genbart(y ~ ., data = d_nb, family = negbin(), control = ctrl)

summary(fit_nb)$aux
```

## Counts with too many zeros

`zi_poisson()` and `zi_negbin()` give the count mean and the structural-zero
probability separate forests, so the excess-zero mechanism can depend on the
predictors rather than being a single constant.

```{r zi}
log_mu <- 1 + 1.5 * (train$x[, 1] - 0.5)
p_zero <- plogis(-0.5 + 2 * (train$x[, 2] - 0.5))

d_zi <- train$d
d_zi$y <- ifelse(rbinom(n, 1, p_zero) == 1, 0, rpois(n, exp(log_mu)))

mean(d_zi$y == 0)

fit_zi <- genbart(y ~ ., data = d_zi, family = zi_poisson(), control = ctrl)

link <- predict(fit_zi, type = "link")
c(count = cor(link[, "count"], log_mu),
  zero = cor(link[, "zero"], qlogis(p_zero)))
```

Each component finds its own driver, which is visible in the splitting counts:

```{r zi-usage}
usage <- summary(fit_zi)$usage

head(usage$count, 3)
head(usage$zero, 3)
```

## Ordinal outcomes

`ordinal()` uses the cumulative-link parameterization of `MASS::polr()`, with
the first cutpoint pinned at zero for identifiability, so only the spacing
between cutpoints is estimated.

```{r ordinal}
cuts <- c(-1, 0.5, 2)
probs <- vapply(train$eta, function(e) diff(c(0, plogis(cuts - e), 1)),
                numeric(4))

d_ord <- train$d
d_ord$y <- factor(apply(probs, 2, function(p) sample.int(4, 1, prob = p)),
                  levels = 1:4, ordered = TRUE)

fit_ord <- genbart(y ~ ., data = d_ord, family = ordinal(), control = ctrl)

# Only the gaps are identified, so compare those.
rbind(estimated = colMeans(fit_ord$aux)[-1],
      truth = cuts[-1] - cuts[1])
```

Category probabilities and the most probable class both come from `predict()`:

```{r ordinal-pred}
head(predict(fit_ord, type = "prob"), 3)

table(predicted = predict(fit_ord, type = "class"), observed = d_ord$y)
```

## Unordered categories

`multinomial()` fits one forest per category and leaves the model unidentified:
adding any function of the predictors to every category's forest leaves the
probabilities alone. This is the parameterization of Murray (2021). The proper
leaf prior makes the posterior proper regardless, and everything identified --
a probability, an odds ratio -- comes out of the draws as usual. Its point is
that the prior is symmetric in the categories, whereas singling out a reference
makes the fit depend on which one you singled out.

```{r multinomial}
lin <- cbind(2 * sin(pi * train$d$x1), 3 * (train$d$x2 - 0.5)^2 - 0.5,
             -1.5 * train$d$x3)
lin <- lin - rowMeans(lin)
P <- exp(lin) / rowSums(exp(lin))

d_mn <- train$d
d_mn$y <- factor(c("a", "b", "c")[apply(P, 1, function(p) sample.int(3, 1, prob = p))],
                 levels = c("a", "b", "c"))

fit_mn <- genbart(y ~ ., data = d_mn, family = multinomial(), control = ctrl)

# One forest per category, and the fitted probabilities.
names(fit_mn$eta)

round(head(predict(fit_mn, type = "prob"), 3), 3)
```

Pass `reference` to get the other parameterization, in which the fitted
functions are log odds against that category and are read the way
`nnet::multinom()` coefficients are:

```{r multinomial-ref}
fit_ref <- genbart(y ~ ., data = d_mn, family = multinomial(reference = "a"),
                   control = ctrl)

names(fit_ref$eta)
```

## Links the package does not compile

The links listed in `?genbart-families` are the ones the sampler evaluates in
compiled code. Any other link works for the five families with a single mean and
a conventional link, applied from R: the additive predictor is mapped to the
scale the compiled family works on, and the chain rule carries the derivatives
back. Nothing about the call changes.

```{r cauchit}
# Heavy tails: a Cauchy link keeps a small probability of the wrong outcome even
# far out, which a logit link cannot represent.
d_bin <- train$d
d_bin$y <- rbinom(n, 1, pcauchy(4 * train$eta))

test_bin <- test$d
test_bin$y <- rbinom(n, 1, pcauchy(4 * test$eta))

fit_cauchit <- genbart(y ~ ., data = d_bin, family = binomial("cauchit"),
                       control = ctrl)

fit_cauchit
```

Held out, the right link earns its keep. `type = "density"` evaluates the
conditional density of the outcome at the value the outcome actually took, so
summing its log is a proper score:

```{r cauchit-score}
log_score <- function(fit) {
  sum(log(predict(fit, newdata = test_bin, type = "density")))
}

fit_logit <- genbart(y ~ ., data = d_bin, family = binomial("logit"),
                     control = ctrl)

c(cauchit = log_score(fit_cauchit), logit = log_score(fit_logit))
```

A link you write yourself is supplied the way `glm()` takes one, as a
`link-glm` object. Only `linkinv` and `linkfun` are needed; `mu.eta` saves the
sampler from differencing the composition.

```{r own-link}
half_probit <- structure(
  list(linkfun = function(mu) qnorm(mu) / 2,
       linkinv = function(eta) pnorm(2 * eta),
       mu.eta  = function(eta) 2 * dnorm(2 * eta),
       valideta = function(eta) TRUE, name = "half-probit"),
  class = "link-glm")

fit_half <- genbart(y ~ ., data = d_bin, family = binomial(half_probit),
                    control = ctrl)

fit_half$family$link
```

Three caveats. This route costs a call into R for every leaf the sampler visits,
so it is slower than a compiled link. The leaf prior scale is calibrated for the
compiled link, so an unusual one may want `sigma_mu` set by hand. And the
additive predictor is unconstrained, so a link whose inverse has a restricted
range -- `poisson("identity")`, `Gamma("inverse")` -- will produce non-finite
densities for some predictors; those proposals are rejected rather than breaking
the chain, but the fit will be poor.

## Supplying a likelihood

The sampler needs one thing from a family: the log density of an observation and
its first two derivatives with respect to each additive predictor. Because that
is the whole interface, it can be handed over from R. `custom_family()` takes
the log density and differences it for the derivatives.

Here is the Poisson family written out by hand, checked against the compiled
one. Terms free of the predictor may be dropped -- they cancel from every
acceptance ratio -- which is why `lgamma(y + 1)` is absent.

```{r custom-poisson}
by_hand <- custom_family(function(y, eta) y * eta[, 1] - exp(eta[, 1]),
                         start = log(mean(d_pois$y)))

fit_hand <- genbart(y ~ ., data = d_pois, family = by_hand, control = ctrl)
fit_builtin <- genbart(y ~ ., data = d_pois, family = poisson(), control = ctrl)

c(supplied = cor(predict(fit_hand, type = "link"), train$eta + 1),
  compiled = cor(predict(fit_builtin, type = "link"), train$eta + 1))
```

More than one predictor is a matter of saying so. This is a location-scale
model: the mean and the log standard deviation each get their own forest.

```{r custom-two}
d_ls <- train$d
log_sd <- -1 + d_ls$x3
d_ls$y <- train$eta + rnorm(n, sd = exp(log_sd))

fit_ls <- genbart(
  y ~ .,
  family = custom_family(
    function(y, eta) dnorm(y, eta[, 1], exp(eta[, 2]), log = TRUE),
    num_predictors = 2, start = c(0, -1), name = "location-scale by hand"),
  data = d_ls, control = ctrl)

eta_ls <- predict(fit_ls, type = "link")

c(mean = cor(eta_ls[, 1], train$eta), log_sd = cor(eta_ls[, 2], log_sd))
```

The function is called once per leaf per Fisher-scoring step with the
observations reaching that leaf, so it must be vectorized over `y` and the rows
of `eta`, and it must return exactly one value per row. Supplying `derivatives`
turns three calls into one and removes the differencing error:

```{r custom-derivs}
sharper <- custom_family(
  function(y, eta) y * eta[, 1] - exp(eta[, 1]),
  derivatives = function(y, eta, h) {
    list(score = y - exp(eta[, h]), info = exp(eta[, h]))
  },
  start = log(mean(d_pois$y)))

cor(predict(genbart(y ~ ., data = d_pois, family = sharper, control = ctrl),
            type = "link"),
    train$eta + 1)
```

What this route cannot do: it cannot draw a nuisance parameter, so a dispersion
has to be fixed inside the closure; the response must be numeric; and since the
package cannot know what the mean of your density is,
`predict(type = "response")` returns the additive predictors rather than a
fitted mean.

## Right-censored survival times

Survival families take a `Surv()` object on the left-hand side. They model
log time as the predictor plus a scaled error, so the error distribution
selects the survival family.

```{r aft}
log_t <- train$eta + 0.8 * rlogis(n)
cutoff <- quantile(log_t, 0.7)

d_aft <- train$d
d_aft$time <- exp(pmin(log_t, cutoff))
d_aft$status <- as.numeric(log_t <= cutoff)

mean(d_aft$status == 0)  # censored fraction

fit_aft <- genbart(survival::Surv(time, status) ~ ., data = d_aft,
                   family = loglogistic_aft(), control = ctrl)

c(correlation = cor(predict(fit_aft, type = "link"), train$eta),
  sigma = mean(fit_aft$aux[, "sigma"]))
```

On the response scale a survival model reports the median survival time:

```{r aft-pred}
head(predict(fit_aft, type = "response"))
```

## Bounded outcomes with mass at the endpoints

`ordbeta()` is the ordered beta regression of Kubinec (2023), for a proportion
or slider response that piles up at zero and one. One predictor drives both the
endpoint probabilities and the beta mean in between.

```{r ordbeta}
eta_ob <- 2 * (train$x[, 1] - 0.5) + 1.5 * (train$x[, 2] - 0.5)
mu_ob <- plogis(eta_ob)

u <- runif(n)
p_one <- plogis(eta_ob - 1.5)
p_zero <- 1 - plogis(eta_ob + 1.5)

y_ob <- rbeta(n, mu_ob * 8, (1 - mu_ob) * 8)
y_ob[u <= p_zero] <- 0
y_ob[u >= 1 - p_one] <- 1

d_ob <- train$d
d_ob$y <- y_ob

c(at_zero = mean(y_ob == 0), at_one = mean(y_ob == 1))

fit_ob <- genbart(y ~ ., data = d_ob, family = ordbeta(), control = ctrl)

summary(fit_ob)$aux
```

## Soft rules versus hard rules

Soft rules are the default. A decision rule becomes a logistic gate rather than
a step, so the fitted function is smooth. That helps when the truth is smooth,
which it usually is, and costs roughly four times as much per iteration. Both
arms below get the same reduced budget, so the timings are comparable with each
other rather than representative of a full run.

```{r soft-hard}
d_gauss <- train$d
d_gauss$y <- train$eta + rnorm(n)

rmse <- function(a, b) sqrt(mean((a - b)^2))

timed <- function(soft) {
  started <- proc.time()[["elapsed"]]
  fit <- genbart(y ~ ., data = d_gauss,
                 control = genbart_control(num_trees = 20, num_burn = 300,
                                           num_save = 300, soft = soft))
  data.frame(
    rules = if (soft) "soft" else "hard",
    test_rmse = rmse(predict(fit, newdata = test$d), test$eta),
    seconds = round(proc.time()[["elapsed"]] - started, 1)
  )
}

rbind(timed(TRUE), timed(FALSE))

truth_sd <- sd(test$eta)
```

For reference, the standard deviation of the true function is
`r round(truth_sd, 2)`, so both are fitting real structure; soft rules are
the more accurate of the two here.

## Missing predictor values

The default `na.action` drops any row with a missing value, which is what most
model-fitting functions in R do and is often a lot of data. Passing
`na.action = na.pass` keeps those rows and lets the splitting rules deal with
the missing values instead. A rule on a variable that has any is drawn as one of
three, with equal probability: the values below the cutpoint go left and missing
goes with them; they go left and missing goes right; or the rule ignores the
cutpoint and splits on whether the value is there at all.

This is *missingness incorporated in attributes*, and the third rule is the
interesting one, because it is what lets the model use a variable whose
**absence** is the signal.

```{r missing}
d_mis <- train$d
d_mis$y <- train$eta + rnorm(n, sd = 0.4)

gone <- runif(n) < 0.3
d_mis$x1[gone] <- NA

nrow(na.omit(d_mis))              # what the default would fit to

fit_mis <- genbart(y ~ ., data = d_mis, na.action = na.pass, control = ctrl)

c(rows_used = fit_mis$n, correlation = cor(predict(fit_mis), train$eta))
```

Here is the case imputation cannot reach. The response depends on whether `x1`
was recorded and on nothing else about it, so filling the gaps in with a
plausible number destroys exactly the information that matters:

```{r missing-informative}
d_inf <- train$d
d_inf$y <- 2 * gone + rnorm(n, sd = 0.3)
d_inf$x1[gone] <- NA

fit_inf <- genbart(y ~ ., data = d_inf, na.action = na.pass, control = ctrl)
p_inf <- predict(fit_inf)

c(where_missing = mean(p_inf[gone]), where_present = mean(p_inf[!gone]))
```

The truth is 2 and 0. The splitting counts show where the model spent its
attention:

```{r missing-counts}
round(head(sort(colMeans(fit_inf$counts$eta), decreasing = TRUE), 4), 2)
```

Three things to know. Because the choice is drawn from its prior along with the
variable and the cutpoint, it cancels from every acceptance ratio, and a
variable with no missing values is not given the extra draw at all -- complete
data reproduces the sampler exactly as it was. A missing value takes a hard path
through the tree even when the rules are soft, which is right: there is nothing
about being absent to smooth over. And `predict()` accepts missing values only
in columns that had them at fitting time, because nowhere else does any rule
carry an answer.

The last point is about what is being estimated. This targets the mean of the
response given the predictors **and the pattern of missingness**, which is what
you want for prediction. If the estimand is a regression or causal effect
defined on complete data, multiple imputation is the right tool and this is not.

## Variable selection

Splitting variables are drawn from a sparsity-inducing Dirichlet prior, so the
model filters out predictors it does not need. `summary()` reports how often
each one was split on.

```{r selection}
fit_gauss <- genbart(y ~ ., data = d_gauss, control = ctrl)

round(summary(fit_gauss)$usage$eta, 2)
```

The five predictors that matter separate cleanly from `x6` to `x8`, which enter
the Friedman function not at all. Factors are expanded to an indicator per level
but share a single weight in this prior, so a factor is selected as a whole
rather than one level at a time.

## The conditional density of the outcome

`type = "density"` evaluates the likelihood contribution of each observation at
its own outcome value, which is why the outcome has to be present in `newdata`.
It is a density for a continuous response, a probability for a discrete one, and
a survival probability for a censored time.

```{r density}
held_out <- test$d
held_out$y <- rpois(n, exp(test$eta + 1))

fit_pois <- genbart(y ~ ., data = d_pois, family = poisson(), control = ctrl)

head(predict(fit_pois, newdata = held_out, type = "density"))
```

Summed on the log scale this is a held-out log score, which is a fair way to
compare models. Here a Poisson fit is compared with a negative binomial fit on
data that really are Poisson:

```{r logscore}
log_score <- function(fit) {
  sum(predict(fit, newdata = held_out, type = "density", log = TRUE))
}

fit_pois_nb <- genbart(y ~ ., data = d_pois, family = negbin(), control = ctrl)

c(poisson = log_score(fit_pois), negbin = log_score(fit_pois_nb))
```

The two are close, as they should be: the negative binomial nests the Poisson
and pays only for estimating a dispersion the data do not need.

## Speed, and several chains at once

A non-conjugate sampler costs more than a conjugate one, and it is worth knowing
how much and why. Fitting a leaf means building a Laplace approximation to its
conditional posterior and then evaluating a Metropolis ratio, where a conjugate
sampler draws directly; that is about three passes over the data where one would
do. Soft rules cost a further factor of three, since a leaf touches every
observation rather than only those in its cell.

Where a data augmentation exists that makes the conditional Gaussian, all of
that collapses: the Laplace approximation stops being an approximation, so
Fisher scoring lands in one step and the leaf refresh becomes a Gibbs step.
`augment = TRUE` does this for a Bernoulli response with a probit link, following
Albert and Chib (1993).

```{r augment}
d_pr <- train$d
d_pr$y <- rbinom(n, 1, pnorm(train$eta))

fast <- genbart_control(num_trees = 20, num_burn = 300, num_save = 300,
                        augment = TRUE)

c(direct = system.time(genbart(y ~ ., d_pr, family = binomial("probit"),
                               control = ctrl))[["elapsed"]],
  augmented = system.time(genbart(y ~ ., d_pr, family = binomial("probit"),
                                  control = fast))[["elapsed"]])
```

Judge it on effective samples per second rather than on the clock: the augmented
chain mixes about two thirds as well, so the net gain is smaller than the timing
suggests. On a 614-row problem it came out at 5.7 times the speed and 3.8 times
the effective samples per second.

Independent chains run in parallel under whatever backend you have planned, and
give a convergence diagnostic the package otherwise has no way to produce. The
chain is the only parallel axis a sampler like this has, since each sweep
conditions on the one before it.

```{r chains}
future::plan(future::multicore, workers = 2)

fit_chains <- genbart(y ~ ., d_pr, family = binomial("probit"), chains = 4,
                      control = fast)

fit_chains$rhat
```

Values much above 1.01 mean the chains have not agreed and the run is too short,
which is exactly what they say here: this vignette runs 300 draws per chain to
build quickly, and this problem needs several thousand. That is the point of
having the diagnostic.

The draws are pooled, so `predict()` and `summary()` work on them unchanged. One
`set.seed()` before the call reproduces the whole run whatever the backend,
because each chain is given its own random number stream.

```{r chains-reset, include = FALSE}
future::plan(future::sequential)
```

## Practical notes

A few things worth knowing before relying on the output.

**Reproducibility.** The sampler draws from R's own generator, so
`set.seed()` is all that is needed.

**Weights** behave as they do in `glm()`: they multiply each observation's
log-likelihood contribution, so integer weights are exactly equivalent to
replicating rows. That makes them correct for frequency weights and for the
number of trials in a binomial model. Treat them with care as survey or
balancing weights, since credible intervals will then reflect the sum of the
weights rather than the actual amount of information.

**Interval coverage.** In simulation at the default settings, nominal 95%
intervals for the additive predictor cover the truth 0.95 of the time for a
Gaussian response, 0.96 for Poisson and gamma, and 0.91 for a binary response,
which carries the least information per observation. What coverage is lost is
lost to shrinkage bias rather than to poor mixing: the posterior mean sits about
0.8 posterior standard deviations from the truth, and running more or longer
chains does not move coverage. Treat pointwise intervals for the regression
function as good but not exact.

**Weakly identified nuisance parameters** — a negative binomial dispersion, a
survival shape — mix more slowly than the regression function, and the data
often do not pin them down. Look at their draws before interpreting them.

**Separable responses.** The leaf scale is drawn under a half-Cauchy prior,
which has no upper bound. When the predictors nearly separate a binary
response the likelihood rewards an unbounded predictor and that prior is not
enough to hold the scale down: in a fully separated example the drawn scale
wandered between 3 and 9 times its prior median over 1600 draws without
settling, and the additive predictor reached 110. `genbart()` warns when the
scale settles more than five times above its prior median. The remedy is
`update_sigma_mu = FALSE` in `genbart_control()`, which pins the scale and
brought the same example back to a maximum predictor of 7.

**The number of trees matters less than it looks.** The default leaf scale is
proportional to \(1/\sqrt{m}\), so the ensemble's prior standard deviation at
any point is the same whatever \(m\) is, and the drawn scale keeps it there:
across 5 to 200 trees on a 400-observation binary problem the posterior mean of
\(\sqrt{m}\,\sigma_\mu\) stayed between 1.9 and 2.1 while \(\sigma_\mu\)
itself fell from 0.87 to 0.13. Held-out log score varied by 1.0 out of 227
across that range, which is inside the replicate-to-replicate noise, while cost
rose from 1.5 to 54 seconds. More trees buy a richer approximation, not a
different model; there is little to gain from tuning \(m\) and much to lose in
time.

**The leaf scale is raised from near zero** over the first quarter of warmup.
Linero (2025) describes this as essential: started at its full value the sampler
can settle early into a poor configuration and fail to move. It is controlled by
`sigma_mu_ramp` in `genbart_control()`.

## References

Kubinec, R. (2023). Ordered beta regression: a parsimonious, well-fitting model
for continuous data with lower and upper bounds. *Political Analysis*, 31(4),
519--536.

Kapelner, A., & Bleich, J. (2015). Prediction with missing data via Bayesian
additive regression trees. *Canadian Journal of Statistics*, 43(2), 224--239.

Murray, J. S. (2021). Log-linear Bayesian additive regression trees for
multinomial logistic and count regression models. *Journal of the American
Statistical Association*, 116(534), 756--769.

Twala, B. E. T. H., Jones, M. C., & Hand, D. J. (2008). Good methods for coping
with missing data in decision trees. *Pattern Recognition Letters*, 29(7),
950--956.

Albert, J. H., & Chib, S. (1993). Bayesian analysis of binary and polychotomous
response data. *Journal of the American Statistical Association*, 88(422),
669--679.

Linero, A. R. (2025). Generalized Bayesian additive regression trees models:
beyond conditional conjugacy. *Journal of the American Statistical
Association*, 120(549), 356--369.

Linero, A. R., & Yang, Y. (2018). Bayesian regression tree ensembles that adapt
to smoothness and sparsity. *Journal of the Royal Statistical Society Series
B*, 80(5), 1087--1110.
