---
title: "Generalized BART with bartisan"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Generalized BART with bartisan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.6, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 3)

has_gg <- rlang::is_installed("ggplot2")
```

```{r libs}
library(bartisan)

set.seed(2026)
```

This vignette describes what the model is and how it is fitted. It is the
reference document rather than the entry point: `vignette("bartisan")` shows how
to use the package on a real analysis, and `vignette("families")` covers choosing
a likelihood. Read this one when you want to know what the fit is doing, why the
defaults are what they are, or what an advanced setting changes.

Fits below use a deliberately small chain so the vignette builds quickly: 20
trees and 300 draws after 300 warmup, on 400 observations. The defaults are 50
trees and 500 draws after 500 warmup.

```{r control}
ctrl <- bartisan_control(num_trees = 20, num_burn = 300, num_draws = 300)

n <- 400
```

# The model

## A sum of trees

A binary regression tree is a pair $(T, M)$, where $T$ is the tree structure,
meaning the splitting rules at the interior nodes, and
$M = \{\mu_1, \dots, \mu_B\}$ collects one parameter for each of its $B$ leaves.
The tree defines a function $g(x; T, M)$ that sends $x$ down the tree according
to the rules and returns the parameter in the leaf it reaches. A single tree is
a step function.

BART sums many of them [@chipman2010]:

$$f(x) = \sum_{m=1}^{M} g(x; T_m, M_m).$$

Each tree is kept small by its prior, so no single one explains much. The sum is
flexible while each term is a weak learner, which is what makes the fit stable.
Interactions come free: a path through a tree that splits on $x_1$ and then on
$x_2$ is an interaction between them, and nothing had to be specified for it to
appear.

In ordinary BART the sum is the conditional mean of a Gaussian outcome,

$$Y = f(x) + \varepsilon, \qquad \varepsilon \sim \mathrm{N}(0, \sigma^2).$$

## The generalization

*bartisan* keeps the sum of trees and replaces what sits on top of it. The forest
supplies an additive predictor $\eta(x)$, and the outcome follows whatever
density the family names,

$$\eta(x) = \sum_{m=1}^{M} g(x; T_m, M_m), \qquad
  Y \mid x \sim p\{y \mid \eta(x), \theta\},$$

with $\theta$ collecting any parameters of the density that are not functions of
$x$, such as a residual standard deviation or a shape. These are the nuisance
parameters, and they are drawn alongside the trees.

Some families need more than one function of $x$. A location-scale model has a
mean and a log standard deviation; a zero-inflated count has a rate and a
probability of a structural zero. Those get one forest each:

$$\eta_h(x) = \sum_{m=1}^{M_h} g(x; T_{hm}, M_{hm}), \qquad h = 1, \dots, H,$$

and the sampler cycles over all the forests. `vignette("families")` says how many
each family uses.

## The prior

Three pieces, all with defaults chosen so that the fit is regularized without
being tuned [@chipman2010].

**Tree structure.** A node at depth $d$ is split rather than left as a leaf with
probability

$$\alpha (1 + d)^{-\beta}, \qquad \alpha = 0.95, \; \beta = 2.$$

This falls off quickly, so trees are shallow: most have two or three leaves. It
is the main reason a forest of 50 trees does not overfit.

**Leaf values.** Each $\mu_b$ is normal with mean zero and a standard deviation
that shrinks as trees are added, so that the sum has a sensible scale whatever
$M$ is. The `k` argument in `bartisan_control()` sets how hard this shrinks:
larger `k` pulls the fit toward the intercept-only model. The default is a
compromise that works across a wide range of problems, and it is rarely worth
changing.

**Nuisance parameters.** Each family sets its own, for example a half-Cauchy on a
scale parameter.

The prior does the work that cross-validation does in other tree ensembles. There
is no learning rate, no depth limit, and no early stopping.

## Soft rules

A rule in classic BART is a step: an observation with $x_j \le c$ goes left and
everything else goes right. The fitted function is then a step function, which is
a poor description of a smooth truth.

Soft rules replace the step with a gate [@linero2018]. An observation goes left
with weight

$$\psi\left(\frac{x_j - c}{\tau}\right),$$

where $\psi$ is a cumulative distribution function and $\tau$ is a bandwidth, and
right with the remaining weight. Every observation reaches every leaf with some
weight, and the fitted function is smooth. As $\tau \to 0$ the gate becomes a
step and the hard rule is recovered.

The `gate` argument chooses $\psi$: `"smoothstep"` (the default),
`"smootherstep"`, `"logistic"`, or `"hard"`. The first two are polynomial and
reach exactly zero and one outside a finite window, which is faster than the
logistic and usually just as good. `bandwidth` sets the prior mean of $\tau$,
which is drawn rather than fixed.

Soft rules cost about three times as much per iteration and are usually worth
it:

```{r softhard}
friedman <- function(n) {
  x <- as.data.frame(matrix(runif(n * 10), n, 10))
  names(x) <- paste0("x", 1:10)
  x$eta <- 10 * sin(pi * x$x1 * x$x2) + 20 * (x$x3 - 0.5)^2 +
    10 * x$x4 + 5 * x$x5
  x
}

train <- friedman(n)
test  <- friedman(1000)
train$y <- train$eta + rnorm(n)

timed <- function(gate) {
  started <- proc.time()[["elapsed"]]
  fit <- bartisan(y ~ . - eta, data = train, family = gaussian(),
                  control = bartisan_control(num_trees = 20, num_burn = 300,
                                             num_draws = 300, gate = gate))
  data.frame(rules = gate,
             test_rmse = sqrt(mean((predict(fit, newdata = test) - test$eta)^2)),
             seconds = round(proc.time()[["elapsed"]] - started, 1))
}

rbind(timed("smoothstep"), timed("hard"))
```

The true function has a standard deviation of about
`r round(sd(test$eta), 1)`, so both are fitting real structure and the soft fit
is the more accurate.

## Sparsity

By default the variable a rule splits on is drawn from a categorical
distribution whose probabilities have a Dirichlet prior [@linero2018sparse]:

$$s \sim \mathrm{Dirichlet}(\alpha/p, \dots, \alpha/p),$$

with $\alpha$ itself drawn. Small $\alpha$ concentrates the probability on a few
predictors, so the forest can stop splitting on the rest entirely. This is DART,
and it is what `sparsity = TRUE` means. It is the default.

The alternative, `sparsity = FALSE`, gives every predictor the same splitting
probability, which is classic BART.

The difference matters for variable selection. @chipman2010 note that counting
splits works poorly when there are many trees, "because the redundancy offered by
so many trees tends to mix many irrelevant predictors in with the relevant ones",
and recommend reducing the number of trees so that predictors compete. The
Dirichlet prior addresses the same problem directly, by letting irrelevant
predictors be dropped rather than crowded out.

Both effects are visible on the Friedman function, where `x1` to `x5` matter and
`x6` to `x10` are noise. The table reports the average posterior probability of
being used, over three replicates at $n = 500$:

```{r sparsitytable, echo = FALSE}
knitr::kable(data.frame(
  Trees = c(10, 20, 50, 100),
  `noise, sparsity = FALSE` = c(0.28, 0.50, 0.95, 1.00),
  `noise, sparsity = TRUE` = c(0.09, 0.08, 0.09, 0.14),
  check.names = FALSE),
  caption = "Mean prop_used for the five noise predictors. The five real predictors sit at 1.00 in every cell.")
```

Without the sparsity prior the advice to use fewer trees is essential: at 50
trees the noise predictors are used in 95% of draws and are indistinguishable
from the real ones. With it, the noise predictors stay near zero at every tree
count, and reducing the trees buys little. The recommendation is a workaround for
the absence of the prior rather than a property of variable selection in general.

Two things this does not fix. Sparsity needs signal: on a small sample with a
weak fit nothing separates, and the honest reading of a flat table is that the
data cannot rule anything out. And it says nothing about effect size.
`vignette("importance")` covers both.

# How it is fitted

## Bayesian backfitting

The trees are updated one at a time. To update tree $m$, subtract the
contribution of every other tree from the predictor, leaving a partial residual
that only tree $m$ has to explain, then draw $(T_m, M_m)$ given that residual.
Cycling over all trees, and then over the nuisance parameters, is one sweep. The
draws saved after warmup are the posterior sample.

## Moving the trees

A tree is proposed by a small local change: growing a leaf into two, pruning a
pair of leaves back into one, or changing a splitting rule. The proposal is
accepted or rejected by a Metropolis-Hastings step. Because growing and pruning
change the number of parameters, this is a reversible-jump sampler.

The acceptance ratio needs the likelihood of the tree with the leaf parameters
integrated out,

$$p(\text{residual} \mid T) = \int p(\text{residual} \mid T, M) \, p(M) \, dM.$$

## Why the general case is hard, and what this package does

For a Gaussian outcome with a normal leaf prior that integral is a normal
integral and has a closed form. This is what confines classic BART to Gaussian
likelihoods, or to likelihoods that can be made Gaussian by augmentation.

For anything else the integral has no closed form. @linero2025 replaces it with a
Laplace approximation: expand the log integrand around its mode, integrate the
resulting quadratic, and use that both to score the tree and to propose the leaf
values. A Metropolis correction then makes the chain target the exact posterior.
The approximation only has to be good enough to be accepted often; it does not
have to be right.

The practical consequence is the reason the package exists. A family needs only
the log density of one observation and its first two derivatives with respect to
the predictor. Anything that can supply those can be fitted, which is why
`custom_family()` takes an R function and works.

The sampler recognizes three shapes of the leaf-level target and uses the
cheapest machinery each admits:

| Shape | Meaning | Cost |
| --- | --- | --- |
| Quadratic | the target is exactly Gaussian, so the Laplace step is exact | closed form, no rejection |
| Exponential | of the form $a\eta + b e^{r\eta}$, as for a count or a Weibull | one mode-solve |
| General | anything else | Newton iteration |

A family that reaches the quadratic shape is several times faster than one that
does not. This is what data augmentation is for.

## Data augmentation

Some likelihoods are not Gaussian but become Gaussian once an extra latent
variable is imagined and drawn alongside everything else.

- A **probit** model is a Gaussian one on a latent scale that is truncated by the
  observed category [@albert1993]. Draw the latent value and the target is
  exactly quadratic.
- A **logistic** model becomes Gaussian conditional on a Pólya-Gamma variable
  [@polson2013], which acts as a per-observation precision.
- A **censored** outcome becomes an ordinary one once the unobserved value is
  imputed from its truncated distribution.

`augment = TRUE`, the default where a scheme exists, turns these on. They change
the sampler and not the model: the posterior is the same, reached faster.
`vignette("families")` says which families have one.

## Nuisance parameters

Anything in the density that is not a function of $x$ is drawn once per sweep,
after the trees. A residual standard deviation, a negative binomial dispersion,
the cutpoints of an ordinal model, and the baseline hazard of a proportional
hazards model are all handled this way. They come back in `fit$aux` and are
summarized by `summary()`.

A likelihood written with `custom_family()` can have them too, by naming them in
`aux_names`. There is no prior argument: a parameter with a restricted range is
handled by writing the transform into the density, exactly as it would be for a
real predictor.

## The Dirichlet process mixture

Two families, `dpm()` and `dpm_aft()`, do not assume a shape for the error at
all. They model it as a mixture of normals with a Dirichlet process prior on the
mixing distribution [@escobar1995; @george2019], which lets the number of
components grow with the data. The mixture is constrained to have mean zero so
that the forest still carries the conditional mean.

Conditional on which component each observation belongs to, the model is Gaussian
and the target is quadratic, so the extra flexibility is cheap. `error_density()`
reports the fitted error distribution, which is the diagnostic the family exists
to provide. Because the mixture can collapse to a single component, it costs
little when a single normal was right.

# Practical matters

## Missing predictor values

Missing predictor values are handled natively, by treating missingness as
something to split on [@twala2008]. A rule can send missing values left, send
them right, or split on missingness itself, and which of these is used is part of
the posterior. No imputation is required and rows are not dropped.

```{r missing}
d_miss <- train
d_miss$x1[sample(n, 60)] <- NA

fit_miss <- bartisan(y ~ . - eta, data = d_miss, family = gaussian(),
                     control = ctrl)

sqrt(mean((predict(fit_miss, newdata = test) - test$eta)^2))
```

The fit uses the incomplete variable rather than discarding it. This is a
prediction tool and not a substitute for thinking about why values are missing;
it assumes the pattern of missingness is itself informative, which is sometimes
true and sometimes not.

## Several chains

`chains` runs the sampler more than once from different starting points. The
default is one, which is enough for estimates. Running several is what makes
`fit$rhat` available, which is the only way to see whether the sampler has
converged.

```{r chains}
fit_chains <- bartisan(y ~ . - eta, data = train, family = gaussian(),
                       chains = 4, control = ctrl)

fit_chains$rhat
```

Read this table selectively. `aux.sigma` and `loglik` are close to 1, which is
what you want, and they respond to longer chains in the usual way.

The `eta` row does not, and it is worth knowing why before it alarms you. It is
the worst value over all `r n` observations, so it is a maximum by construction,
and forests mix slowly on their fitted values: two chains can visit quite
different sets of trees that imply very similar predictions, which inflates a
between-chain statistic without meaning the answers disagree. Adding draws does
not reliably bring it down. Values above 1.05 on this row are ordinary for a
BART fit, and it is characteristic of the method rather than of this
implementation.

What to do instead is check that the quantities you will report are stable. If
`sigma` and the log likelihood have converged and an `avg_comparisons()` estimate
is the same across chains, the fit is usable whatever the `eta` row says.

Chains run sequentially unless a `future` plan is set, in which case they run in
parallel. `vignette("diagnostics")` covers this properly, including which
statistic to look at when the fitted values are the estimand.

## Choosing the settings

Most of `bartisan_control()` does not need to be touched. In rough order of how
often changing one is worth it:

| Setting | When to change it |
| --- | --- |
| `num_burn`, `num_draws` | when `rhat` says the chain has not converged |
| `num_trees` | more for a complex function and a large sample; fewer to speed up |
| `gate` | `"hard"` when the truth really is a step function, or for speed |
| `sparsity` | `FALSE` to recover classic BART, which is rarely what you want |
| `k` | to shrink harder toward the mean, in a very small sample |

Everything else exists so that the checks in the test suite can be written, or
because a specific family needs it.

The single decision that matters more than all of these is the family.
`vignette("families")` covers it.

## Where the time goes

Cost per sweep is roughly linear in the number of observations and the number of
trees. The family matters more than either: a family that reaches the quadratic
target is several times faster than one that does not, which is why
`lognormal_aft()` fits in about a second where `weibull_aft()` takes eight on the
same data.

# Further reading

@hill2020 is a review of BART covering the model, its extensions and its
applications. @chipman2010 is the original. @linero2018 introduces soft rules and
@linero2018sparse the sparsity prior. @linero2025 is the sampler this package
implements. For causal inference with BART, @hill2011 is the starting point and
@hahn2020 the standard treatment of what goes wrong when treatment assignment is
strongly predicted by the covariates. @murray2021 gives an alternative route to
count and multinomial outcomes.

# References
