---
title: "Generalized BART with bartisan"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Generalized BART with bartisan}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{=html}
<style>
pre {overflow-x: auto;}
pre code {word-wrap: normal; white-space: pre;}
</style>
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.6, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 3)

```

```{r libs}
library(bartisan)

set.seed(2026)
```

## Introduction

This vignette describes what the model is and how it is fitted. It is the reference document rather than the entry point: `vignette("bartisan")` shows how to use the package on a real analysis, and `vignette("families")` covers choosing a likelihood. This is the one to read to find out what the fit is doing, why the defaults are what they are, or what an advanced setting changes.

In this guide we will first set out the model, which is a sum of trees supplying an additive predictor to an arbitrary likelihood, and then the prior that regularizes it. Next we will describe how it is fitted, which is where the generalization actually lives, and what data augmentation buys where it is available. We then turn to the practical matters that come up in a real fit, to what the sampler costs against the other BART packages, and to the checks that say the implementation is right.

Fits below use a deliberately small chain so that the vignette builds quickly: 20 trees and 300 draws after 300 warmup iterations, on 400 observations. The defaults are 50 trees and 800 draws after 200 warmup iterations.

```{r control}
ctrl <- bartisan_control(num_trees = 20, num_burn = 300, num_draws = 300)

n <- 400
```

## The Model

### A Sum of Trees

A binary regression tree is a pair $(T, M)$, where $T$ is the tree structure, meaning the splitting rules at the interior nodes, and $M = \{\mu_1, \dots, \mu_B\}$ collects one parameter for each of its $B$ leaves. The tree defines a function $g(x; T, M)$ that sends $x$ down the tree according to the rules and returns the parameter in the leaf it reaches. A single tree is a step function.

BART sums many of them [@chipman2010]:

$$f(x) = \sum_{m=1}^{M} g(x; T_m, M_m).$$

Each tree is kept small by its prior, so no single one explains much. The sum is flexible while each term is a weak learner, which is what makes the fit stable. Interactions come free: a path through a tree that splits on $x_1$ and then on $x_2$ is an interaction between them, and nothing had to be specified for it to appear.

In ordinary BART, the sum is the conditional mean of a Gaussian outcome,

$$Y = f(x) + \varepsilon, \qquad \varepsilon \sim \mathrm{N}(0, \sigma^2).$$

### The Generalization

*bartisan* keeps the sum of trees and replaces what sits on top of it. The forest supplies an additive predictor $\eta(x)$, and the outcome follows whatever density the family names,

$$\eta(x) = \sum_{m=1}^{M} g(x; T_m, M_m), \qquad
  Y \mid x \sim p\{y \mid \eta(x), \theta\},$$

with $\theta$ collecting any parameters of the density that are not functions of $x$, such as a residual standard deviation or a shape. These are the nuisance parameters, and they are drawn alongside the trees.

Some families need more than one function of $x$. A location-scale model has a mean and a log standard deviation; a zero-inflated count has a rate and a probability of a structural zero. Those get one forest each:

$$\eta_h(x) = \sum_{m=1}^{M_h} g(x; T_{hm}, M_{hm}), \qquad h = 1, \dots, H,$$

and the sampler cycles over all the forests. `vignette("families")` says how many forests each family uses.

### The Prior

The model prior has three pieces, all with defaults chosen so that the fit is regularized without being tuned [@chipman2010].

**Tree structure.** A node at depth $d$ is split rather than left as a leaf with probability

$$\gamma (1 + d)^{-\beta}, \qquad \gamma = 0.95, \; \beta = 2,$$

which are the `gamma` and `beta` arguments of `bartisan_control()`.

This falls off quickly, so trees are shallow: most have two or three leaves. It is the main reason a forest of 50 trees does not overfit.

**Leaf values.** Each $\mu_b$ is normal with mean zero and a standard deviation that shrinks as trees are added, so that the sum has a sensible scale whatever $M$ is. The `k` argument in `bartisan_control()` sets how hard this shrinks: larger values of `k` pull the fit toward the intercept-only model. The default is a compromise that works across a wide range of problems, and it is rarely worth changing.

**Nuisance parameters.** Each family sets its own, for example a half-Cauchy on a scale parameter.

The prior does the work that cross-validation does in other tree ensembles. There is no learning rate, no depth limit, and no early stopping.

### Soft Rules (`gate`)

A rule in classic BART is a step: an observation with $x_j \le c$ goes left and everything else goes right. The fitted function is then a step function, which is a poor description of a smooth truth.

Soft rules replace the step with a gate (i.e., a smooth function of the distance from the cutpoint) [@linero2018]. An observation goes left with weight

$$\psi\left(\frac{x_j - c}{\tau}\right),$$

where $\psi$ is a cumulative distribution function and $\tau$ is a bandwidth, and right with the remaining weight. An observation near a cutpoint reaches both of that node's children, so the fitted function is smooth rather than a step. As $\tau \to 0$ the gate becomes a step and the hard rule is recovered.

The `gate` argument chooses $\psi$: `"smoothstep"` (the default), `"smootherstep"`, `"logistic"`, or `"hard"`. How far the smoothing reaches is what separates them. The logistic has unbounded support, so under it every observation really does reach every leaf with some weight. The two polynomial gates are the cumulative distribution functions of symmetric Beta kernels and reach exactly zero and one outside a finite window, so an observation far enough from a cutpoint takes one side only and the subtree on the other side is never visited. That skipped work turns out to be worth less than it sounds; what makes them faster is that a polynomial needs no `exp()`, and they keep that advantage even at a bandwidth wide enough to truncate nothing at all. `bandwidth` sets the prior mean of $\tau$, which is drawn rather than fixed, and it means the same amount of smoothing under every gate, the kernels being matched on their standard deviation rather than on their width.

Soft rules cost about three times as much per iteration but are usually worth it:

```{r softhard}
friedman <- function(n) {
  x <- as.data.frame(matrix(runif(n * 10), n, 10))
  names(x) <- paste0("x", 1:10)
  x$eta <- 10 * sin(pi * x$x1 * x$x2) + 20 * (x$x3 - 0.5)^2 +
    10 * x$x4 + 5 * x$x5
  x
}

train <- friedman(n)
test  <- friedman(1000)
train$y <- train$eta + rnorm(n)

timed <- function(gate) {
  started <- proc.time()[["elapsed"]]
  fit <- bartisan(y ~ . - eta, data = train, family = gaussian(),
                  control = bartisan_control(num_trees = 20, num_burn = 300,
                                             num_draws = 300, gate = gate))
  data.frame(rules = gate,
             test_rmse = sqrt(mean((predict(fit, newdata = test) - test$eta)^2)),
             seconds = round(proc.time()[["elapsed"]] - started, 1))
}

rbind(timed("smoothstep"), timed("hard"))
```

The true function has a standard deviation of about `r round(sd(test$eta), 1)`, so both are fitting real structure and the soft fit is the more accurate.

### The Predictor Transform (`x_transform`)

Numeric predictors are mapped to $[0, 1]$ before any rule sees them, and the map is not cosmetic. A cutpoint is drawn uniformly on a node's live range in the transformed coordinate, so the map decides where a rule *can* split, and under soft rules it is also the scale on which $\tau$ is measured. The `x_transform` argument of `bartisan_control()` chooses it. The three options fail in different places, which is the whole of the choice between them.

`"range"` rescales linearly, $T(x) = (x - \min x) / (\max x - \min x)$, and so preserves the original spacing. That makes it the most accurate of the three on a well-behaved predictor, and, being the only affine map among them, the only one a derivative passes through cleanly. Its failure is a predictor whose bulk sits inside a sliver of its range, which a few extreme values are enough to produce: a cutpoint drawn uniformly on the range then almost never lands where the structure is. In a simulation with 1% of a predictor at 300 times the scale of the rest and a truth that oscillates within the bulk, it was 4 times worse than the alternatives at $n = 500$ and 6 times worse at $n = 2000$, with 95% intervals covering .24 rather than .95. The deterioration with $n$ is the signature: the extremes grow with the sample, so the bulk occupies an ever smaller share of the range.

`"quantile"` maps each predictor through its own empirical distribution function, which is what `SoftBart::softbart()` does. It cannot fail that way, because it places cutpoints by rank and knows nothing about spacing. It fails instead by being a step function, so the fit is a step function of the original predictor: there is no derivative to take, and a relationship that is straight in the predictor becomes a staircase wherever the data have gaps. On two tight clusters with a linear truth it was nearly three times worse than the alternatives.

`"smoothcdf"`, the default, smooths that same distribution function and so does neither. The estimate is

$$\hat F_h(x) = \frac{1}{n} \sum_{i=1}^n K\left(\frac{x - x_i}{h}\right),$$

where $K$ is the integrated Epanechnikov kernel and $h = 2.1 \hat\sigma n^{-1/3}$, with $\hat\sigma$ the smaller of the standard deviation and the interquartile range over 1.349 so that a long tail sets the bandwidth from the bulk rather than from itself. The $n^{-1/3}$ rate is the one for a distribution function rather than the $n^{-1/5}$ that is right for a density [@tenreiro2006], a rate @azzalini1981 had pointed out for a second-order approximation. The result is stretched affinely so that the smallest and largest observations land on 0 and 1, as they do under the other two maps; normalizing on the smoothed values instead would leave a few percent of the coordinate outside the data, where no cutpoint can be useful. Cutpoints then land where the data are, as under `"quantile"`, and the map is strictly increasing and differentiable, as under `"range"`. Across eight data-generating processes it was never the most accurate and never far from it, where each of the other two was badly wrong somewhere.

It is not a free lunch for slopes. Writing the fit as $f(T(x))$, a derivative in the original coordinate is $f'(T(x)) T'(x)$: an affine $T$ has a known constant derivative, so only $f'$ is estimated, where a smoothed distribution function contributes an estimated density and the two errors multiply. `vignette("effects")` measures what that costs and recommends refitting with `"range"` when a slope is the quantity being reported.

The kernel is Epanechnikov rather than Gaussian because its support is bounded, which is what makes the estimate cheap to compute exactly. Written directly, $\hat F_h$ on a grid of $m$ points is $nm$ kernel evaluations, tens of millions per predictor on a large fit. Almost all of that work is avoidable, because almost every weight is a 0 or a 1 rather than something in between: an observation further below a grid point than the kernel reaches has passed it entirely and contributes 1, one further above contributes 0, and only those within $h$ need evaluating. With the data and the grid both sorted, the two ends of that window advance monotonically, so each is found by a pointer that never goes back and the sweep costs $O(n + m)$.

### Sparsity (`sparsity`)

By default, the variable a rule splits on is drawn from a categorical distribution whose probabilities have a Dirichlet prior [@linero2018sparse]:

$$s \sim \mathrm{Dirichlet}(\alpha/p, \dots, \alpha/p),$$

with $\alpha$ itself drawn. Small $\alpha$ concentrates the probability on a few predictors, so the forest can stop splitting on the rest entirely. This is DART, and it is what `sparsity = TRUE` means. It is the default.

The alternative, `sparsity = FALSE`, gives every predictor the same splitting probability, which is classic BART.

The difference matters for variable selection. @chipman2010 note that counting splits works poorly when there are many trees, "because the redundancy offered by so many trees tends to mix many irrelevant predictors in with the relevant ones", and recommend reducing the number of trees so that predictors compete. The Dirichlet prior addresses the same problem directly, by letting irrelevant predictors be dropped rather than crowded out.

Both effects are visible on the Friedman function, where `x1` to `x5` matter and `x6` to `x10` are noise. The table reports the average posterior probability of being used, over three replicates at $n = 500$:

```{r sparsitytable, echo = FALSE}
knitr::kable(data.frame(
  Sparsity = c("`sparsity = FALSE`", "`sparsity = TRUE`"),
  `10 trees` = c(0.28, 0.09),
  `20 trees` = c(0.50, 0.08),
  `50 trees` = c(0.95, 0.09),
  `100 trees` = c(1.00, 0.14),
  check.names = FALSE),
  caption = "Mean prop_used for the five noise predictors. The five real predictors sit at 1.00 in every cell.")
```

Without the sparsity prior, the advice to use fewer trees is essential: at 50 trees the noise predictors are used in 95% of draws and are indistinguishable from the real ones. With it, the noise predictors stay near zero at every tree count, and reducing the trees buys little. The recommendation is a workaround for the absence of the prior rather than a property of variable selection in general. However, sparsity needs signal, so on a small sample with a weak fit, nothing separates, and the honest reading of a flat table is that the data cannot rule anything out; and it says nothing about effect size, since a predictor can be split on constantly and move the prediction very little. `vignette("importance")` covers both.

## The Sampler

### Bayesian Backfitting

The trees are updated one at a time. To update tree $m$, subtract the contribution of every other tree from the predictor, leaving a partial residual that only tree $m$ has to explain, then draw $(T_m, M_m)$ given that residual. Cycling over all trees, and then over the nuisance parameters, is one sweep. The draws saved after warmup are the posterior sample.

### Moving the Trees

A tree is proposed by a small local change: growing a leaf into two, pruning a pair of leaves back into one, or changing a splitting rule. The proposal is accepted or rejected by a Metropolis-Hastings step. Because growing and pruning change the number of parameters, this is a reversible-jump sampler.

The acceptance ratio needs the likelihood of the tree with the leaf parameters integrated out,

$$p(\text{residual} \mid T) = \int p(\text{residual} \mid T, M) \, p(M) \, dM$$

### The General Case, and the Laplace Approximation

For a Gaussian outcome with a normal leaf prior, that integral is a normal integral and has a closed form. This is what confines classic BART to Gaussian likelihoods, or to likelihoods that can be made Gaussian by augmentation.

For anything else, the integral has no closed form. @linero2025 replaces it with a Laplace approximation: expand the log integrand around its mode, integrate the resulting quadratic, and use that both to score the tree and to propose the leaf values. A Metropolis correction then makes the chain target the exact posterior. The approximation only has to be good enough to be accepted often; it does not have to be right.

The practical consequence is the reason the package exists. A family needs only the log density of one observation and its first two derivatives with respect to the predictor. Anything that can supply those can be fit, which is why `custom_family()` takes an R function and works.

The sampler recognizes three shapes of the leaf-level target and uses the cheapest machinery each admits:

| Shape | Meaning | Cost |
|------------------------|------------------------|------------------------|
| Quadratic | the target is exactly Gaussian, so the Laplace step is exact | closed form, no rejection |
| Exponential | of the form $a\eta + b e^{r\eta}$, as for a count or a Weibull | one mode-solve |
| General | anything else | Newton iteration |

A family that lands higher in this table is several times faster than one that does not. This is what data augmentation is for.

### Data Augmentation (`augment`)

Some likelihoods move up the table once an extra latent variable is imagined and drawn alongside everything else. The variable is not part of the model; it is machinery, and the posterior over everything else is unchanged.

Several such rewritings are in use here, and they do not all arrive at the same shape. **A probit model** is a Gaussian one on a latent scale truncated by the observed category [@albert1993], so drawing the latent value makes the target exactly quadratic. **A logistic model** is Gaussian conditional on a Pólya-Gamma variable [@polson2013], which acts as a per-observation precision, and an ordinal complementary log-log model is Gaussian conditional on an exponential waiting time. **A censored outcome** becomes an uncensored one once the unobserved value is imputed from its truncated distribution, which for the log-normal and log-logistic survival families leaves a quadratic target. **A negative binomial** is a Poisson whose rate carries a gamma prior, so drawing the rate leaves the *exponential* shape rather than the quadratic one, which is a smaller gain. And **a zero-inflated count** is blocked by a mixture rather than by a link, so what is drawn there is the indicator of which component produced each observation; that separates the two forests, and for a zero-inflated negative binomial a second rewriting goes on top of the first.

`augment = TRUE`, which is the default wherever a rewriting exists, turns these on. They change the sampler and not the model: the posterior is the same, reached faster. `vignette("families")` says which families have one, and the section on speed below measures what each is worth.

### Nuisance Parameters

Anything in the density that is not a function of $x$ is drawn once per sweep, after the trees. A residual standard deviation, a negative binomial dispersion, the cutpoints of an ordinal model, and the baseline hazard of a proportional hazards model are all handled this way. They come back in `fit$aux` and are summarized by `summary()`.

A likelihood written with `custom_family()` can have them too, by naming them in `aux_names`. There is no prior argument: a parameter with a restricted range is handled by writing the transform into the density, exactly as it would be for a real predictor.

### The Dirichlet Process Mixture

Two families, `dpm()` and `dpm_aft()`, do not assume a shape for the error at all. They model it as a mixture of normals with a Dirichlet process prior on the mixing distribution [@escobar1995; @george2019], which lets the number of components grow with the data. The mixture is constrained to have mean zero so that the forest still carries the conditional mean.

Conditional on which component each observation belongs to, the model is Gaussian and the target is quadratic, so the extra flexibility is cheap. `error_density()` reports the fitted error distribution, which is the diagnostic the family exists to provide. Because the mixture can collapse to a single component, it costs little when a single normal was right.

## Practical Matters

### Missing Predictor Values

Missing predictor values are handled natively, by treating missingness as something to split on [@twala2008; for BART, @kapelner2015]. A rule can send missing values left, send them right, or split on missingness itself, and which of these is used is part of the posterior. No imputation is required and rows are not dropped.

```{r missing}
d_miss <- train
d_miss$x1[sample(n, 60)] <- NA

fit_miss <- bartisan(y ~ . - eta, data = d_miss, family = gaussian(),
                     control = ctrl)

sqrt(mean((predict(fit_miss, newdata = test) - test$eta)^2))
```

The fit uses the incomplete variable rather than discarding it. Note that this is a prediction tool and not a substitute for thinking about why values are missing: what it estimates is the mean of the response given the predictors and the pattern of missingness, which is the quantity prediction calls for, and which is not the regression or causal effect defined on complete data that multiple imputation targets.

### Several Chains (`chains`)

`chains` runs the sampler more than once from different starting points. The default is one, which is enough for estimates. Running several is what makes a between-chain comparison possible, which is the only way to see whether the sampler has converged, and `diagnose()` is where that comparison is computed. It is not part of the fit: the per-observation statistics cost more than the sampling does, so they run when they are asked for.

```{r chains}
fit_chains <- bartisan(y ~ . - eta, data = train, family = gaussian(),
                       chains = 4, control = ctrl)

diagnose(fit_chains)$table
```

This table is read selectively. `aux.sigma` and `loglik` are close to 1, which is what convergence looks like, and they respond to longer chains in the usual way.

The `eta` row does not, and it is worth knowing why before it causes alarm. It summarizes the worst 5% of the observations rather than the single worst, so it is a high percentile by construction, and forests mix slowly on their fitted values: two chains can visit quite different sets of trees that imply very similar predictions, which inflates a between-chain statistic without meaning that the answers disagree. Adding draws does not reliably bring it down. Values above 1.05 on this row are ordinary for a BART fit, and are characteristic of the method rather than of this implementation.

What to check instead is the quantity that will be reported, on its own draws. `diagnose()` takes the output of `estimate_effect()` for exactly this, and `posterior::summarise_draws()` takes anything else that can be arranged as draws by chains. A high `eta` row is not on its own a reason to discard a fit, but neither is a low one a reason to trust an effect read off it: a contrast can mix badly where the function it contrasts mixes well, so it has to be diagnosed rather than inferred.

Chains run sequentially unless a `future` plan is set, in which case they run in parallel:

``` r
future::plan("multisession", workers = 4)

fit <- bartisan(y ~ ., data = d, chains = 4)
diagnose(fit)
```

The chain is the only parallel axis the *sampler* has, since a sweep conditions on the one before it, and it is also what makes a convergence diagnostic possible. Outside the sampler, a plan is also used when one is set: the convergence pass splits its per-observation columns across workers, and `estimate_effect()` predicts the treatment levels in parallel. Any *future* backend works, including *mirai*'s, and one `set.seed()` reproduces the whole run whatever the backend.

`diagnose()` is the one call for whether the fit can be reported. It computes split-R-hat and the bulk and tail effective sample sizes for every scalar the sampler draws, for the fitted function twice over, once averaged across observations and once over its worst 5% of them, and for the size of the forest itself, then says which of them fall short and what to change. Handed the output of `estimate_effect()` instead of a fit, it reports on the contrast and on the two potential outcomes it is a contrast of. Two of its choices are worth knowing about. It repeats R-hat on the second half of the draws alone, which is what distinguishes a warmup that ended too early from chains that have each settled somewhere different: discarding the early draws is what more `num_burn` would have done, so if that fixes R-hat then warmup was the problem, and the advice says so rather than listing everything a reader might try. And it leaves the leaf scale out, because that parameter mixes badly in every implementation (1.12 in *dbarts* and 1.16 in *stochtree* against 1.19 here, on the same data) and its disagreement never reaches the fitted function.

A within-chain drift test would answer the warmup question more directly, and is not offered, because it cannot be made to work at a forest's autocorrelation. Three versions were calibrated against stationary series, where there is nothing to find by construction: taking each half's Monte Carlo error from that half alone fires 31% of the time at an autocorrelation of .995 against a nominal 5%, taking it from the whole chain holds specificity under 3% but then misses a six-standard-deviation trend three times in four, and batch means catch everything and fire 92% of the time on a chain that has converged. Re-running a statistic that is already calibrated sidesteps the choice.

`vignette("diagnostics")` covers all of this from the reader's side, including which statistic to look at when the fitted values are the estimand.

### Choosing the Settings

Most of `bartisan_control()` does not need to be touched. In rough order of how often changing one is worth it:

| Setting | When to change it |
|------------------------------------|------------------------------------|
| `num_burn`, `num_draws` | when `rhat` says the chain has not converged |
| `num_trees` | more for a complex function and a large sample; fewer to speed up |
| `gate` | `"hard"` when the truth really is a step function, or for speed |
| `x_transform` | `"range"` when a slope is the quantity being reported |
| `sparsity` | `FALSE` to recover classic BART, which is rarely the goal |
| `k` | to shrink harder toward the mean, in a very small sample |

Everything else exists so that the checks in the test suite can be written, or because a specific family needs it.

The single decision that matters more than all of these is the family. `vignette("families")` covers it.

## Speed

A non-conjugate sampler costs more than a conjugate one, and it is worth being precise about how much, and about what the extra buys. Cost per sweep is roughly linear in the number of observations and in the number of trees, but the family matters more than either: one that reaches the quadratic target is several times faster than one that does not, which is why `lognormal_aft()` fits in about a second where `weibull_aft()` takes eight on the same data.

### Gains From Data Augmentation (`augment`)

A major contributor to *bartisan*'s speed is data augmentation. Where the rewriting reaches a Gaussian conditional the Laplace approximation stops being an approximation at all, which is the largest of the gains: `augment` uses @albert1993 for a probit link and for an ordinal probit, Pólya-Gamma augmentation [@polson2013] for a logit link, an ordinal logit, or a multinomial, and imputation of the censored failure times for the log-normal and log-logistic survival models. The negative binomial and the zero-inflated families are rewritten too, but to a Poisson rather than to a Gaussian, and the two paragraphs after the table say what that costs them.

Each of them is faster, from 1.2 times for the negative binomial under soft rules to 31 for an ordinal probit with hard rules, with a median of 8.7. What each costs in mixing is the question, and the answer is not that they all cost something: over 15 replicate datasets the worst-mixing quantity's effective sample size came out a median of 0.89 times what the direct likelihood gives, and of the seventeen cases below, seven have an 80% interval lying entirely below 1, three lie entirely above it, and seven include it. Effective draws per second, which is the ratio to judge, favors augmentation in fifteen of the seventeen, the two negative-binomial rows being the exceptions:

| Family | Speed | ESS, worst quantity | ESS, median quantity | ESS per second |
|-----------------------|-------|--------------------|-----------|-----------|
| `lognormal_aft()`, soft rules | 11x | 1.49x [1.14, 1.68] | 1.06x | **16x** |
| `ordinal("probit")`, hard rules | 31x | 0.50x [0.37, 0.84] | 0.53x | **16x** |
| `ordinal("probit")`, soft rules | 24x | 0.64x [0.50, 0.67] | 0.57x | **15x** |
| `multinomial()`, soft rules | 8.1x | 1.49x [1.33, 2.04] | 0.87x | **12x** |
| `loglogistic_aft()`, hard rules | 8.7x | 1.21x [0.96, 2.56] | 0.98x | **11x** |
| `binomial("probit")` | 9.8x | 1.13x [0.97, 1.27] | 1.09x | **10x** |
| `multinomial()`, hard rules | 12x | 0.89x [0.81, 0.99] | 0.84x | **10x** |
| `lognormal_aft()`, hard rules | 14x | 0.61x [0.45, 1.22] | 0.88x | **8.8x** |
| `binomial("logit")` | 6.7x | 1.31x [1.04, 1.99] | 1.11x | **8.8x** |
| `zi_negbin()`, hard rules | 6.7x | 1.02x [0.79, 1.22] | 1.02x | **6.6x** |
| `ordinal("logit")`, hard rules | 15x | 0.43x [0.38, 0.64] | 0.49x | **6.5x** |
| `ordinal("logit")`, soft rules | 12x | 0.50x [0.46, 0.65] | 0.61x | **6.3x** |
| `zi_poisson()`, hard rules | 5.8x | 1.15x [0.74, 1.42] | 1.13x | **5.9x** |
| `loglogistic_aft()`, soft rules | 6.9x | 0.67x [0.54, 0.87] | 0.74x | **4.6x** |
| `zi_poisson()`, soft rules | 3.7x | 1.14x [0.96, 2.25] | 1.30x | **4.5x** |
| `negbin()`, hard rules | 1.8x | 0.52x [0.47, 0.97] | 0.84x | 1.0x |
| `negbin()`, soft rules | 1.2x | 0.66x [0.49, 1.06] | 0.83x | 0.8x |

Every cell is the median over replicates of the ratio *within* a replicate, which is the reading the design supports: both arms see the same data, so the paired ratio is the quantity with the least noise in it, and a median rather than a mean because a wall-clock timing can be disturbed by work outside the fit. The bracketed range is an 80% bootstrap interval over the replicates, and it is in the table because without it the column invites conclusions it cannot support: at three replicates the same quantity varied by a median factor of 4.5 within a single cell, since the worst-quantity effective sample size is a minimum over many quantities and so is a high-variance thing to estimate.

`augment` is on by default for every family that has a rewriting: the binomial, ordinal, multinomial, negative binomial, zero-inflated, and survival families. The negative binomial's is a Poisson whose rate comes from a gamma, which needs one gamma draw per observation rather than a Pólya-Gamma one, and it is the one family whose two rows differ in kind rather than degree. Under hard rules the target reaches the exponential form and the rewriting buys time at a real cost in mixing; under soft rules it cannot, and the rewriting instead buys a little time at no measurable cost in mixing. Neither comes out ahead on effective draws per second, at 1.0 and 0.8 times, which makes the negative binomial the one family whose rewriting does not pay for itself; `augment = FALSE` is worth trying whenever a negative binomial fit's diagnostics look poor.

The survival families are the clearest case of what the rewriting is for. Right-censoring is what makes their likelihood expensive: an observed failure contributes a density in the additive predictor and a censored one contributes a survival function, and the two have different shapes, so nothing about the target is exploitable. Imputing each censored failure time above its censoring time replaces the survival term with a density, and then every observation contributes the same quadratic shape. `weibull_aft()` is absent from the table because it needs none of this, its likelihood being already exponential in the sense above, censoring included, which also makes it the slowest of the three at the default soft gate, where that form cannot be used.

The zero-inflated families are the odd ones out, because what blocks them is a mixture rather than a link: the zero contributes $\log[\pi + (1 - \pi) P_0]$, so neither the count predictor nor the inflation predictor has a shape to exploit. Introducing the indicator of *which* component produced each observation separates them, and the count forest then sees a plain Poisson or negative binomial, and the inflation forest a Bernoulli logistic likelihood that Pólya-Gamma handles.

The ordinal probit is the largest of these because the target it replaces is the most expensive one in the package: two cumulative-normal evaluations per observation per pass, and no exploitable shape, so every trial value of a leaf parameter costs its own pass. It comes with one subtlety worth naming. The cutpoints are *not* drawn from their conditional given the latent normals, which would be uniform between the two order statistics that bracket each one, pinning them to an interval of width $O(1/n)$ and mixing worse the more data there is. They are drawn from the ordinal likelihood with the latent normals integrated out, and the latent normals are redrawn immediately afterward: a partially collapsed Gibbs sampler [@vandyk2008], which is the standard remedy [@cowles1996].

The ordinal logit has no latent normal, its latent variable being logistic, and the usual route to one goes through the Kolmogorov-Smirnov density, which needs a sampler of its own. However, @polson2013's Theorem 1 at $a = 1$, $b = 2$ says that the standard logistic density is $\frac{1}{4} E[\exp(-w x^2 / 2)]$ with $w \sim \mathrm{PG}(2, 0)$, so a logistic residual is a normal whose precision is Pólya-Gamma; and because $\mathrm{PG}(b, c)$ is $\mathrm{PG}(b, 0)$ tilted by $\exp(-c^2 w / 2)$, the precision's conditional given a residual $r$ is exactly $\mathrm{PG}(2, \lvert r \rvert)$. That is an integer-parameter draw, which the exact Devroye sampler already in the package covers, so the logit link gets the same quadratic target with nothing approximate anywhere. The Poisson and gamma families need no rewriting, their targets being already in a shape the sampler collapses to a single pass.

## Correctness

The test suite includes some checks worth knowing about.

Shrinking the prior weights to nothing makes the likelihood constant, so the posterior collapses to the tree prior. The sampled trees then reproduce the prior's branching process to within a fraction of a percent, which is a direct check that the birth, death, and change moves satisfy detailed balance.

Every family's analytic score is checked against a central difference of its own log density, which is an independent route to the same quantity. This is the check that matters for the sampler, since a wrong score sends the Fisher-scoring proposal to the wrong place.

Over 40 replicate datasets at $n = 500$ with 10 predictors and the default settings, 95% credible intervals for the additive predictor covered the truth .964 of the time for a Gaussian response, .961 binomial, .973 Poisson, and .970 gamma, each with a Monte Carlo standard error near .006. Nominal coverage is .95, so these intervals are mildly conservative rather than deficient. The ratio of absolute bias to posterior standard deviation runs .69 to .74, so the posterior spread is comfortably larger than the distance from the truth, which is the same fact read a second way. Four chains and five times the draws move neither number, so mixing is not what sets them. All of this is one data-generating process: a mean function of moderate spread mapped to each family's link, which for the binomial keeps the success probabilities informative. A response whose probabilities sit near 0 or 1 carries much less information per observation, and pointwise intervals for the regression function are best treated as approximate on that account rather than because of anything measured here.

## Comparison With Other BART Packages

Below we compare *bartisan*'s capabilities to those of other popular BART R packages: [*dbarts*](https://cran.r-project.org/package=dbarts) 0.9.34, [*BART*](https://cran.r-project.org/package=BART) 2.9.10, [*flexBART*](https://cran.r-project.org/package=flexBART) 2.0.3, [*SoftBart*](https://cran.r-project.org/package=SoftBart) 1.0.3, [*bartMachine*](https://cran.r-project.org/package=bartMachine) 1.4.2, and [*stochtree*](https://cran.r-project.org/package=stochtree) 0.4.5. A dash means the package does not offer the feature.


::: wide-table
|   | bartisan | dbarts | BART | flexBART | SoftBart | bartMachine | stochtree |
|---------|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|
| **Responses** |  |  |  |  |  |  |  |
| Gaussian | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ |
| Binary, probit | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ |
| Binary, logit | ✓ | — | ✓ | ✓ | — | — | — |
| Count: Poisson, negative binomial | ✓ | — | — | — | — | — | — |
| Gamma, beta, ordered beta, Tweedie | ✓ | — | — | — | — | — | — |
| Ordinal | ✓ 3 links | — | — | — | — | — | ✓ cloglog |
| Multinomial | ✓ logit, probit | — | ✓[^1] | — | — | — | — |
| Zero-inflated counts | ✓ | — | — | — | — | — | — |
| Survival | ✓ 3 AFT, PH | — | ✓ **+ recurrent, competing risks** | — | — | — | — |
| Heteroskedastic | ✓ | — | — | ✓ | — | — | ✓ |
| Error distribution itself modeled (DPM) | ✓ | — | — | — | — | — | — |
| A likelihood written by the user | ✓ | — | — | — | — | — | — |
| **Rules and priors** |  |  |  |  |  |  |  |
| Hard decision rules | ✓ | ✓ | ✓ | ✓ | — | ✓ | ✓ |
| Soft decision rules | ✓ 4 gates | — | — | — | ✓ | — | — |
| Dirichlet sparsity prior (DART) | ✓ | — | ✓ | ✓ | ✓ | — | — |
| Splitting weights fixed by the user | ✓ | — | — | — | ✓ groups | ✓ | — |
| Categorical splits on level subsets | ✓ | — | — | ✓ **+ nested, network** | — | — | — |
| Missing predictors, no imputation | ✓ default | — | — | — | — | ✓ | — |
| **Structure** |  |  |  |  |  |  |  |
| Random intercepts | ✓ | ✓ | — | — | — | — | ✓ |
| Varying coefficients | ✓ | — | — | ✓ | ✓ | — | ✓ |
| Treatment-effect (BCF) structure | ✓ | — | — | — | — | — | ✓ |
| Formula interface | ✓ | ✓ | — | ✓ | — | — | — |
| **Running it** |  |  |  |  |  |  |  |
| Multiple chains | ✓ | ✓ | ✓ | ✓ | — | ✓ | ✓ |
| Threads inside one chain | — | ✓ | ✓ | — | — | ✓ | ✓ |
| Grow-from-root warm start | — | — | — | — | — | — | ✓ |
| Cross-validation over hyperparameters | — | ✓ | — | — | — | ✓ | — |
| Save and reload a fitted model | RDS | — | — | — | — | ✓ | ✓ JSON |
| **Reading the fit** |  |  |  |  |  |  |  |
| `predict()` on new data | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ |
| Convergence diagnostics built in | ✓ | — | ✓ | — | — | ✓ | — |
| Variable importance | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ | — |
| Formal variable-selection test | ✓ inclusion | — | ✓ permutation | — | ✓ inclusion | ✓ permutation | — |
| Partial dependence | ✓ | ✓ | — | — | ✓ | ✓ | — |
| Interaction detection | ✎ | — | — | — | — | ✓ | — |
| Counterfactual estimands with intervals | ✓ | ✎ (through [*bartCause*](https://cran.r-project.org/package=bartCause)) | — | — | — | — | ✎ (through [*bartCause*](https://cran.r-project.org/package=bartCause)) |
| ATE, ATT or per-unit treatment effects | ✓ | ✎ (through [*bartCause*](https://cran.r-project.org/package=bartCause)) | — | — | — | — | ✎ (through [*bartCause*](https://cran.r-project.org/package=bartCause)) |
| Cross-validated model comparison (`loo`) | ✓ | — | — | — | — | — | — |
:::


[^1]: *BART*'s support is a set of binary fits rather than a multinomial model, which is worth knowing before the checkmark is read as a like-for-like. `BART::mbart2()` fits $K$ independent one-vs-rest binary probit or logit BARTs, each to the indicator `(y == h)`, and then normalizes their latent predictors afterward, so no forest is ever fitted to a multinomial likelihood. `BART::mbart()` instead fits $K - 1$ binary models to nested subsets, the $h$th to `y == h` among the observations with `y >= h`, and multiplies them as a continuation-ratio. That second one is an exact factorization of the multinomial mass function, so it is a coherent model, but each conditional carries its own independent forest and prior, and the factorization runs over the categories in sorted order, so the prior it puts on a probability vector is not exchangeable in the categories. *bartisan*'s `multinomial()` couples the forests through the likelihood itself instead: conditional on the others, the likelihood for category $j$ is exactly binomial-logistic in $\eta_j - \log C_j$ with $C_j$ summing $e^{\eta}$ over the rest, which is what the Pólya-Gamma augmentation exploits and what keeps the prior symmetric in the categories [@murray2021].

**✎ means a helper package covers it, not this one.** Interaction detection is the remaining one, and *bartisan* does it through *marginaleffects*, because a fit works with it and every estimand there is computed by pushing the draws through: `avg_comparisons(variables = "x", by = "z")` and `hypotheses(~pairwise)` are interaction detection, and neither needs code here. `partial_dependence()` averages the fitted surface over the sample at each value of one or two predictors, and `plot()` on a fit is the same thing drawn. `marginaleffects::plot_predictions()` remains the one to reach for when the grid needs more control than that.

The variable-selection row divides differently, since different procedures get called a test there. One is the posterior inclusion probability under the Dirichlet sparsity prior, thresholded at .5 to give the median probability model of @barbieri2004. *SoftBart* reports it as `posterior_probs()`, and it is `colMeans(var_counts > 0)` on the splitting counts: the `prop_used` column of `variable_importance()` is that same quantity, under a prior that is on by default here as it is there. The other is a permutation test [@bleich2014], which refits the model on responses whose link to the predictors has been broken and thresholds the splitting shares against the resulting null rather than against .5. *bartMachine* offers all three of its thresholds and *BART* offers the third one, `BART::mc.wbart.gse()`; *bartisan* offers none of them.

The other columns worth reading as gaps rather than as differences are these. There are no threads inside a chain, so a single chain is single-core here where *dbarts*, *BART*, *bartMachine*, and *stochtree* all use several; chains do run in parallel, but it does not help one chain. There is no cross-validation over `k`, `num_trees`, or the tree prior, which `dbarts::xbart()` and `bartMachine::bartMachineCV()` both automate. There is no grow-from-root warm start, which is *stochtree*'s way of shortening burn-in. And there is no JSON serialization, so a fit round-trips through `saveRDS()` and not into another language. On survival, *BART* is ahead: recurrent events and competing risks are there and here they are not.

Some other packages are single-purpose rather than general, so they are not columns above: [*bcf*](https://CRAN.R-project.org/package=bcf) fits Bayesian causal forests only, [*VCBART*](https://github.com/skdeshpande91/VCBART) varying-coefficient models only, and *bartCause* wraps *dbarts* and *stan4bart* for causal estimands.

## Notes and Limitations

Soft rules cost more per iteration than hard ones because an observation near a cutpoint has to be followed down both sides of it rather than into one cell, so a leaf that a hard rule would skip still has to be evaluated. The gap is about a factor of $B$, the number of leaves, and the reason it comes out near three rather than near a hundred is the tree prior: it keeps trees to about two and a half leaves on average, so there is not much for the softening to multiply.

`tweedie()` has no closed form density at a positive response, and normalizing it takes an infinite series [@dunn2005]. However, writing the log density in exponential-dispersion form [@jorgensen1987] separates it into a part that moves with the predictor, which is closed form, and a normalizing term that does not depend on the predictor at all; the second goes in the eta-free part, which is evaluated once per sweep rather than at every leaf, and cancels from every acceptance ratio in between. Its length depends on the response and the dispersion but never on the mean, so it does not grow as the forest moves. Drawing `power` costs about 45% again, because the slice sampler has to re-sum the series at each candidate value and cannot use the table of log-gammas that a fixed power allows.

`ordbeta()` and `Beta()` are expensive families because every evaluation of their log density needs two log-gamma calls that no amount of restructuring removes. `ordinal()` is at the other end, and only because of augmentation: all three of its links have a latent-variable representation that the sampler uses by default (a normal for the probit, a normal with a Pólya-Gamma precision for the logit, and an exponential waiting time for the complementary log-log), which is what makes it almost as fast as `binomial()`.

The leaf scale is raised from near zero over the first quarter of warmup. @linero2025 describes this as essential: started at its full value, the sampler can settle early into a poor configuration and fail to move. It is controlled by `sigma_mu_ramp` in `bartisan_control()`.

Nuisance parameters that are weakly identified, such as the negative binomial dispersion or the shape of a survival model, mix more slowly than the regression function, and the data often do not pin them down. Their draws are worth checking before they are interpreted.

The leaf scale is drawn under a half-Cauchy prior, which has no upper bound. When the predictors nearly separate a binary response the likelihood rewards an unbounded predictor and that prior does not hold the scale down: in a fully separated example the drawn scale wandered between 3 and 9 times its prior median over 1600 draws without settling. `bartisan()` warns when the scale settles more than five times above its prior median, and `update_sigma_mu = FALSE` pins it.

A link supplied from R costs a call into the interpreter for every leaf the sampler visits, and `custom_family()` costs three unless the derivatives are supplied. Both are usable at the sizes a BART model is usually fit at, and both are slower than the compiled equivalent.

## Further Reading

@hill2020 is a review of BART covering the model, its extensions and its applications. @chipman2010 is the original. @linero2018 introduces soft rules and @linero2018sparse the sparsity prior. @linero2025 is the sampler this package implements. For causal inference with BART, @hill2011 is the starting point and @hahn2020 the standard treatment of what goes wrong when treatment assignment is strongly predicted by the covariates. @murray2021 gives an alternative route to count and multinomial outcomes.

## References
