---
title: "Which variables matter"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Which variables matter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.4, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 3)

has_me <- rlang::is_installed("marginaleffects")
run <- rlang::is_installed("MASS")
```

## Introduction

A fitted forest spends its splitting rules on some predictors and not others, and
counting how it spends them is the usual measure of variable importance for BART.
`variable_importance()` reports the count.

This vignette covers what that number means, how to tell whether a difference in
it is real, and three things it is not. The last part matters more than the
first: variable importance is the most over-read output in machine learning, and
the failure modes are specific and checkable.

```{r libs, eval = run}
library(bartisan)

data("birthwt", package = "MASS")

d <- birthwt
d$race  <- factor(d$race,  levels = 1:3, labels = c("white", "black", "other"))
d$smoke <- factor(d$smoke, levels = 0:1, labels = c("no", "yes"))
d$ht    <- factor(d$ht,    levels = 0:1, labels = c("no", "yes"))
d$ui    <- factor(d$ui,    levels = 0:1, labels = c("no", "yes"))
d$low   <- NULL
```

## The measure

```{r vi, eval = run}
set.seed(2026)

fit <- bartisan(bwt ~ ., data = d, family = gaussian(), chains = 4)

variable_importance(fit)
```

Two columns, answering different questions.

`splits` is the average number of splitting rules the forest gives a predictor in
one posterior draw, with an interval. It says how much of the fitted structure
the predictor accounts for.

`prop_used` is the proportion of draws in which the predictor received at least
one rule. Read this one first. Under the default prior it behaves like a
posterior probability that the predictor belongs in the model, and it is the
column to use if the question is which variables to keep.

The reason it behaves that way is the prior. By default the variable a rule
splits on is drawn from a Dirichlet-distributed set of probabilities
[@linero2018sparse], which lets the forest concentrate on a few predictors and
drop the rest entirely. A predictor that contributes nothing can fall to
`prop_used` near zero, which is not possible under classic BART, where every
predictor keeps a fixed share of the splitting probability.

## Calibrating with noise

The table above does not separate anything: every predictor is used in at least
84% of draws. Before concluding that all eight matter, check what a variable that
certainly does not matter would look like on this data. Add some.

```{r noise, eval = run}
set.seed(11)
d_noise <- d
for (j in 1:3) d_noise[[paste0("noise", j)]] <- rnorm(nrow(d))

set.seed(2026)
fit_noise <- bartisan(bwt ~ ., data = d_noise, family = gaussian(), chains = 4)

variable_importance(fit_noise)
```

The three noise variables land in the middle of the table, above `ptl`, `ftv` and
`age`. Since they are random numbers with no relation to birth weight, the
ordering in the lower half of this table carries no information. The honest
conclusion is that this sample cannot distinguish the weaker predictors from
noise, not that they are all important.

This technique costs three lines and settles the question. Use it whenever an
importance table is going to be interpreted. A gap between the real predictors
and the noise controls is what makes the table readable; without one, only the
top of the table means anything.

## When it does work

The failure above is a property of this dataset, not of the method. With more
data and a stronger signal the separation is sharp. The following uses the
Friedman function, where `x1` to `x5` enter the outcome and `x6` to `x10` are
noise, at 500 observations.

```{r friedman}
friedman <- function(n) {
  x <- as.data.frame(matrix(runif(n * 10), n, 10))
  names(x) <- paste0("x", 1:10)
  x$y <- 10 * sin(pi * x$x1 * x$x2) + 20 * (x$x3 - 0.5)^2 +
    10 * x$x4 + 5 * x$x5 + rnorm(n)
  x
}

set.seed(7)
fit_fr <- bartisan(y ~ ., data = friedman(500), family = gaussian())

variable_importance(fit_fr)
```

The five real predictors sit at 1.00 and four of the five noise predictors below
0.13. The fifth, `x8`, comes in around 0.4, which is a useful reminder that even
a clean separation has a straggler: a noise variable will occasionally be picked
up, and a single moderate value is not evidence of anything. The gap that matters
runs from about 0.4 to 1.00, and it is wide.

There is no threshold that is correct in general. Look for the gap, cut inside
it, and check that the conclusion does not depend on where.

## Correlated predictors

This is the failure mode most likely to mislead, because it produces a confident
and wrong answer rather than an uninformative one.

```{r corr}
set.seed(5)
n <- 400
cd <- data.frame(x1 = runif(n))
cd$x1_copy <- cd$x1 + rnorm(n, sd = 0.01)   # almost the same variable
cd$x2 <- runif(n)
cd$x3 <- runif(n)
cd$y <- 3 * cd$x1 + rnorm(n, sd = 0.3)      # only x1 is in the truth

fit_corr <- bartisan(y ~ ., data = cd, family = gaussian())

variable_importance(fit_corr)
```

The outcome depends on `x1`. The forest used `x1_copy` in every draw and `x1` in
under 10% of them. Read as a selection result, this table says to keep the
variable that is not in the truth and drop the one that is.

Nothing has gone wrong with the fit. The two variables carry the same
information, so a tree that splits on either fits equally well, and the forest
settled on one arbitrarily. Predictions are unaffected. What is affected is any
statement about which variable matters.

The same thing happens to the effects:

```{r correff, eval = has_me}
library(marginaleffects)

avg_comparisons(fit_corr, variables = c("x1", "x1_copy"))
```

All of the association is attributed to the copy and none to the original.
Moving both together, which is what a change in the underlying quantity would
mean, recovers the truth:

```{r corrjoint}
lo <- transform(cd, x1 = 0.25, x1_copy = 0.25)
hi <- transform(cd, x1 = 0.75, x1_copy = 0.75)

drawn <- rowMeans(predict(fit_corr, newdata = hi, draws = TRUE) -
                    predict(fit_corr, newdata = lo, draws = TRUE))

round(c(estimate = mean(drawn), quantile(drawn, c(0.025, 0.975))), 3)
```

The true difference over that range is 1.5. So the model knows the relationship;
it just cannot say which of two identical columns it belongs to.

The practical rule is to treat a set of correlated predictors as one unit. Decide
in advance which variables measure the same underlying thing, and report and move
them together.

## Three things importance is not

**It is not an effect size.** How often a predictor is split on and how much it
moves the outcome are different quantities, and they can disagree in both
directions. A variable with a small effect over a range that gets split
repeatedly will show high usage. If the question is how much the outcome changes,
`avg_comparisons()` answers it directly and reports an interval.
See `vignette("effects")`.

**It is not a test.** `prop_used` is a posterior probability under a particular
prior. It has no error rate attached, and the choice of cutoff is yours. The
noise-control technique above is the closest thing to a calibration and it is
informal.

**It is not causal.** A ranking of predictors describes this fitted function on
this sample. It says nothing about what would happen if any of them were changed,
and a variable can rank highly because it is a consequence of the outcome, a
proxy for a confounder, or correlated with something that matters.
See `vignette("causal")`.

## A note on the number of trees

A recommendation that circulates is to reduce the number of trees when using BART
for variable selection. It comes from @chipman2010, who observe that counting
splits works poorly with many trees because "the redundancy offered by so many
trees tends to mix many irrelevant predictors in with the relevant ones", and
that predictors compete for splits when the forest is small.

The advice is correct for classic BART and largely unnecessary here. Measured on
the Friedman function at 500 observations, three replicates, the mean `prop_used`
for the five noise predictors was:

```{r treetable, echo = FALSE}
knitr::kable(data.frame(
  Trees = c(10, 20, 50, 100),
  `sparsity = FALSE` = c(0.28, 0.50, 0.95, 1.00),
  `sparsity = TRUE (default)` = c(0.09, 0.08, 0.09, 0.14),
  check.names = FALSE))
```

Without the sparsity prior the advice is essential: at 50 trees the noise
predictors are used in 95% of draws and cannot be told apart from the real ones.
With the prior, which is on by default, they stay near zero at every tree count.
The recommendation addresses the same problem the prior addresses, and there is
little left for it to do. Reducing the trees also costs predictive accuracy, so
it is not free.

If you want to check, fit at the default and again at 20 trees and see whether
the conclusion changes. It usually will not.

## Where to go next

`vignette("effects")` covers how much predictors move the outcome, which is the
question importance is usually a proxy for. `vignette("comparison")` covers
choosing between models that include different variables, which is the other way
to ask whether a variable earns its place.

## References
