---
title: "Survival analysis in bartisan"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Survival analysis in bartisan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 3.6, dpi = 110, fig.align = "center",
                      warning = FALSE, message = FALSE)
options(digits = 3)

has_gg <- rlang::is_installed("ggplot2")
has_surv <- rlang::is_installed("survival")

# Precomputed by _dev/survival-sim.R and friends, so that this vignette builds
# without refitting anything. See the section "A comparison across six truths".
res <- readRDS("survival-results.rds")

# One consistent ordering of the families and the truths across every table and
# figure. Needed whether or not ggplot2 is available, so it sits outside the
# guard below.
fam_levels <- res$meta$fam_levels
agg <- res$agg
agg$family <- factor(agg$family, levels = fam_levels)
agg$truth <- factor(agg$truth, levels = res$meta$truth_order)
agg <- agg[order(agg$truth, agg$family), ]

# One row per truth, one column per family, for the summary tables.
pivot <- function(value, families = fam_levels) {
  wide <- data.frame(truth = levels(agg$truth))
  for (f in families) {
    wide[[f]] <- agg[[value]][match(paste(wide$truth, f),
                                    paste(agg$truth, agg$family))]
  }
  names(wide)[1] <- "Truth"
  wide
}

if (has_gg) {
  library(ggplot2)
  theme_set(theme_bw(base_size = 9) +
              theme(panel.grid.minor = element_blank(),
                    strip.background = element_rect(fill = "grey93",
                                                    colour = NA),
                    legend.position = "bottom",
                    legend.title = element_blank()))
  fam_colours <- setNames(
    c("#B2182B", "#EF8A62", "#F4A582", "#4393C3", "#2166AC", "#1B7837"),
    fam_levels)
}
```

```{r libs}
library(bartisan)
library(survival)

set.seed(2026)

# Small chains in the live examples, so that this vignette builds quickly. The
# defaults are 50 trees and 500 draws after 500 warmup iterations, which is
# what the simulations reported below use.
ctrl <- bartisan_control(num_trees = 10, num_burn = 150, num_save = 150)
```

This vignette is about right-censored time-to-event data. It covers the five families that take a censored response, what each one is estimating, which shapes of hazard each can and cannot represent, how they behave under misspecification and under heavy censoring, and the route to take when the proportional hazards assumption fails. `vignette("families")` gives the short version; this is the long one.

The recurring theme is that these models disagree about *what the predictor means*, and that the disagreement mostly stops mattering once you ask for the quantity you actually wanted — survival at a horizon — rather than reading coefficients.

## The response

Every family here takes a `survival::Surv()` object, or equivalently a two-column numeric matrix of non-negative times and 0/1 event indicators. A 1 means the event was observed at that time; a 0 means the subject was still event-free when last seen, so the true time is somewhere above it.

```{r data}
n <- 400
d <- data.frame(x1 = runif(n), x2 = runif(n), x3 = runif(n),
                trt = rbinom(n, 1, 0.5))

# A log-normal accelerated failure time truth, censored at about 30%. The
# treatment multiplies survival time by exp(0.5).
m <- 1 + 1.5 * sin(3 * d$x1) - 0.8 * d$x2 + 0.5 * d$trt
latent <- exp(m + 0.7 * rnorm(n))
cens <- rexp(n, 0.05)

d$time <- pmin(latent, cens)
d$status <- as.numeric(latent <= cens)

mean(d$status)
```

Only right censoring is supported: `Surv(time, status)` and nothing else. Left-truncated, interval-censored and competing-risks data need a different likelihood, and `custom_family()` is the place to write one.

Censoring is assumed **independent** of the event time given the covariates, as it is in every model in this vignette and in `survival::coxph()`. That is an assumption about the data, not about the family, so no choice below relaxes it.

## The five families

```{r famtable, echo = FALSE}
knitr::kable(data.frame(
  Family = c("`weibull_aft()`", "`loglogistic_aft()`", "`lognormal_aft()`",
             "`ph()`", "`dpm_aft()`"),
  Model = c("$\\log T = \\eta + \\sigma\\epsilon$, $\\epsilon$ standard Gumbel",
            "the same with $\\epsilon$ standard logistic",
            "the same with $\\epsilon$ standard normal",
            "$\\lambda(t \\mid x) = \\lambda_0(t)\\,e^{r(x)}$",
            "$\\log T = m(x) + W$, $W$ a mixture of normals"),
  `The forest gives` = c("a log time ratio", "a log time ratio",
                         "a log time ratio", "a log hazard ratio",
                         "$E[\\log T \\mid x]$"),
  `Drawn nuisance` = c("$\\sigma$", "$\\sigma$", "$\\sigma$",
                       "$\\lambda_0$ on a grid", "the error density"),
  check.names = FALSE))
```

All five put a BART forest on a single additive predictor and leave something else free. What differs is which part is parametric.

### The three accelerated failure time families

These model the log event time directly,

$$\log T = \eta(x) + \sigma\epsilon,$$

with $\epsilon$ of a fixed shape and $\sigma$ drawn and reported in `fit$aux`. Multiplying survival time by a constant adds a constant to $\eta$, which is what "accelerated failure time" means: covariates stretch or compress the time axis without changing the shape of the survival curve.

```{r aftfits}
fit_w <- bartisan(Surv(time, status) ~ x1 + x2 + x3 + trt, data = d,
                  family = weibull_aft(), control = ctrl)
fit_ln <- bartisan(Surv(time, status) ~ x1 + x2 + x3 + trt, data = d,
                   family = lognormal_aft(), control = ctrl)

c(weibull = mean(fit_w$aux[, "sigma"]), lognormal = mean(fit_ln$aux[, "sigma"]))
```

They differ only in the error's shape, and therefore in the shape of the hazard:

- **`weibull_aft()`** has a **monotone** hazard — increasing, decreasing or flat, but never turning around. It is the only family in the package that is simultaneously an accelerated failure time and a proportional hazards model, so it is the one parametric choice that supports a hazard-ratio reading. It is about seven times slower than the other two — see [Speed](#speed).
- **`loglogistic_aft()`** allows a hazard that rises and then falls, and has the heaviest tails of the three, so a handful of very long survivors move it least.
- **`lognormal_aft()`** also allows a non-monotone hazard, with thinner tails than the log-logistic.

In practice the log-logistic and log-normal are hard to tell apart from data and the Weibull is distinguishable from both, so the informative comparison is Weibull against either of the other two, by `loo()`.

### `ph()`: a free baseline hazard

`ph()` gives up on a parametric time distribution and models the hazard instead [@basak2024]:

$$\lambda(t \mid x) = \lambda_0(t)\exp\{r(x)\},$$

with $\lambda_0$ constant within each of `num_bins` pieces whose edges sit at evenly spaced quantiles of the observed times. This is the piecewise-exponential proportional hazards model with a forest on the log hazard ratio. The baseline can take any shape at all, including one that turns over, which none of the three parametric families can do.

```{r phfit}
fit_ph <- bartisan(Surv(time, status) ~ x1 + x2 + x3 + trt, data = d,
                   family = ph(), control = ctrl)

head(round(colMeans(fit_ph$aux), 3))
```

The bin hazards come back as `lambda1`, `lambda2`, … alongside `lambda_rate`, the drawn rate of their own Gamma prior — the hazards are shrunk towards each other through it, which is what keeps a fine grid from overfitting.

One point of interpretation matters here. The predictor and the baseline are identified only *jointly*: multiplying $\lambda_0$ by a constant and subtracting its log from $r(x)$ leaves the likelihood unchanged. The fit resolves this by letting the baseline carry the level and reporting $r(x)$ centered, so **`fit_ph`'s predictor is a contrast, not a level**. Differences between two values of $r$ are meaningful; a single value is not.

### `dpm_aft()`: the error distribution estimated rather than assumed

`dpm_aft()` is an accelerated failure time model with the shape of the error left free [@henderson2020], and is the family used when a `Surv` response is given and no `family` is named:

$$\log T = m(x) + W, \qquad W \sim \text{a Dirichlet process mixture of normals},$$

with the mixture constrained to mean zero, so $m(x)$ is the conditional mean of $\log T$. It is `dpm()`'s error model with censoring added: each sweep, a right-censored log-time is imputed from the mixture component it currently sits in, truncated below at its censoring time.

```{r dpmfit}
fit_dpm <- bartisan(Surv(time, status) ~ x1 + x2 + x3 + trt, data = d,
                    family = dpm_aft(), control = ctrl)

round(colMeans(fit_dpm$aux), 3)
```

`error_density()` reports what the error distribution came out looking like, which is the diagnostic the model exists to provide:

```{r errdens, fig.height = 2.6, eval = has_gg}
ed <- error_density(fit_dpm)

# The normal a lognormal_aft() fit would have assumed, given the same spread.
sd_fitted <- mean(fit_dpm$aux[, "error_sd"])

ggplot(ed, aes(at, mean)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey85") +
  geom_line(linewidth = 0.5) +
  stat_function(fun = dnorm, args = list(sd = sd_fitted),
                linetype = 2, colour = "#B2182B") +
  labs(x = "error in log T", y = "density",
       subtitle = "fitted error density (solid) against the normal of the same spread (dashed)")
```

Because the mixture can collapse to a single component, it costs almost nothing when a single normal was the right answer — the property that makes it a reasonable default rather than a specialist tool. Nor does it cost much time: conditional on which component each observation sits in, the model is Gaussian, so it takes the same fast path `lognormal_aft()` does and the mixture update on top is cheap. It runs at about twice `lognormal_aft()`'s cost and a third of `weibull_aft()`'s.

## What the predictor means: three estimands, not one

The five families report three genuinely different quantities, and they are not rescalings of each other.

```{r estimandtable, echo = FALSE}
knitr::kable(data.frame(
  Family = c("the three `*_aft()`", "`ph()`", "`dpm_aft()`"),
  `A one-unit increase in the predictor means` = c(
    "survival time multiplied by $e$, at every quantile of the curve",
    "the hazard multiplied by $e$, at every time",
    "the mean of $\\log T$ raised by one"),
  check.names = FALSE))
```

A time ratio and a hazard ratio move in opposite directions — a covariate that raises the hazard shortens survival — and their magnitudes are related only through the shape of the baseline. Under a Weibull with shape $k$ the two coincide up to a factor: a log hazard ratio of $\beta$ is a log time ratio of $-\beta/k$. Away from the Weibull there is no such correspondence at all.

`dpm_aft()`'s predictor is a third thing. It is the mean of $\log T$, not the median and not a quantile, so it is only a time ratio to the extent that the error density is symmetric — which is exactly what `dpm_aft()` declines to assume.

### The estimand you usually want

None of the three is normally the question. The question is usually about survival at a horizon: the difference in one-year survival between treated and untreated, say. That is a contrast in $S(t \mid x)$ at a fixed $t$, and it is on the same scale — a probability — no matter which family produced it.

```{r survtype}
head(predict(fit_ph, type = "survival", times = c(1, 2, 5)), 3)
```

and inside the estimand machinery:

```{r survest, eval = rlang::is_installed("marginaleffects")}
library(marginaleffects)

# The average difference in survival at t = 5 that the treatment is worth.
avg_comparisons(fit_ph, variables = "trt", type = "survival", times = 5)
```

One time per call. **marginaleffects** warns that it does not recognize `times`, because it checks the dots against a whitelist hardcoded per model class and offers no hook for registering an argument; it passes the argument through regardless and the result is correct. The warning is suppressed here.

This works identically for all five families. It is the reason `type = "survival"` exists rather than leaving the curve to be assembled by hand from the draws, and it is the recommended way to report a survival model from this package: it sidesteps the estimand mismatch above, it comes with a posterior interval, and it is comparable across families in a way that the predictors are not.

`type = "response"` is the median survival time for all five, which is the other scale-free summary and is often easier to communicate than a curve.

## Which hazard shapes each family can represent

Most of the differences in what follows come down to one picture. The parametric families constrain the hazard's shape; `ph()` does not.

```{r hazshapes, echo = FALSE, fig.height = 3.1, eval = has_gg}
tt <- seq(0.05, 3, length.out = 200)

haz <- rbind(
  data.frame(family = "weibull_aft()", label = "shape 0.7",
             t = tt, h = 0.7 * tt^(0.7 - 1)),
  data.frame(family = "weibull_aft()", label = "shape 1.8",
             t = tt, h = 1.8 * tt^(1.8 - 1)),
  data.frame(family = "loglogistic_aft()", label = "shape 2.5",
             t = tt, h = (2.5 * tt^(2.5 - 1)) / (1 + tt^2.5)),
  data.frame(family = "lognormal_aft()", label = "sigma 0.6",
             t = tt, h = dnorm(log(tt) / 0.6) /
               (0.6 * tt * pnorm(-log(tt) / 0.6))),
  data.frame(family = "ph()", label = "free baseline",
             t = tt, h = approx(c(0, 0.4, 0.8, 1.3, 2, 3),
                                c(0.3, 1.5, 2.2, 1.1, 0.5, 0.8),
                                xout = tt, method = "constant",
                                rule = 2)$y))
haz$family <- factor(haz$family, levels = fam_levels)

ggplot(haz, aes(t, h, group = interaction(family, label))) +
  geom_line(aes(colour = family), linewidth = 0.6) +
  facet_wrap(~ family, nrow = 1) +
  scale_colour_manual(values = fam_colours, guide = "none") +
  coord_cartesian(ylim = c(0, 2.6)) +
  labs(x = "time", y = "hazard",
       subtitle = "the Weibull hazard cannot turn around; the other two can turn over exactly once; ph()'s can do anything")
```

`dpm_aft()` is not in the figure because its hazard shape is whatever the fitted error mixture implies, which can be multi-modal — a hazard with more than one peak, which none of the other four can produce.

## A comparison across six truths

`r sprintf("The simulation behind this section fits each family to %d training observations and scores it on %d held-out observations, with %d trees and %d draws after %d warmup, over %d replicates and at about 30%% censoring.", res$meta$n_train, res$meta$n_test, res$meta$num_trees, res$meta$num_save, res$meta$num_burn, res$meta$n_rep)`
The script is `_dev/survival-sim.R` in the package sources.

Every truth uses the same nonlinear function of five covariates, so the families differ only in how well they cope with the *shape* of the time distribution built around it. The six truths are:

```{r truthtable, echo = FALSE}
knitr::kable(data.frame(
  Truth = c("Weibull PH", "hazard turns over", "log-normal errors",
            "bimodal errors", "heavy-tailed errors", "crossing hazards"),
  `What it is` = c(
    "proportional hazards with a Weibull baseline",
    "proportional hazards, baseline hazard rises then falls",
    "accelerated failure time, normal errors",
    "accelerated failure time, two-component error",
    "accelerated failure time, $t_3$ errors",
    "the covariate effect reverses sign over time"),
  `Exactly right for` = c(
    "`weibull_aft()` and `ph()` both", "`ph()`", "`lognormal_aft()`",
    "`dpm_aft()`",
    "none; `dpm_aft()` and `loglogistic_aft()` nearest",
    "none of the five, nor the discrete-time route exactly"),
  check.names = FALSE))
```

The primary metric is the RMSE of the fitted survival function $S(t \mid x)$ against the true one, on held-out subjects and over twenty times spanning the 5th to 95th percentile of the event-time distribution. It is used because it is on the same scale for every family, which the predictors are not.

```{r rmsefig, echo = FALSE, fig.height = 4.4, eval = has_gg}
ggplot(agg, aes(s_rmse, family, colour = family)) +
  geom_point(size = 1.9) +
  geom_linerange(aes(xmin = s_rmse - s_rmse_se, xmax = s_rmse + s_rmse_se),
                 linewidth = 0.5) +
  facet_wrap(~ truth, ncol = 2) +
  scale_colour_manual(values = fam_colours, guide = "none") +
  scale_y_discrete(limits = rev(fam_levels)) +
  expand_limits(x = 0) +
  labs(x = expression(RMSE~of~S(t*"|"*x)~~~"(lower is better)"), y = NULL,
       subtitle = "held-out error in the survival function, one standard error over replicates")
```

```{r rmsetable, echo = FALSE}
knitr::kable(pivot("s_rmse"), digits = 3, row.names = FALSE,
             caption = "RMSE of S(t | x) on held-out data. Lower is better.")
```

Five things to read off it.

**Being exactly right is worth surprisingly little.** On the Weibull truth, which `weibull_aft()` and `ph()` both describe exactly, `weibull_aft()` leads at 0.045 — and `ph()` and `dpm_aft()`, neither of which is given the shape, are level with it at 0.046. Being *wrong* costs far more than being right gains: `lognormal_aft()` pays 0.068 on the same data. So the question worth asking of a family is not how much it wins when its assumption holds, but how much it loses when it does not.

**`ph()` wins where the baseline turns over, and only there.** At 0.045 against 0.055 for the best parametric alternative, it is clearly ahead on the case the three parametric families structurally cannot fit. But this is the one truth in the table where it leads. On the two accelerated failure time truths it is among the worst — 0.069 on log-normal errors against 0.049 — because proportional hazards is an assumption too, and it is the wrong one there.

**When the error is badly shaped, `dpm_aft()` wins by a wide margin.** On the two-component error it is three times more accurate than anything else and 300 log points ahead, because no other family in the set can represent a bimodal time distribution at any value of its parameters. The heavy-tailed case is milder, and there `loglogistic_aft()` matches it — which is what the log-logistic's heavy tails are for.

**`dpm_aft()` costs nothing when it is not needed, and it is the most consistent family in the table.** On log-normal errors it sits at 0.0491 against the correctly specified `lognormal_aft()`'s 0.0498; on the Weibull truth, 0.0457 against `weibull_aft()`'s 0.0449. It is best or tied-best on four of the six truths and never worse than third on any. That asymmetry — large gains when the assumption is wrong, no measurable loss when it is right — is what makes it a sensible default for anyone without a view about the error's shape.

**When hazards cross, the RMSE column understates the damage.** Every family lands between 0.080 and 0.096, which looks like a mild penalty. It is not: the [ranking column](#ordering) shows that the models have absorbed a reversing covariate effect into almost no covariate effect at all. See [the discrete-time route](#nonprop) below.

### The same picture as fitted curves

The aggregate numbers hide *how* the misspecified fits are wrong. Here are the fitted survival curves for three held-out subjects under the two truths where the families disagree most, against the truth in black.

```{r curvefig, echo = FALSE, fig.height = 4.8, eval = has_gg}
keep <- c("truth", "weibull_aft()", "dpm_aft()", "ph()",
          "discrete-time probit")
cv <- res$curves[res$curves$truth %in% c("hazard turns over",
                                         "crossing hazards") &
                   res$curves$family %in% keep, ]
# Truths as rows in the intended order, families as columns, and the three
# subjects as lines *within* each panel -- so that a model which collapses the
# three onto one curve is visible as such.
cv$truth <- factor(cv$truth, levels = c("hazard turns over",
                                        "crossing hazards"))
cv$family <- factor(cv$family, levels = keep,
                    labels = c("the truth", "weibull_aft()", "dpm_aft()",
                               "ph()", "discrete time"))
cv$subject <- factor(cv$subject, levels = c("poor prognosis", "typical",
                                            "good prognosis"))

ggplot(cv, aes(time, surv, colour = subject)) +
  geom_line(linewidth = 0.55) +
  facet_grid(truth ~ family) +
  scale_colour_manual(values = c("poor prognosis" = "#B2182B",
                                 "typical" = "#999999",
                                 "good prognosis" = "#2166AC")) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "time", y = expression(S(t*"|"*x)),
       subtitle = "three held-out subjects: the truth in the first column, four fits beside it")
```

In the top row every fit recovers the separation between the three subjects; they differ in the level. `weibull_aft()` sends the good-prognosis curve to zero well before the truth does, because a monotone hazard cannot flatten out the way this baseline does; `ph()` and `dpm_aft()` track it, and the discrete-time fit is visibly steppy at this grid resolution.

The bottom row is the one to look at. The true curves separate widely and then **reverse** — the subject with the best prognosis at the middle of the grid has the worst survival by the end of it, and the three curves cross. No fit reproduces that, which is the structural limitation stated as a picture. But the symptom is not a distorted covariate effect; it is a missing one. `ph()` and the discrete-time fit put all three subjects on essentially the same curve. `weibull_aft()` and `dpm_aft()` keep a little separation, in one fixed order, which is right over part of the range and wrong over the rest. A model that must apply the same covariate effect at every time, given an effect that is positive early and negative late, averages it to approximately nothing.

Two cautions on reading this panel. It shows three subjects, not the sample: the discrete-time route scores much better than `ph()` on the ranking measured over all 700, and that advantage is not what this figure is showing. And the survival RMSE for this row was an unremarkable 0.08 to 0.10 — the failure is severe and the headline metric does not say so.

### Recovering the ordering of risk {#ordering}

A weaker question than the curve itself is whether the model gets the *ordering* of subjects right at each time. Misspecification costs much less here.

```{r rankfig, echo = FALSE, fig.height = 3.4, eval = has_gg}
ggplot(agg, aes(rank, family, colour = family)) +
  geom_point(size = 1.9) +
  facet_wrap(~ truth, ncol = 3) +
  scale_colour_manual(values = fam_colours, guide = "none") +
  scale_y_discrete(limits = rev(fam_levels)) +
  labs(x = "mean Spearman correlation with the true S(t | x)  (higher is better)",
       y = NULL,
       subtitle = "how well the ordering of subjects by survival is recovered")
```

On all five truths where the covariate effect is constant in time, every family recovers the ordering at a correlation between 0.92 and 0.99, whether or not it has the shape right. The spread across families is a few hundredths where the spread in the survival curve was a factor of three.

The crossing truth is the exception, and there the column collapses: the discrete-time model reaches 0.47, the accelerated failure time families 0.24 to 0.29, and `ph()` and `weibull_aft()` essentially zero — 0.04 and $-0.03$, which is to say no information about who is at risk at a given time. This is the same erasure the curves showed, measured.

The practical reading: **if all you need is who is at higher risk, and the covariate effect does not move with time, the choice of family barely matters.** It matters when you need the curve, an absolute probability, or the shape of the hazard — and it matters enormously when the effect does move with time, which is the one case where the ordering is not recoverable at all.

### The error density, recovered

On the bimodal truth `dpm_aft()`'s advantage is not a tuning gain; it is that it can see the shape the other families are assuming away.

```{r densfig, echo = FALSE, fig.height = 2.9, eval = has_gg}
dn <- res$dens[res$dens$truth %in% c("bimodal errors", "log-normal errors"), ]
truthdens <- do.call(rbind, lapply(split(dn, dn$truth), function(z) {
  at <- z$at
  h <- if (z$truth[1] == "bimodal errors")
    0.5 * dnorm(at, 1.3, 0.35) + 0.5 * dnorm(at, -1.3, 0.35)
  else dnorm(at, 0, 0.7)
  data.frame(truth = z$truth[1], at = at, mean = h)
}))

ggplot(dn, aes(at, mean)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#4393C3", alpha = 0.25) +
  geom_line(colour = "#2166AC", linewidth = 0.6) +
  geom_line(data = truthdens, colour = "black", linetype = 2, linewidth = 0.5) +
  facet_wrap(~ truth, scales = "free") +
  labs(x = "error in log T", y = "density",
       subtitle = "dpm_aft()'s fitted error density (blue, with interval) against the truth (dashed)")
```

It finds both components when they are there and collapses to something close to a single normal when they are not. That is the whole argument for the family.

### Held-out log scores

The log score is the other natural metric, and it can be made comparable across all five families — but only after a correction, which is the subject of [a trap below](#measure). With the correction applied:

```{r lstable, echo = FALSE}
knitr::kable(pivot("logscore", setdiff(fam_levels, "discrete-time probit")),
             digits = 0, row.names = FALSE,
             caption = "Held-out log predictive score on the density of T, summed over the 700 held-out observations. Higher is better.")
```

The two metrics agree on which family to prefer on five of the six truths. The exception is the turning-over baseline, where `ph()` leads on the survival curve but sits third on the log score — behind `dpm_aft()` and `loglogistic_aft()`, and by three log points, which is inside the noise. The reason is that the log score is evaluated at each subject's own observed time, which is by construction where the data are dense, while the survival RMSE is averaged over a grid that reaches into the tail. Families differ less in the middle than at the edges.

Where the shape of the density is genuinely wrong the log score is the harsher of the two: on the bimodal truth `dpm_aft()` is ahead by more than 300 log points, a much larger gap than the threefold one in the RMSE table.

That the correction works at all is worth one check. On the Weibull truth, where `weibull_aft()` and `ph()` fit the same model by different parameterizations, they score $-191$ and $-200$. Without the Jacobian they would be several hundred apart, which is the size of $\sum \delta_i \log t_i$ on this data and has nothing to do with either fit.

## Censoring

More censoring means less information, so every family gets worse. The question is whether it changes which family to prefer.

```{r censfig, echo = FALSE, fig.height = 3.2, eval = has_gg}
sw <- res$sweep_agg
sw$family <- factor(sw$family, levels = fam_levels)

ggplot(sw, aes(censor, s_rmse, colour = family)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.6) +
  scale_colour_manual(values = fam_colours) +
  scale_x_continuous(labels = function(z) paste0(100 * z, "%")) +
  labs(x = "fraction censored", y = expression(RMSE~of~S(t*"|"*x)),
       subtitle = "turning-over baseline, censoring swept from none to 70% (note the y axis does not start at zero)")
```

**`ph()` is best at every level**, from no censoring to 70%, and no parametric family overtakes it at any point on this truth. So the recommendation is stable in the amount of censoring, which is the question that mattered.

The degradation is gentle up to about half the observations censored — `ph()` goes from 0.044 to 0.055 — which is the usual reassurance about likelihood-based survival analysis: a censored observation is not a missing one, it contributes $S(t)$ and that is real information. Beyond half, the curves steepen.

**One line does not behave**, and it is worth understanding rather than passing over. `weibull_aft()` is the *worst* family in the panel with no censoring at all, at 0.097, and it gets **better** as censoring increases, reaching 0.072 at 70%. Everything else degrades monotonically. The likely reason is that this truth has a heavy, polynomial tail, which a Gumbel error in log time cannot represent; with no censoring the fit is dragged by observed times far out in that tail, and censoring truncates exactly the observations it cannot accommodate. Censoring is protecting a misspecified model from the part of the distribution it gets wrong. That is a caution about reading a fit's apparent quality off a heavily censored sample, not a reason to want censoring.

One expectation the sweep did not bear out: `dpm_aft()` might have been expected to suffer most under heavy censoring, since it estimates a whole density from log-times that are increasingly imputed rather than observed. It does not. Over the sweep its error grows by a factor of 1.4, against 1.5 for both `ph()` and `loglogistic_aft()`. Its advantage is not fragile in the way the extra machinery suggests.

## When hazards are not proportional {#nonprop}

Every family above assumes that covariates act the same way at every time — the accelerated failure time families by stretching the time axis, `ph()` by scaling the hazard. When that fails, and in particular when survival curves cross, none of them can fit the data, and the failure is structural rather than a matter of flexibility.

The model to reach for is a **discrete-time hazard**, and it needs no new family. Expand the data to one row per subject per time point up to their own, then fit `binomial("probit")` with **time as a covariate**. This is the nonparametric survival BART of @sparapani2016; because the hazard becomes an arbitrary function of $(t, x)$, nothing is assumed about proportionality at all.

```{r dtcode, eval = has_surv}
# The grid: quantiles of the observed event times.
edges <- quantile(d$time[d$status == 1], seq(0.05, 0.95, length.out = 15))

expand_dt <- function(d, edges, xnames) {
  reached <- pmin(pmax(findInterval(d$time, edges, rightmost.closed = TRUE), 1L),
                  length(edges))
  long <- d[rep(seq_len(nrow(d)), reached), xnames, drop = FALSE]
  long$tbin <- unlist(lapply(reached, seq_len))
  long$ev <- 0
  long$ev[cumsum(reached)] <- d$status    # the event lands on the last row
  long
}

long <- expand_dt(d, edges, c("x1", "x2", "x3", "trt"))
c(subjects = nrow(d), rows = nrow(long))

fit_dt <- bartisan(ev ~ ., data = long, family = binomial("probit"),
                   control = ctrl)
```

The survival function is the running product of one minus the fitted hazards:

```{r dtsurv, eval = has_surv}
grid_dat <- d[rep(1:3, each = length(edges)), c("x1", "x2", "x3", "trt")]
grid_dat$tbin <- rep(seq_along(edges), 3)

h <- matrix(predict(fit_dt, newdata = grid_dat, type = "response"),
            nrow = length(edges))
round(t(apply(1 - h, 2, cumprod))[, c(1, 8, 15)], 3)
```

`binomial("probit")` is the fastest family in the package, so the cost is the expansion rather than the sampler: the number of rows is the sum of how many grid points each subject reaches. Fifteen to twenty grid points is usually plenty; a finer grid multiplies the rows without adding much.

**How it compares.** From the same simulation, on the two truths that separate it from `ph()`:

```{r dtcompare, echo = FALSE}
sel <- agg[agg$truth %in% c("hazard turns over", "crossing hazards") &
             agg$family %in% c("ph()", "weibull_aft()", "dpm_aft()",
                               "discrete-time probit"),
           c("truth", "family", "s_rmse", "s_worst", "rank")]
sel <- sel[order(sel$truth, sel$s_rmse), ]
names(sel) <- c("Truth", "Family", "S RMSE", "worst t", "ranking")
knitr::kable(sel, digits = 3, row.names = FALSE)
```

The trade is sharper than one might hope, in both directions.

**Under proportional hazards it is the worst model in the whole comparison**, not a close second — 0.073 against `ph()`'s 0.045 on the turning-over baseline, and 0.076 against 0.046 on the Weibull. It nests proportional hazards, so this is not bias; it is the cost of making the forest learn the baseline through splits on time while it is also learning the covariate effect, where `ph()` draws the baseline in closed form and spends the whole forest on $x$. Freedom is not free, and here it is expensive.

**Under crossing hazards it does not win on the survival curve either** — 0.085, behind the accelerated failure time families at 0.080. What it wins is the thing that matters: the ranking, at 0.47 against 0.29 and below for everything else. It is the only model in the set that recovers *any* of the reordering, and recovering some of a reversing effect while getting the level slightly worse is the better failure of the two.

Three further costs. The expansion inflates the data, so a large study with a fine grid gets slow. The grid is a real choice, unlike `ph()`'s bins, because it sets the resolution of the hazard in $t$ *and* the resolution at which non-proportionality can be detected at all. And there is no `sigma`, no baseline and no error density to report — the model is a hazard surface, summarized through `type = "response"` and the curve, and `type = "survival"` does not apply to it because as far as the package is concerned it is a binomial fit.

**When to use it.** When you have a substantive reason to expect the covariate effect to move with time — a treatment whose benefit accrues or wears off, a risk factor that matters only early. Not as a hedge: under proportional hazards it is the most expensive option in the comparison, so reaching for it "just in case" costs about as much as being wrong about the error distribution. If the question is whether hazards are proportional, fitting it alongside `ph()` and comparing the fitted curves is a reasonable diagnostic; adopting it by default is not.

## Three traps

### The log score does not use the same measure for every family {#measure}

`predict(type = "density")` returns, for each observation, its contribution to the likelihood — a density for an event and $S(t)$ for a censored observation. **The density is not on the same scale for every family.** The accelerated failure time families, `dpm_aft()` included, report the density of $\log T$; `ph()` reports the density of $T$. Censored contributions are survival probabilities and carry no measure at all.

So the two differ by a Jacobian, on events only:

```{r jacobian, eval = has_surv}
# Put an accelerated failure time family's log score on the density of T,
# so that it is comparable with ph()'s.
log_score_T <- function(fit, newdata) {
  ld <- sum(predict(fit, newdata = newdata, type = "density", log = TRUE))
  if (identical(fit$family$family, "ph")) ld
  else ld - sum(newdata$status * log(newdata$time))
}

c(lognormal = log_score_T(fit_ln, d), ph = log_score_T(fit_ph, d))
```

Without the correction the comparison is meaningless: on one of the simulations above the uncorrected gap between `ph()` and the accelerated failure time families exceeded a thousand log points and reversed which family looked better. `loo()` and `waic()` inherit the same issue, since they are built on the same pointwise densities — they are valid for comparing two accelerated failure time families, or two `ph()` fits, and invalid across the two groups.

If you would rather not think about it, compare on $S(t \mid x)$ from `predict(type = "survival")`, which is a probability for every family and needs no correction.

### `ph()`'s predictor is a contrast, not a level

Because the baseline absorbs the level, a single fitted value of `ph()`'s predictor has no interpretation on its own; only differences do. `predict(type = "link")` is centered accordingly. If you want a level, ask for a quantity that has one: `type = "survival"`, or `type = "response"` for the median time.

### `num_bins` is not a tuning parameter

`ph()` exposes `num_bins`, and it is fair to ask whether you are being made to make a choice. Sweeping it over a sixty-fold range on the same two proportional-hazards truths, at 700 observations where the default is 9:

```{r binsweep, echo = FALSE}
knitr::kable(res$bins, digits = 3, row.names = FALSE,
             caption = "Each cell is the turning-over baseline first, the Weibull baseline second.")
```

**From 4 bins to 100 the estimates are flat** — a twenty-five-fold range, no trend in either error column, and every difference inside the replicate-to-replicate spread, which is about 0.009 and 0.005 respectively. The default sits near the bottom of that plateau, and moving anywhere inside it changes nothing you would notice.

At 250 bins something does happen, consistently in both truths and both error columns: the error rises by about 20% and the Pareto-$k$ diagnostics start to be troubled. This is over-parameterization becoming visible — 250 bins on 700 observations gives about 200 effective parameters, and the shrinkage through `lambda_rate` is no longer enough to absorb it. Push further, to one piece per event time, and each observation's density is inflated by a parameter only that observation informs: measured that way the effective parameter count exceeded the sample size and more than half the Pareto-$k$ values went above 0.7, at which point `loo()` and `waic()` stop being usable at all.

So the shape of the thing is a wide flat plateau with a cliff a long way past the default, not a peak you have to find. `num_bins` is exposed for confirming that on your own data rather than for tuning: fit at the default and again at two or three times it, and if the answers agree — they will — stop thinking about it.

This is also the answer to a question that comes up: why not fit Cox's *partial* likelihood, which needs no grid at all? Because it couples observations through risk sets, so it does not decompose into a sum over the observations reaching a leaf, which is what this sampler requires. `ph()` fits the full likelihood of the piecewise-exponential model instead, which does decompose, and which approaches the partial likelihood as the bins shrink. A grid-free version was built and measured against this one [@linero2022]; it was no more accurate, and it broke `loo()`, for the reason above.

## Speed {#speed}

```{r timefig, echo = FALSE, fig.height = 2.6, eval = has_gg}
tm <- res$timing
tm$family <- factor(tm$family, levels = fam_levels)

ggplot(tm, aes(secs, family, fill = family)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = fam_colours, guide = "none") +
  scale_y_discrete(limits = rev(fam_levels)) +
  labs(x = "seconds per fit", y = NULL,
       subtitle = "700 observations, 50 trees, 500 draws after 500 warmup")
```

The spread is about eightfold, and it does not follow the families' flexibility at all.

`lognormal_aft()` and `loglogistic_aft()` are the fastest, at about 1.2 seconds, because both reach the sampler's quadratic fast path through data augmentation — the augmented target is exactly quadratic in the predictor, so the leaf draw is a closed form rather than a Laplace approximation. `dpm_aft()` comes next at about 2.8 seconds, because conditional on which mixture component each observation currently sits in it is *also* exactly Gaussian, and so takes the same fast path; the Dirichlet process update on top of that is cheap. `ph()` and the discrete-time route sit at about six seconds. The slowest family is `weibull_aft()`, at about 8.4 seconds, because its likelihood takes the exponential path rather than the quadratic one. None of these is slow enough at this sample size to drive the choice. Fit the one that matches the question; speed is a reason to prefer `lognormal_aft()` only when you are fitting many models rather than one.

## Choosing

```{r choosetable, echo = FALSE}
knitr::kable(data.frame(
  `If` = c("you have no view about the shape of anything",
           "you want the same, but the fit has to be quick",
           "the shape of the baseline hazard is the point",
           "you want a hazard ratio from a parametric fit",
           "you only need to know who is at higher risk",
           "the covariate effect may move with time",
           "you want to report a treatment effect"),
  Use = c("`dpm_aft()`",
          "`lognormal_aft()` or `loglogistic_aft()`",
          "`ph()`",
          "`weibull_aft()`",
          "any of them, if the effect is constant in time",
          "the discrete-time route, not a family",
          "any of them, through `type = \"survival\"`"),
  check.names = FALSE))
```

`dpm_aft()` heads the table, and is the default for an unnamed `Surv` response, on the evidence above rather than on principle: it was best or tied-best on four of the six truths and never worse than third, and on the two truths where a parametric family was exactly right it matched that family to the third decimal. It is also, contrary to what its flexibility suggests, one of the cheaper families to fit — about twice `lognormal_aft()` and a third of `weibull_aft()`. The second row is for fitting many models rather than one.

Two closing points that the numbers above support and that are easy to lose sight of.

The families disagree far more about the *density* than about the *ordering*. If your question is comparative — who is at higher risk, which arm does better — you have a lot of freedom in the choice and should spend your effort elsewhere. If your question is an absolute probability at a horizon, the choice matters and the comparison above is worth taking seriously.

And the failure that costs most is non-proportionality, not the shape of the error. Getting the error wrong cost a factor of three at worst, and `dpm_aft()` insures against it for nothing. A covariate effect that reverses over time cost every family in the table the entire covariate signal — an effect estimated at approximately zero, with a survival RMSE that looked unremarkable while it happened. `ph()` and `dpm_aft()` each relax one thing while holding proportionality or acceleration fixed; the discrete-time route is the only option here that relaxes *that*, and it is worth its cost exactly when the substance suggests an effect that moves with time.

## References
