---
title: "Varying Coefficients and Random Intercepts"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Varying Coefficients and Random Intercepts}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{=html}
<style>
pre {overflow-x: auto;}
pre code {word-wrap: normal; white-space: pre;}
</style>
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.2, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 4)

has_loo <- rlang::is_installed("loo")
has_nlme <- rlang::is_installed("nlme")
run <- TRUE
```

## Introduction

A BART model is a sum of trees, and a sum of trees has no coefficients. It predicts, and an effect is read off it as a contrast between two predictions, which is what `estimate_effect()` and the *marginaleffects* functions in `vignette("effects")` do. That is the right reading for most predictors. For some, though, we want the effect to *be* a parameter: something with a prior of its own, a posterior of its own, and a value we can report as a coefficient rather than reconstruct as a difference. The treatment in a causal analysis is the usual case, and so is any predictor whose effect is the whole point of the model.

`vc()` is how a predictor is given a coefficient in *bartisan*, and the coefficient it gets is a function rather than a number: a forest of its own that says how the effect moves with the other predictors. This is the varying-coefficient model of @deshpande2026, of which the Bayesian causal forest of @hahn2020 is the case of one binary covariate and @woody2020 the case of one continuous one. `bcf()` fits that special case with the settings the causal problem wants, and this guide is where the general model is described.

In this guide we will write the model down and say what each of its parts is, then declare a varying coefficient with `vc()` and work through its three arguments: which predictors the coefficient varies with, where the control function is read, and what happens when the coefficient is not allowed to vary at all, which is how a linear term is written into an otherwise nonparametric model. Next we'll read the fit, both as coefficients and as effects, and see how the settings of a fit are keyed by forest once there is more than one. Then we'll show how `bcf()` writes its model in these terms. Finally, because they are declared in the same formula and explained nowhere else, we'll cover random intercepts.

Throughout we use the `rhc` dataset (see `vignette("bartisan")` or `help("rhc", package = "bartisan")`), predicting `death` from the covariates and from `rhc`, whether the patient received right heart catheterization. The fits use the default single chain; `vignette("diagnostics")` covers checking convergence, which we skip here for brevity.

```{r libs, eval = run}
library(bartisan)

data("rhc")
```

## The Model

A model with varying coefficients is

$$g(\mu_i) = f_0(Z_i) + \sum_j (X_{ij} - c_j)\, f_j(Z_i)$$

where $g$ is the link, $Z_i$ is the vector of predictors, and each $X_{ij}$ is a predictor that has been given a coefficient. $f_0$ is a forest we will call the **control function**, following @hahn2020: the surface for the outcome when every $X_j$ is at its centering value $c_j$. Each $f_j$ is a forest of its own, and it is the coefficient on $X_j$: how much the linear predictor moves per unit of $X_j$, as a function of whatever $f_j$ is allowed to split on. With a single binary $X$, $f_1$ is the conditional treatment effect on the link scale, one value per covariate profile.

All of the forests are fitted at once. That is what buys a prior on the coefficient itself. Were $X_j$ simply one predictor among many in a single forest, its effect would be whatever difference that forest happened to produce, regularized by the same prior that regularizes everything else, and @hahn2020 show that shrinking the prognostic surface then shrinks the effect along with it. With $f_j$ separate, it can have a prior of its own, and in particular a tighter one, since the ways an effect varies are usually simpler than the surface it sits on.

## Declaring a Varying Coefficient (`vc()`)

`vc()` is a marker written inside the formula. It is never evaluated as a function (calling it directly is an error) but is read out of the formula by `bartisan()`, which builds the forests it asks for.

```{r vc, eval = run}
set.seed(2026)

fit_vc <- bartisan(death ~ age + sex + race + edu + aps +
                     meanbp + resp + hema + pafi +
                     paco2 + crea + surv2m + card +
                     vc(rhc),
                   data = rhc, family = binomial())

fit_vc
```

The print shows two forests where a fit without `vc()` would show one. These forests are the control function, named `(Intercept)`, and the coefficient, named for its covariate. Both are fitted with the same number of trees here, which is the default and is taken up below.

A predictor that has been given a coefficient should be kept out of the control function. Were `rhc` among $f_0$'s predictors as well as multiplying $f_1$, the two would not be separately identified, since any function of `rhc` can move from one to the other without changing a single prediction.

We can examine the structure of the forest for the `rhc` coefficient using `summary()`, which includes a table for each forest in the model:

```{r}
summary(fit_vc)
```

We can see that `age`, `paco2`, and `surv2m` were used in every tree in the control function (`(Intercept)`) forest, but `hema` was the most frequently used predictor in the varying coefficient (`rhc`) forest.

### Effect Modifiers (`modifiers`)

By default, a coefficient is allowed to vary with every predictor in the model except the covariate itself, and `modifiers`, the second argument to `vc()`, is a one-sided formula that narrows that:

```{r vcmod, eval = run}
set.seed(2026)

fit_aps <- bartisan(death ~ age + sex + race + edu + aps +
                      meanbp + resp + hema + pafi +
                      paco2 + crea + surv2m + card +
                      vc(rhc, ~ aps + age),
                    data = rhc, family = binomial())
```

Now the effect of catheterization may differ by severity of illness (`aps`) and by `age` and by nothing else, which is a statement about the model and not a restriction the data can undo. Narrowing is worth doing when there is reason to think the effect depends on a few things, since a forest asked to search fourteen predictors for heterogeneity that lives in two will spend some of its prior on the other twelve.

Calling `summary()` on `fit_aps`, we can see that the only variables used in the varying coefficient forest are indeed `aps` and `age`, as requested:

```{r}
summary(fit_aps)
```

The other thing `modifiers` decides is whether the effect is linear in the covariate. As written, the model is $z\, f_1(Z)$, which for a binary $z$ is no assumption at all, there being only two values, and for a continuous one says the effect is proportional to it. That is a real restriction, and a numeric covariate is allowed to modify its own coefficient to remove it:

```{r vcself, eval = FALSE}
# The effect of `aps` may itself change across `aps`, so the dose response is
# a curve rather than a line through the origin
bartisan(death ~ age + sex + vc(aps, ~ aps + age + sex),
         data = rhc, family = binomial())
```

This fits $\text{aps} \cdot f_1(\text{aps},\, \text{age},\, \text{sex})$, so the slope moves across the range of `aps`, and it is worth reaching for whenever the effect of a continuous predictor might not be proportional to it. A categorical covariate is removed from its own forests instead: a level's indicator is nonzero only on the rows where that level holds, and the variable is constant on exactly those rows, so a split on it would separate rows that contribute from rows that contribute nothing.

### Centering the Covariate (`center`)

The $c_j$ in the model are the values at which the control function is read, and `center` chooses them. This is a reparameterization of $f_0$ alone: every coefficient and every estimand is identical under any choice, and what changes is what the control function means and how well the two forests mix. The default, `"auto"`, picks by the covariate. A `0`/`1` covariate is left at zero, so $f_0$ is the surface among the binary predictor's reference level, which is a quantity with a meaning of its own. Any other numeric covariate is centered at its mean, because zero may be nowhere near the data and a control function read there would be an extrapolation. `"zero"`, `"mean"`, `"mid"` (the midpoint of the range), or a specific number override that.

A factor is handled differently. It is always fitted mean-centered and gets one forest per level, coded symmetrically rather than as contrasts against whichever level happened to sort first, and `coef()` recenters the coefficients to sum to zero across the levels. That coding leaves the reference level a reporting choice: `center` names the level to report against, and no refit is needed to change it.

`center = "estimate"` is different in kind. Rather than subtract a number from the covariate, it gives each of the covariate's values a coefficient of its own and *draws* it, so that every contrast carries the same prior whatever the number of values and no value is a reference. At two values it restricts nothing, and it is what `bcf()` uses for a binary treatment: what it buys is that the answer stops depending on which arm was written as 1. Above two values every contrast becomes one shared shape times a scalar, where the symmetric coding gives each level its own, so it is the parsimonious model against a general one, and the one to reach for when the levels plausibly differ in degree rather than in kind. It needs a covariate with between two and twenty distinct values, and a family whose leaf target is quadratic in the predictor it feeds, which is why `bcf()` falls back to the fixed coding for a family that is not.

### A Constant Coefficient (`vc(z, ~ 1)`)

Naming no modifier at all leaves the coefficient's forest nothing to split on. Every tree in it is then a stump, the coefficient is a single drawn number, and the covariate enters the model as a linear term while everything else stays nonparametric. That is the semiparametric case of the General BART model of @tan2019, a forest and a parametric part taking disjoint sets of predictors, and it is the shape *stan4bart* and *flexBART* fit^[Both packages' formula interfaces write it as `y ~ bart(x1 + x2) + x3`, with `x3` linear].

Comparing the constant coefficient against the varying one is a question neither variable importance nor a dropped predictor answers: not whether `rhc` matters, but whether what it does depends on anything else. `loo_compare()` from *loo* answers it; `vignette("comparison")` covers what it is doing.

```{r vcconstant, eval = run && has_loo}
set.seed(2026)

constant <- bartisan(death ~ age + sex + race + edu + aps +
                       meanbp + resp + hema + pafi +
                       paco2 + crea + surv2m + card +
                       vc(rhc, ~ 1),
                     data = rhc, family = binomial())

constant

library(loo)
loo_compare(list(constant = loo(constant), varying = loo(fit_vc)))
```

The constant coefficient is not beaten, so there is no evidence here that the effect of catheterization on the log odds of death varies with the other covariates. That buys an interpretation the varying fit cannot offer, because one number describes the whole sample:

```{r vccoef, eval = run && has_loo}
b <- coef(constant, draws = TRUE)[["rhc"]][, 1]

c(log_OR = mean(b),
  OR = exp(mean(b)),
  lower = quantile(exp(b), .025, names = FALSE),
  upper = quantile(exp(b), .975, names = FALSE)) |>
  round(3)
```

An odds ratio of about 1.4, conditional on the predictors in the forest. A constant conditional odds ratio is a much more clinically meaningful effect summary than a marginal effect measure because it describes the estimated effect for each patient profile as determined by the included covariates, as opposed to an average effect across a whole population. This is the same interpretation a coefficient in a logistic regression would have, but fit with BART, we can be reasonably confident that any nonlinearities and interactions among covariates are adjusted for, whereas a logistic regression must impose the model structure a researcher believes to be there.

It is important to remember that the coefficient is drawn under the leaf prior of its own forest and not under a prior written for a regression coefficient, so it is shrunk toward zero and is a regularized log odds ratio rather than a maximum likelihood one; because the leaf scale is divided by $\sqrt{m}$ for a forest of $m$ trees, that prior does not change with the number of trees. Also, the comparison between a fixed and a varying coefficient is far better powered on some outcomes than others. On a Gaussian outcome with a coefficient truly ranging from 0.3 to 2.5, the varying coefficient was easily preferred at $n = 1000$; on a binary outcome with the same coefficients it  both models were equally preferred. A null result on a binary outcome at a few hundred observations says little, and should not be read as evidence that an effect is constant.

## Reading a Fit

### The Coefficients (`coef()`)

`coef()` returns the varying coefficients and nothing else. The control function is a prediction rather than a coefficient, and `predict()` reports it.

```{r coef, eval = run}
head(coef(fit_vc))
```

`coef()` returns one row per observation and one column per coefficient, which is what a coefficient becomes when it is allowed to vary: for the first patient, or more precisely for a patient with the first patient's covariate profile, the log odds of death move by `r round(coef(fit_vc)[1, "rhc"], 3)` under catheterization. The values are on the link scale, so for a binary outcome they are differences in log odds rather than in probability. Supplying `newdata` evaluates the coefficients at other covariate profiles, and setting `draws = TRUE` returns every posterior draw rather than the mean, as a list with one draws-by-observations matrix per coefficient, which is what an interval on any one patient's coefficient needs.

```{r coefdraws, eval = run}
draws <- coef(fit_vc, draws = TRUE)[["rhc"]]

# A 95% interval on the coefficient for the first three patients
t(apply(draws[, 1:3], 2L, quantile, c(.025, .975)))
```

`summary()` on the fit reports the forests as it does for any fit, and `prior_summary()` writes out the prior each one was drawn under, including the leaf scale of the coefficient forest, which is where the regularization that separates the effect from the surface lives.

### Effects From the Coefficients (`estimate_effect()`)

A coefficient on the link scale is one step from an effect on the response scale, and `estimate_effect()` makes it easier to report meaningful contrasts. With `treat` naming the covariate, it predicts outcomes under each of the predictor's values and contrasts them. Setting `estimand = "CATE"` returns one effect per unit, here on the risk difference scale, and `plot()` on that draws them ordered by their estimate with the marginal effect as a separate interval past the right edge.

```{r cate, eval = run, fig.height = 3.4}
cate <- estimate_effect(fit_vc, treat = "rhc", estimand = "CATE")

plot(cate)
```

The marginal effect, effects by subgroup, or the effect as a ratio are all the same call with different arguments. For example, setting `comparison = "or"` displays the effects on the conditional odds ratio scale. `?estimate_effect` documents these arguments, and `vignette("causal")` is where the assumptions live that turn any of them into a causal effect.

## Settings That Vary by Forest

Every forest has its own prior, and every setting in `bartisan_control()` that could mean something different for one forest than for another may be given once, to apply to all, or once per forest. Given per forest, the values are either positional in the order the forests are listed in the fit's print, or keyed by the forest names. Both of these ask for fifty trees in the control function and twenty-five in the coefficient:

```{r trees, eval = FALSE}
bartisan(death ~ age + sex + race + edu + aps +
           meanbp + resp + hema + pafi +
           paco2 + crea + surv2m + card +
           vc(rhc),
         data = rhc, family = binomial(), num_trees = c(50L, 25L))

bartisan(death ~ age + sex + race + edu + aps +
           meanbp + resp + hema + pafi +
           paco2 + crea + surv2m + card +
           vc(rhc),
         data = rhc, family = binomial(),
         num_trees = c("(Intercept)" = 50L, rhc = 25L))
```

A forest a named argument does not mention keeps that argument's own default rather than borrowing another forest's value. The settings this covers are `num_trees`, `k`, `sigma_mu`, `sparsity`, `split_prior`, `bandwidth`, `gamma`, `beta`, the four `alpha` arguments, and the three `update_` flags; `?bartisan_control` lists them.

Fewer trees for a coefficient than for the control function is the setting worth knowing about, and it is the one `bcf()` makes by default. A coefficient forest describes how an effect varies, and that is usually a simpler surface than the outcome's, so it needs less capacity; and with less capacity it competes less with the control function for the same variation, which is what keeps the two well separated in the posterior. When a fit with a varying coefficient mixes badly, a coefficient forest that is too large is the first thing to try.

`sparsity` is the other setting that reads differently on a coefficient forest. On the control function it selects among the predictors the way it does in any BART fit. On a coefficient forest what it selects among is the *modifiers*, and dropping all of them leaves an effect that does not vary rather than an effect that is zero, since nothing can drop the covariate the forest multiplies. That is the shrinkage a heterogeneity model wants, and it is why `bcf()` leaves `sparsity` on where a single-forest treatment model turns it off; `vignette("effects")` covers the single-forest case.

### Families With Several Additive Predictors

`gaussian_ls()`, `zi_poisson()` and the other families with more than one additive predictor fit one control function per parameter, and each parameter's formula carries its own `vc()` terms. The forests are then two-dimensional, a control function and its coefficients for each parameter, and are named accordingly:

```{r vcmulti, eval = FALSE}
# Forests: mean, mean:z, log_sd
bartisan(list(mean = y ~ x1 + x2 + vc(z),
              log_sd = ~ x1 + x2),
         data = d,
         family = gaussian_ls())

# Forests: mean, mean:z, log_sd, log_sd:z. One formula reaches every parameter,
# which is the rule every per-forest argument follows too
bartisan(y ~ x1 + x2 + vc(z), data = d, family = gaussian_ls())

# Each coefficient with modifiers of its own
bartisan(list(mean = y ~ x1 + x2 + vc(z, ~ x2),
              log_sd = ~ x1 + x2 + vc(z, ~ x1)),
         data = d, family = gaussian_ls())
```

So the same covariate may have a coefficient on more than one parameter. `z` shifting the mean and `z` widening the spread are different questions, and both are answered at once, and `coef()` returns one column per coefficient, named `mean:z` and `log_sd:z` for the forests they come from.

Those names are also how the per-forest settings are keyed, and every forest can be set at once. With four forests, a control function and a coefficient for each parameter, `num_trees` takes four values, either positionally in the order the forests are listed in the fit's print or by name:

```{r vcmultitrees, eval = FALSE}
# The mean's control function and coefficient, then the log standard
# deviation's, in the order the print lists them
bartisan(y ~ x1 + x2 + vc(z), data = d, family = gaussian_ls(),
         num_trees = c(50L, 20L, 15L, 10L))

# The same, by name; the order then does not matter
bartisan(y ~ x1 + x2 + vc(z), data = d, family = gaussian_ls(),
         num_trees = c("mean"   = 50L, "mean:z"   = 20L,
                       "log_sd" = 15L, "log_sd:z" = 10L))
```

A forest a named vector leaves out keeps the argument's own default rather than taking its parameter's value: `num_trees = c(mean = 50L, log_sd = 15L)` gives both coefficient forests the default of 50, not 50 and 15. Name the coefficient forests too when they are meant to differ.

## Bayesian Causal Forests (`bcf()`)

The Bayesian causal forest of @hahn2020 is a varying coefficient model with one binary covariate, the treatment, and a handful of choices made for the causal problem. `bcf()` implements this model with these modifications to support treatment effect estimation. Given a formula for the covariates and the treatment named in `treat`, it writes a `bartisan()` call that could have been written directly:

```{r bcf, eval = run}
set.seed(2026)

fit_bcf <- bcf(death ~ age + sex + race + edu + aps +
                 meanbp + resp + hema + pafi +
                 paco2 + crea + surv2m + card,
               treat = ~ rhc, data = rhc, family = binomial())

fit_bcf
```

`bcf()` makes the following modifications to a standard `bartisan()` call:

1. The treatment gets a `vc()` term, so its effect is a forest with a prior of its own, which is everything above. 2. The estimated propensity score, fitted by a logistic BART model on the same covariates, goes into the control function as a predictor named `.propensity` and not into the effect forest, so that the control function absorbs the selection into treatment while the effect stays free of it; that is the remedy @hahn2020 propose for what they call regularization-induced confounding.
3. The effect forest gets fewer trees than the control function, 25 against 50, for the reason given in the section above.
4. A binary treatment's coding is drawn rather than fixed, which is `center = "estimate"`, so the answer is the same whichever arm was written as 1. 
5. `sparsity` is left on, since the treatment is the coefficient rather than a predictor a splitting proportion could drop (note this is in line with the `bartisan()` default, but not what is otherwise recommend for treatment effect estimation without a varying coefficient model).

`moderators` is `vc()`'s `modifiers` under the name the causal literature uses, and the propensity model is `bartisan()` with `propensity_args` for its settings.

Written out by hand, and leaving aside that the propensity score would have to be fitted and added to the data first, the model is

```{r bcfhand, eval = FALSE}
bartisan(death ~ age + sex + race + edu + aps +
           meanbp + resp + hema + pafi +
           paco2 + crea + surv2m + card + .propensity +
           vc(rhc, center = "estimate"),
         data = rhc, family = binomial(), num_trees = c(50L, 25L))
```

and every method that works on a `bartisan()` fit works on a `bcf()` fit unchanged, since the second is the first with the treatment recorded. What the record adds is that `estimate_effect()` needs no `treat`, `print()` names the treatment and its moderators, and `plot()` on the fit is `plot(estimate_effect(fit, estimand = "CATE"))`:

```{r bcfplot, eval = run, fig.height = 3.4}
plot(fit_bcf)
```

The reasons to write the model by hand rather than through `bcf()` are the ones this guide is about: a coefficient on something other than a treatment, more than one of them, a continuous covariate whose effect should not be linear in it, or a family with several additive predictors. `vignette("causal")` covers what has to be true of the design for any of these effects to be read causally, which no model supplies.

## Random Intercepts (`(1 | group)`)

A `(1 | group)` term in the formula adds an intercept per level of `group`, drawn from a common mean-zero normal whose standard deviation is itself drawn, under the same half-Cauchy prior the leaf scale uses [@gelman2006; @polson2012]. Several grouping factors are allowed, and `(1 | a/b)` expands to nesting as it does in *lme4*.

The `rhc` data has no grouping structure, so we simulate one: forty groups of about fifteen observations each, a nonlinear effect of one covariate, and a group effect with a standard deviation of one half.

```{r resim, eval = run}
set.seed(2026)

n_groups <- 40L
d <- data.frame(g = factor(sample(n_groups, 600L, replace = TRUE)),
                x = runif(600L))
u <- rnorm(n_groups, 0, 0.5)
d$y <- sin(2 * pi * d$x) + u[d$g] + rnorm(600L, 0, 0.5)

fit_re <- bartisan(y ~ x + (1 | g), data = d)

fit_re
```

The intercepts are in `fit_re$ranef` and the standard deviation they were drawn under in `fit_re$tau`, one matrix of draws per additive predictor with a column per grouping factor, and `summary()` reports that scale beside the leaf scale. `ranef()` is the generic from *nlme*, which *lme4* re-exports, so either qualification reaches the method; `draws = TRUE` gives the draws, which is what an interval on a group needs.

```{r ranef, eval = run && has_nlme}
# How much of the variation is between groups: the posterior of the group scale
quantile(fit_re$tau[[1L]][, "g"], c(.025, .5, .975))

# The intercepts, as posterior means, against the values they were drawn from
re <- nlme::ranef(fit_re)$g
cor(re[, "(Intercept)"], u)

# With intervals, for the first four groups
apply(nlme::ranef(fit_re, draws = TRUE)$g[["(Intercept)"]][, 1:4], 2L,
      quantile, c(.025, .975))
```

A posterior mean is the wrong summary for a level with few observations, which is the case a group intercept exists for, and a level whose interval covers zero is one the data had little to say about. The intercepts are shrunk toward zero by their prior and are deviations from the additive predictor, so they come out approximately centered without being constrained to sum to zero exactly; nothing is lost by that, since the level of the fitted function is the additive predictor's.

There are a few details about this term worth knowing. Only intercepts are supported, and a random slope is refused rather than ignored. A random intercept is a scalar entering the predictor with weight one for the observations in its level, which is exactly what a leaf is once its gate is removed, so the sampler's leaf machinery handles it, closed forms included; a slope is a scalar multiplying a covariate, which is a different shape of parameter. A variable whose effect varies by group belongs in the fixed part of the formula, where a tree can split on the group and on the variable together and get an interaction of any shape. A grouping factor can also go in the fixed part, where a tree splits on it like anything else, and with few large groups that is the better choice: the group means are well determined without pooling, and a split can interact the group with the covariates. The random intercept wins where there are many small groups, which is where partial pooling earns its keep. A level not seen at fitting time is given the prior mean of zero when predicting, with a warning.

A family with several additive predictors gets a separate set of intercepts for each, independent of one another, so a zero-inflated count model has a group effect on the count part and another on the inflation part, and the columns of `ranef()` are named for the predictors rather than `(Intercept)`. A group intercept reaches every control function and no coefficient, since a group-varying coefficient would be a random slope. `prior_summary()` reports the prior the scale was drawn under.

## Further Reading

`?vc` is the reference for the arguments used here, and `?bcf` for what it writes into a `bartisan()` call and how it handles a categorical or continuous treatment. `vignette("causal")` covers the assumptions under which a coefficient on a treatment is a causal effect, and `vignette("effects")` the contrasts and curves that read an effect off a fit without a coefficient. `vignette("comparison")` covers `loo_compare()`, which is what the constant-against-varying comparison above rests on. `?bartisan` documents the `(1 | group)` syntax, and `?ranef.bartisan_fit` what comes back.

## References
