---
title: "Getting started with bartisan"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Getting started with bartisan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.4, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 4)

has_me <- rlang::is_installed("marginaleffects")
has_gg <- rlang::is_installed("ggplot2")
has_loo <- rlang::is_installed("loo")
has_mass <- rlang::is_installed("MASS")
has_bp <- rlang::is_installed("bayesplot")
run <- has_mass
```

## Introduction

*bartisan* fits Bayesian additive regression trees (BART) using the same interface
as `glm()`. You supply a formula, a data frame, and a family, and the model
estimates the relationship between the predictors and the outcome without you
having to say what shape that relationship takes. Nonlinearity and interactions
are found rather than specified.

This vignette walks through a complete analysis: fitting a model, checking that
it worked, seeing which predictors it uses, reading off the effects, predicting
for new observations, and comparing models. It assumes you are comfortable with
regression but does not assume familiarity with machine learning or Bayesian
methods.

The main thing to take from it is that the defaults are meant to be used. The
priors, the number of trees, and the sampler settings are chosen to work across a
wide range of problems, and tuning them is rarely where the gains are. Almost
everything below is a single function call with no arguments beyond the formula
and the data.

Each section ends with a pointer to a vignette that covers the same ground in
more depth.

```{r libs, eval = run}
library(bartisan)
```

## The data

We use the birth weight data from @venables2002, collected at Baystate Medical
Center in 1986. The outcome is birth weight in grams, and the predictors are
characteristics of the mother.

```{r data, eval = run}
data("birthwt", package = "MASS")

d <- birthwt

# Give the categorical variables labels so the output is readable.
d$race  <- factor(d$race,  levels = 1:3, labels = c("white", "black", "other"))
d$smoke <- factor(d$smoke, levels = 0:1, labels = c("no", "yes"))
d$ht    <- factor(d$ht,    levels = 0:1, labels = c("no", "yes"))
d$ui    <- factor(d$ui,    levels = 0:1, labels = c("no", "yes"))

# `low` is birth weight dichotomized at 2500g, so it is the outcome in disguise.
d$low <- NULL

str(d)
```

The predictors are the mother's age (`age`), weight at her last menstrual period
in pounds (`lwt`), race, smoking during pregnancy (`smoke`), number of previous
premature labours (`ptl`), history of hypertension (`ht`), presence of uterine
irritability (`ui`), and number of physician visits in the first trimester
(`ftv`).

The question is how these relate to birth weight, and in particular how much
lower birth weight is among infants of mothers who smoked.

## Fitting the model

```{r fit, eval = run}
set.seed(2026)

fit <- bartisan(bwt ~ ., data = d, family = gaussian(), chains = 4)

fit
```

That is the whole call. `family = gaussian()` says the outcome is continuous with
additive noise, exactly as in `glm()`. The family may be omitted, in which case
it is read off the outcome and reported; naming it is clearer and silences the
message.

`chains = 4` runs the sampler four times from different starting points. The
default is one chain, which is enough to get estimates, but running several is
what makes the convergence diagnostics in the next section available. Each chain
here takes under a second.

The printed summary reports the structure of the fit and `sigma`, the residual
standard deviation, at about 646 grams.

Choosing a family is the one modeling decision that usually matters more than any
sampler setting. `vignette("families")` covers the choice, and
`vignette("survival")` covers censored outcomes.

## Checking the model

Two questions are worth separating: whether the sampler converged, and whether
the model fits.

### Convergence

```{r rhat, eval = run}
fit$rhat
```

`rhat` compares variation between chains to variation within them. Values near 1
indicate the chains have settled on the same answer. `ess_bulk` and `ess_tail`
are effective sample sizes, and count how many independent draws the correlated
ones are worth.

Read the rows that correspond to quantities you will report. Here `aux.sigma`
(the residual standard deviation) and `eta.eta` (the fitted values, summarized
over the observation that mixes worst) are both close to 1, which is what
matters. `sigma_mu.eta` is a prior scale parameter internal to the forest; it
mixes slowly in most BART fits and running longer does not help much. Conclusions
do not rest on it.

If the rows you care about are far from 1, increase `num_burn` and `num_save` in
`bartisan_control()`.

### Fit

A posterior predictive check simulates new outcomes from the fitted model and
compares their distribution to the observed one.

```{r ppcheck, eval = run && has_bp, fig.height = 3.2}
bayesplot::pp_check(fit)
```

The simulated densities (light) track the observed density (dark) closely enough,
with the observed data slightly heavier in the lower tail. Nothing here suggests
the Gaussian family is the wrong choice.

`vignette("diagnostics")` covers both questions in more detail, including trace
plots and what to do when convergence is poor.

## Which predictors the model uses

```{r varimp, eval = run}
variable_importance(fit)
```

`splits` is the average number of splitting rules the forest spends on each
predictor per draw, and `prop_used` is the proportion of draws in which the
predictor received any rule at all.

By default the forest uses a sparsity-inducing prior [@linero2018sparse] that
lets it stop splitting on predictors that do not earn their place, so `prop_used`
behaves like a posterior probability that a predictor belongs in the model. When
a predictor is useless it typically falls to near zero.

Here nothing falls. Every predictor is used in at least 92% of draws, which says
the sample is not large enough to rule any of them out rather than that all of
them matter. With 189 observations and a residual standard deviation not much
below the outcome's own, there is little for the prior to work with. Do not read
this table as a selection result.

Two cautions that apply whatever the sample size. Usage is not effect size: a
predictor can be split on constantly and still move the prediction very little,
and `avg_comparisons()` in the next section is the better guide to that. And when
predictors are correlated the usage distributes among them more or less
arbitrarily, so a predictor can matter and still show low usage because a
correlated partner absorbed it.

`vignette("importance")` covers variable importance and selection.

## Interpreting the fit

A forest has no coefficients, so there is no table of slopes to read. The
question "what is the effect of smoking" is answered by asking the fitted model
what it predicts when everyone smokes, asking again when nobody does, and taking
the difference. `marginaleffects` does this.

```{r effects, eval = run && has_me}
library(marginaleffects)

avg_comparisons(fit)
```

This is the closest analogue to a coefficient table, and it is read the same way.
For binary and categorical predictors the estimate is the average difference in
predicted birth weight between levels. For numeric predictors it is the average
change associated with a one-unit increase.

Infants of mothers who smoked weigh about 290 grams less on average, with an
interval from roughly 500 to 90 grams. Uterine irritability is associated with a
larger difference, about 445 grams. Age and number of physician visits have
intervals comfortably covering zero.

The intervals are posterior credible intervals. A 95% interval is the range the
model considers most plausible given the data, and it already includes
uncertainty about the shape of the relationship, which a linear model excludes by
assuming the shape is known.

For comparison, a linear model gives a somewhat larger smoking coefficient:

```{r lmcompare, eval = run}
round(coef(summary(lm(bwt ~ ., data = d)))["smokeyes", ], 1)
```

The two answers are compatible. The BART estimate is pulled toward zero because
the prior shrinks the fitted function, and its interval is wider because it does
not assume the relationship is linear. When the truth really is linear, a linear
model will be more precise. The value of BART is that you do not have to know
that in advance.

### Looking at a relationship

Effects averaged over the sample hide the shape of the relationship. To see the
shape, plot the model's predictions against one predictor.

```{r pdp, eval = run && has_me && has_gg}
plot_predictions(fit, condition = "lwt")
```

Predicted birth weight rises with the mother's weight up to about 125 pounds and
is roughly flat above that. A linear model fits a straight line through this and
reports a single slope, which understates the relationship at the low end and
overstates it at the high end. Nothing had to be specified to find the bend; it
came out of the fit.

The band is a credible interval, and it is wide at both ends where there are few
observations.

Conditioning on a second variable shows how the relationship differs across
groups:

```{r pdp2, eval = run && has_me && has_gg}
plot_predictions(fit, condition = c("lwt", "smoke"))
```

The two curves are roughly parallel, which says the association with the mother's
weight is similar for smokers and non-smokers and that the smoking difference is
roughly constant across it. If they converged or crossed, that would be an
interaction, and the forest would have found it without being told to look.

`vignette("effects")` covers effects, curves, and interactions.

## Predicting new observations

`predict()` returns the posterior mean prediction.

```{r pred, eval = run}
new_mother <- data.frame(age = 25, lwt = 130, race = "white", smoke = "yes",
                         ptl = 0, ht = "no", ui = "no", ftv = 1)

predict(fit, newdata = new_mother)
```

For a prediction with an interval, use `predictions()`:

```{r predint, eval = run && has_me}
predictions(fit, newdata = new_mother)
```

This interval describes the average birth weight for mothers with these
characteristics. It is not the range within which a single infant's weight is
likely to fall, because it excludes the residual variation around the mean. For
that, simulate outcomes rather than means:

```{r ppi, eval = run && rlang::is_installed("rstantools")}
draws <- rstantools::posterior_predict(fit, newdata = new_mother)

round(quantile(draws, c(0.025, 0.5, 0.975)))
```

The wider interval is the one to quote when the question is about an individual.
Both are needed for different questions, and confusing them is a common mistake.

Note that `predictions()` summarizes the posterior with the median while
`predict()` reports the mean, so the two differ slightly for a skewed posterior.

## Comparing models

Approximate leave-one-out cross-validation estimates how well a model predicts
data it has not seen.

```{r loo, eval = run && has_loo}
library(loo)

loo(fit)
```

`elpd_loo` is the estimated log predictive density on held-out data, where higher
is better. `p_loo` is the effective number of parameters, about 13 here, which is
a measure of how much of the data the forest is actually using. The Pareto k
diagnostics are all good, meaning the approximation is trustworthy for this fit.

Two models can be compared directly. Here we ask whether the predictors beyond
age, weight, and smoking earn their keep:

```{r loocompare, eval = run && has_loo}
set.seed(2026)
reduced <- bartisan(bwt ~ age + lwt + smoke, data = d, family = gaussian(),
                    chains = 4)

loo_compare(loo(fit), loo(reduced))
```

The full model predicts better by about 14 points on the elpd scale, with a
standard error of about 5. A difference of roughly three standard errors is
reasonably clear evidence that the additional predictors help.

Model comparison means something slightly different here than in linear
regression. You are not choosing which terms to include, since the forest handles
that. You are choosing which variables the model is allowed to see, and which
likelihood to use.

`vignette("comparison")` covers model comparison, and the cases where
leave-one-out fails.

## What to be careful about

The model is flexible about the shape of the relationship and nothing else.

It does not make an association causal. The smoking estimate above is an
adjusted association. Whether it can be read as the effect of smoking depends on
whether the variables in the model are enough to account for confounding, which
is a question about the study and not about the fit. A forest fitted to
confounded data returns a confounded answer with a tight interval around it.
`vignette("causal")` covers what is required.

It does not extrapolate reliably. Predictions for predictor values outside the
range of the training data are shrunk toward the overall mean rather than
continuing any trend. This is usually the safer behaviour, but it means the model
is not a tool for forecasting beyond the observed range.

It does not fix a badly chosen family. Getting the outcome distribution wrong
matters more than any sampler setting.

## Where to go next

| Topic | Vignette |
| --- | --- |
| How BART works, and the sampler | `vignette("bartisan")` |
| Choosing a family | `vignette("families")` |
| Convergence and fit | `vignette("diagnostics")` |
| Variable importance and selection | `vignette("importance")` |
| Effects, curves, and interactions | `vignette("effects")` |
| Model comparison | `vignette("comparison")` |
| Causal inference | `vignette("causal")` |
| Censored and survival outcomes | `vignette("survival")` |

`?bartisan-package` has a shorter version of the same map, organized by task.

## References
