---
title: "Getting started with bartisan"
author: "Noah Greifer"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Getting started with bartisan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 3.2, dpi = 110, fig.align = "center",
                      message = FALSE, warning = FALSE)
options(digits = 4, marginaleffects_safe = FALSE)

has_me <- rlang::is_installed("marginaleffects")
has_gg <- rlang::is_installed("ggplot2")
has_loo <- rlang::is_installed("loo")
has_bp <- rlang::is_installed("bayesplot")
```

## Introduction

*bartisan* fits Bayesian additive regression trees (BART) using the same
interface as `glm()`. You supply a formula, a data frame, and a family, and the
model estimates the relationship between the predictors and the outcome without
you having to say what shape that relationship takes. Nonlinearity and
interactions are found rather than specified.

This vignette walks through a complete analysis: fitting a model, checking that
it worked, seeing which predictors it uses, reading off the effects, predicting
for new observations, and comparing models. It assumes you are comfortable with
regression but does not assume familiarity with machine learning or Bayesian
methods.

The main thing to take from it is that the defaults are meant to be used. The
priors, the number of trees, and the sampler settings are chosen to work across a
wide range of problems, and tuning them is rarely where the gains are. Almost
everything below is a single function call with no arguments beyond the formula
and the data.

Each section ends with a pointer to a vignette that covers the same ground in
more depth.

```{r libs}
library(bartisan)
```

## The data

`rhc` records 1500 critically ill patients from the SUPPORT study and whether
each received right heart catheterization, a monitoring procedure, within a day
of arriving in intensive care [@connors1996]. The question the study asked is
whether the procedure helps or harms.

```{r data}
data(rhc)

str(rhc)
```

The outcome comes in two forms. `death` is whether the patient died during
follow-up, and `days` is how long that took. This vignette uses the binary form;
`vignette("survival")` uses the other.

The remaining variables are the patient's age, sex, race and years of education,
seven physiological measurements taken on the first day, whether cardiovascular
disease was among the diagnoses, and `surv2m`, the study's own estimate of the
patient's chance of surviving two months. All were recorded before
catheterization.

## Fitting the model

```{r fit}
set.seed(2026)

fit <- bartisan(
  death ~ rhc + age + sex + race + edu + aps + meanbp + resp + hema +
    pafi + paco2 + crea + surv2m + card,
  data = rhc, family = binomial(), chains = 4
)

fit
```

That is the whole call. `family = binomial()` says the outcome is binary, exactly
as in `glm()`. The family may be omitted, in which case it is read off the
outcome and reported; naming it is clearer and silences the message.

`chains = 4` runs the sampler four times from different starting points. The
default is one chain, which is enough to get estimates, but running several is
what makes the convergence diagnostics in the next section available.

Choosing a family is the one modeling decision that usually matters more than any
sampler setting. `vignette("families")` covers the choice, and
`vignette("survival")` covers censored outcomes such as this one's `days`.

## Checking the model

Two questions are worth separating: whether the sampler converged, and whether
the model fits.

### Convergence

```{r rhat}
fit$rhat
```

`rhat` compares variation between chains to variation within them. Values near 1
indicate the chains have settled on the same answer. `ess_bulk` and `ess_tail`
are effective sample sizes, and count how many independent draws the correlated
ones are worth.

Read the rows that correspond to quantities you will report. `eta.eta` is the
fitted function, summarized over the observation that mixes worst, so it is the
row that matters here. Forests mix slowly on their fitted values and a figure
above 1.01 on that row is ordinary rather than alarming;
`vignette("diagnostics")` explains what to do about it and when to worry.

### Fit

A posterior predictive check simulates new outcomes from the fitted model and
compares their distribution to the observed one.

```{r ppcheck, eval = has_bp, fig.height = 3.0}
bayesplot::pp_check(fit)
```

For a binary outcome this is a weaker check than it is for a continuous one:
there are only two values to get right, so the replicates match unless something
has gone badly wrong. `vignette("diagnostics")` covers what to do instead.

## Which predictors the model uses

```{r varimp}
variable_importance(fit)
```

`splits` is the average number of splitting rules the forest spends on each
predictor per draw, and `prop_used` is the proportion of draws in which the
predictor received any rule at all.

`surv2m` takes the most rules, which is unsurprising: it is a prognostic score
built to predict survival. Age follows. At the bottom, race and sex are used in
about half the draws, which says the model can often do without them.

Two cautions. Usage is not effect size: a predictor can be split on constantly
and still move the prediction very little, and `avg_comparisons()` in the next
section is the better guide to that. And when predictors are correlated the usage
distributes among them more or less arbitrarily.

`vignette("importance")` covers variable importance and selection, including how
to tell whether a difference in this table means anything.

## Interpreting the fit

A forest has no coefficients, so there is no table of slopes to read. The
question "what is the effect of catheterization" is answered by asking the fitted
model what it predicts when every patient receives it, asking again when none
does, and taking the difference. `marginaleffects` does this.

```{r effects, eval = has_me}
library(marginaleffects)

avg_comparisons(fit, variables = "rhc")
```

Catheterization is associated with an increase of about six percentage points in
the probability of death. The interval runs from roughly zero to eleven points,
so the direction is reasonably clear and the size is not.

The lower bound is exactly zero rather than merely close to it, and that is worth
knowing about. The default splitting prior can drop a predictor from the forest
entirely, and in a draw where it drops `rhc` the contrast is exactly zero, so the
posterior has a point mass there. When a particular contrast is the estimand,
`sparsity = FALSE` weights the predictors equally and removes it;
`vignette("causal")` explains the trade and `?bartisan_control` has the
measurements.

Because the outcome is binary, this is a difference in probability, which is
interpretable without reference to the model. That is usually the number to
report. The underlying probabilities are also worth showing:

```{r risks, eval = has_me}
avg_predictions(fit, variables = "rhc")
```

About 63% of patients would be expected to die without catheterization and 69%
with it, averaging over the covariates as they actually occur in this sample.

For comparison, a logistic regression gives a somewhat larger difference:

```{r glmcompare, eval = has_me}
avg_comparisons(glm(formula(fit), data = rhc, family = binomial()),
                variables = "rhc")
```

The two answers are compatible. The BART estimate is pulled toward zero because
the prior shrinks the fitted function, and its interval is wider because it does
not assume the relationship is linear on the log-odds scale. When the truth
really is linear there, logistic regression will be more precise. The value of
BART is that you do not have to know that in advance.

### Looking at a relationship

Effects averaged over the sample hide the shape of the relationship. To see the
shape, plot the model's predictions against one predictor.

Asking for the numbers rather than the plot gives more control over how it is
drawn:

```{r pdp, eval = has_me && has_gg, fig.height = 3.0}
library(ggplot2)

curve <- plot_predictions(fit, condition = "surv2m", draw = FALSE)

ggplot(curve, aes(surv2m, estimate)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  geom_line() +
  labs(x = "estimated probability of surviving two months",
       y = "fitted probability of death") +
  theme_bw(base_size = 9)
```

The fitted probability of death falls from about 0.88 to about 0.45 as the
prognostic score rises, and the fall is not a straight line. A logistic
regression reports one slope on the log-odds scale for the whole range. Nothing
had to be specified to find the shape.

The band is a credible interval, and it widens at the top where few patients
were that healthy.

`vignette("effects")` covers effects, curves, and interactions.

## Predicting new observations

`predict()` returns the posterior mean prediction.

```{r pred}
new_patient <- rhc[1, ]
new_patient$rhc <- 1

predict(fit, newdata = new_patient)
```

For a prediction with an interval, use `predictions()`:

```{r predint, eval = has_me}
predictions(fit, newdata = new_patient)
```

This describes the probability that a patient with these characteristics dies. It
is an interval for that probability, not a statement about which way any
individual patient will go: the outcome itself is either 0 or 1, and a
probability of 0.7 is entirely compatible with survival.

For a continuous outcome the distinction between an interval for the mean and an
interval for a new observation matters a great deal, and
`vignette("effects")` covers it. For a binary outcome the second is rarely what
anyone wants.

## Comparing models

Approximate leave-one-out cross-validation estimates how well a model predicts
data it has not seen.

```{r loo, eval = has_loo}
library(loo)

loo(fit)
```

`elpd_loo` is the estimated log predictive density on held-out data, where higher
is better. `p_loo` is the effective number of parameters, which is a measure of
how much of the data the forest is actually using. The Pareto k diagnostics are
all good, meaning the approximation is trustworthy for this fit.

Two models can be compared directly. Here we ask whether the physiological
measurements earn their keep over knowing the patient's demographics alone:

```{r loocompare, eval = has_loo}
set.seed(2026)
demographics <- bartisan(death ~ rhc + age + sex + race + edu, data = rhc,
                         family = binomial(), chains = 4)

loo_compare(list(full = loo(fit), demographics = loo(demographics)))
```

The full model predicts better by around six times the standard error of the
difference, which is what you would expect: how sick a patient is on arrival is
the main thing that predicts whether they die.

`vignette("comparison")` covers model comparison, and the cases where
leave-one-out fails.

## What to be careful about

The model is flexible about the shape of the relationship and nothing else.

It does not make an association causal. Patients were not randomized to
catheterization; sicker patients were more likely to receive it, which is exactly
the kind of confounding that can produce an apparent harm. Whether the estimate
above can be read as the effect of the procedure depends on whether the
covariates account for that selection, which is a question about the study and
not about the fit. `vignette("causal")` covers what is required, using this same
data.

It does not extrapolate reliably. Predictions for predictor values outside the
range of the training data are shrunk toward the overall mean rather than
continuing any trend.

It does not fix a badly chosen family. Getting the outcome distribution wrong
matters more than any sampler setting.

## Where to go next

| Topic | Vignette |
| --- | --- |
| How BART works, and the sampler | `vignette("bartisan")` |
| Choosing a family | `vignette("families")` |
| Convergence and fit | `vignette("diagnostics")` |
| Variable importance and selection | `vignette("importance")` |
| Effects, curves, and interactions | `vignette("effects")` |
| Model comparison | `vignette("comparison")` |
| Causal inference | `vignette("causal")` |
| Censored and survival outcomes | `vignette("survival")` |

`?bartisan-package` has a shorter version of the same map, organized by task.

## References
