---
title: "bartisan timing benchmark"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
params:
  n: 1000
  p: 10
  reps: 3
  trees: 50
  draws: 1000
  quick: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE,
                      message = FALSE)
options(digits = 4)

# bartMachine starts a JVM the first time it is touched, and these have to be set
# before that happens -- afterwards they are ignored and every fit fails. They
# are set here, in the first chunk, for that reason and no other.
options(java.parameters = c("-Xmx8g", "--add-modules=jdk.incubator.vector",
                            "-XX:+UseZGC"))
```

This document times **bartisan** against the other BART packages on this machine
and reports, for each fit, the wall clock, the effective sample size of the
fitted function, the effective samples per second, and the held-out accuracy.
Speed on its own is not the quantity of interest: a sampler that is twice as fast
and mixes half as well has bought nothing, and one that is fast and wrong has
bought less than nothing.

Render it with defaults, or scale it:

```r
rmarkdown::render("_dev/benchmark.Rmd")
rmarkdown::render("_dev/benchmark.Rmd", params = list(n = 5000, reps = 5))
rmarkdown::render("_dev/benchmark.Rmd", params = list(quick = TRUE))   # a smoke test
```

## Is this build optimized?

**Read this before reading any timing below.** `devtools::load_all()` and
`devtools::install()` compile C++ without optimization by default, which makes
this package five to twenty times slower with nothing else out of place. The
project's `.Rprofile` sets `options(pkg.build_extra_flags = FALSE)` to prevent
it, but that only applies to sessions started in the package directory.

```{r optimized}
library(bartisan)

optimized <- bartisan:::.bartisan_optimized()

if (!optimized) {
  stop("This build is unoptimized; every timing below would be meaningless. ",
       "Rebuild with:  pkgbuild::clean_dll(); R CMD INSTALL --preclean .")
}

optimized
```

## Setup

```{r libs}
library(dbarts)

have <- function(pkg) requireNamespace(pkg, quietly = TRUE)

# Whether a backend can actually fit, not whether its package is installed. The
# two came apart for bartMachine, which needs a Java runtime that R can find:
# loading the package succeeds, `set_bart_machine_num_cores()` succeeds, and then
# every fit fails. So each backend is checked by fitting something tiny.
#
# The usual cause of a bartMachine failure here is JAVA_HOME being unset, which
# leaves /usr/bin/java unable to locate a runtime even when one is installed.
# Setting it in ~/.Renviron fixes it for every R session:
#   JAVA_HOME=/Library/Java/JavaVirtualMachines/temurin-21.jdk/Contents/Home
usable <- function(pkg, probe) {
  if (!have(pkg)) return("not installed")
  out <- tryCatch({ probe(); TRUE },
                  error = function(e) conditionMessage(e))
  if (isTRUE(out)) "yes" else paste("cannot run:", out)
}

xs <- matrix(stats::runif(60), 20, 3)
colnames(xs) <- c("a", "b", "c")
ys <- stats::rnorm(20)

backends <- c(
  bartisan = "yes",
  dbarts = usable("dbarts", function()
    dbarts::bart(xs, ys, ntree = 2L, nskip = 2L, ndpost = 2L, verbose = FALSE,
                 keeptrees = FALSE)),
  BART = usable("BART", function()
    utils::capture.output(BART::wbart(xs, ys, ntree = 2L, nskip = 2L,
                                      ndpost = 2L))),
  stochtree = usable("stochtree", function()
    stochtree::bart(X_train = xs, y_train = ys, num_gfr = 0L, num_burnin = 2L,
                    num_mcmc = 2L, mean_forest_params = list(num_trees = 2L))),
  bartMachine = usable("bartMachine", function() {
    bartMachine::set_bart_machine_num_cores(1L)
    bartMachine::bartMachine(as.data.frame(xs), ys, num_trees = 2L,
                             num_burn_in = 2L,
                             num_iterations_after_burn_in = 2L,
                             verbose = FALSE, serialize = FALSE)
  }))

installed <- backends == "yes"
data.frame(backend = names(backends), status = unname(backends))
```

```{r params}
n     <- if (params$quick) 300  else params$n
p     <- if (params$quick) 5    else params$p
reps  <- if (params$quick) 1L   else params$reps
trees <- if (params$quick) 20L  else params$trees
draws <- if (params$quick) 200L else params$draws

c(n = n, predictors = p, replicates = reps, trees = trees,
  warmup = draws, saved = draws)
```

The Friedman function, which is the standard BART test surface: an interaction,
a quadratic, two linear terms, and `p - 5` predictors that do nothing.

```{r data}
friedman <- function(x) {
  10 * sin(pi * x[, 1] * x[, 2]) + 20 * (x[, 3] - 0.5)^2 + 10 * x[, 4] +
    5 * x[, 5]
}

make_data <- function(n, p, seed) {
  set.seed(seed)
  x <- matrix(runif(2 * n * p), nrow = 2 * n)
  colnames(x) <- paste0("x", seq_len(p))
  eta <- (friedman(x) - 14) / 5
  list(x = x, eta = eta, train = seq_len(n), test = seq_len(n) + n)
}
```

Timing and mixing helpers. `ess` is the usual initial-positive-sequence
estimate; `ess_of` summarizes it over the observations by its median, since there
is one per observation.

```{r helpers}
# Times the best of `reps` runs. A fit that errors -- a package that is
# installed but cannot run here, say -- returns NULL rather than stopping the
# whole document.
timer <- function(expr) {
  best <- Inf
  value <- NULL

  for (i in seq_len(reps)) {
    ok <- TRUE
    t <- system.time(
      value <- tryCatch(eval.parent(substitute(expr)),
                        error = function(e) {
                          ok <<- FALSE
                          message("skipping: ", conditionMessage(e))
                          NULL
                        }))[["elapsed"]]
    if (!ok) return(list(seconds = NA_real_, value = NULL))
    best <- min(best, t)
  }

  list(seconds = best, value = value)
}

ess <- function(z) {
  v <- stats::var(z)
  if (!(v > 0)) return(NA_real_)
  r <- stats::acf(z, plot = FALSE, lag.max = min(200, length(z) - 1))$acf[-1]
  cut <- which(r < 0.05)[1]
  if (is.na(cut)) cut <- length(r)
  length(z) / (1 + 2 * sum(r[seq_len(cut)]))
}

ess_of <- function(draws_matrix) {
  stats::median(apply(draws_matrix, 2L, ess), na.rm = TRUE)
}

results <- list()

# A fit that failed is recorded as a row of NAs rather than dropped. Dropping it
# is how bartMachine went missing from a whole run: its fits were erroring, the
# timer returned NA, and the table simply had no bartMachine in it, which reads
# as "not installed" rather than "installed and broken".
skipped <- list()

record <- function(task, package, call, seconds, draws_matrix, point, truth,
                   score = NA_real_) {
  if (is.na(seconds)) {
    skipped[[length(skipped) + 1L]] <<- data.frame(task = task,
                                                   package = package,
                                                   call = call)
    return(invisible(NULL))
  }

  e <- if (is.null(draws_matrix)) NA_real_ else ess_of(draws_matrix)
  results[[length(results) + 1L]] <<- data.frame(
    task = task, package = package, call = call,
    seconds = seconds,
    ess = e,
    ess_per_sec = e / seconds,
    rmse = sqrt(mean((point - truth)^2)),
    log_score = score)
  invisible(NULL)
}
```

## Gaussian response

The case every package supports, so it is the cleanest comparison of the
samplers themselves.

```{r gaussian}
d <- make_data(n, p, seed = 12)
y <- d$eta + rnorm(2 * n, sd = 0.5)
xtr <- d$x[d$train, , drop = FALSE]; ytr <- y[d$train]
xte <- d$x[d$test, , drop = FALSE];  truth <- d$eta[d$test]
dtr <- data.frame(xtr, y = ytr); dte <- data.frame(xte)

ctrl <- function(...) {
  bartisan_control(num_trees = trees, num_burn = draws, num_draws = draws,
                   verbose = FALSE, ...)
}

# bartisan. `gate` is the one argument that chooses between hard and soft rules
# and, when soft, the shape of the gate; "smoothstep" is the default, and its
# support is bounded, so an observation past it takes one side of a rule outright
# and the subtree on the other side is never visited. See ?bartisan_control.
r <- timer(bartisan(y ~ ., data = dtr, control = ctrl()))
e <- predict(r$value, newdata = dte, type = "link", draws = TRUE)
record("gaussian", "bartisan", "soft, smoothstep gate (default)", r$seconds, e,
       colMeans(e), truth)

r <- timer(bartisan(y ~ ., data = dtr, control = ctrl(gate = "hard")))
e <- predict(r$value, newdata = dte, type = "link", draws = TRUE)
record("gaussian", "bartisan", "hard rules", r$seconds, e, colMeans(e), truth)

for (g in c("smootherstep", "logistic")) {
  r <- timer(bartisan(y ~ ., data = dtr, control = ctrl(gate = g)))
  e <- predict(r$value, newdata = dte, type = "link", draws = TRUE)
  record("gaussian", "bartisan", paste("soft,", g, "gate"), r$seconds, e,
         colMeans(e), truth)
}

# dbarts.
r <- timer(bart(xtr, ytr, xte, ntree = trees, nskip = draws, ndpost = draws,
                verbose = FALSE, keeptrees = FALSE))
record("gaussian", "dbarts", "bart()", r$seconds, r$value$yhat.test,
       colMeans(r$value$yhat.test), truth)

if (installed[["BART"]]) {
  r <- timer(BART::wbart(xtr, ytr, xte, ntree = trees, nskip = draws,
                         ndpost = draws, printevery = draws + 1L))
  if (!is.null(r$value)) {
    record("gaussian", "BART", "wbart()", r$seconds, r$value$yhat.test,
           colMeans(r$value$yhat.test), truth)
  }
}

# stochtree. `num_gfr` is its grow-from-root warm start (He and Hahn 2021),
# which is a different algorithm from the MCMC every other package here runs, so
# it is timed both ways: at 0 for a like-for-like comparison of samplers, and at
# its default for what the package actually recommends.
if (installed[["stochtree"]]) {
  for (gfr in c(0L, 5L)) {
    r <- timer(stochtree::bart(X_train = xtr, y_train = ytr, X_test = xte,
                               num_gfr = gfr, num_burnin = draws,
                               num_mcmc = draws,
                               mean_forest_params = list(num_trees = trees)))
    if (!is.null(r$value)) {
      e <- t(r$value$y_hat_test)
      record("gaussian", "stochtree",
             sprintf("bart(num_gfr = %d)", gfr), r$seconds, e, colMeans(e),
             truth)
    }
  }
}

if (installed[["bartMachine"]]) {
  r <- timer(bartMachine::bartMachine(as.data.frame(xtr), ytr,
                                      num_trees = trees,
                                      num_burn_in = draws,
                                      num_iterations_after_burn_in = draws,
                                      verbose = FALSE, seed = 1))
  if (!is.null(r$value)) {
    point <- bartMachine::bart_predict_for_test_data(r$value,
                                                     as.data.frame(xte),
                                                     truth)$y_hat
    record("gaussian", "bartMachine", "bartMachine()", r$seconds, NULL, point,
           truth)
  }
}
```

## Binary response, probit link

```{r probit}
prob <- pnorm(d$eta)
yb <- rbinom(2 * n, 1, prob)
dtr_b <- data.frame(xtr, y = yb[d$train])
truth_b <- prob[d$test]

# `augment` is on by default, so the comparison needs it switched off
# explicitly. See ?bartisan_control for what it does and what it costs.
r <- timer(bartisan(y ~ ., data = dtr_b, family = binomial("probit"),
                   control = ctrl(augment = FALSE)))
e <- predict(r$value, newdata = dte, type = "response", draws = TRUE)
record("probit", "bartisan", "soft, augment = FALSE", r$seconds, e,
       colMeans(e), truth_b)

r <- timer(bartisan(y ~ ., data = dtr_b, family = binomial("probit"),
                   control = ctrl(augment = TRUE)))
e <- predict(r$value, newdata = dte, type = "response", draws = TRUE)
record("probit", "bartisan", "soft, augment = TRUE", r$seconds, e, colMeans(e),
       truth_b)

r <- timer(bartisan(y ~ ., data = dtr_b, family = binomial("probit"),
                   control = ctrl(gate = "hard", augment = TRUE)))
e <- predict(r$value, newdata = dte, type = "response", draws = TRUE)
record("probit", "bartisan", "hard, augment = TRUE", r$seconds, e, colMeans(e),
       truth_b)

if (installed[["stochtree"]]) {
  r <- timer(stochtree::bart(
    X_train = xtr, y_train = yb[d$train], X_test = xte, num_gfr = 0L,
    num_burnin = draws, num_mcmc = draws,
    general_params = list(outcome_model = stochtree::OutcomeModel(
      outcome = "binary", link = "probit")),
    mean_forest_params = list(num_trees = trees)))
  if (!is.null(r$value)) {
    e <- pnorm(t(r$value$y_hat_test))
    record("probit", "stochtree", "bart(binary/probit)", r$seconds, e,
           colMeans(e), truth_b)
  }
}

r <- timer(bart(xtr, factor(yb[d$train]), xte, ntree = trees, nskip = draws,
                ndpost = draws, verbose = FALSE, keeptrees = FALSE))
e <- pnorm(r$value$yhat.test)
record("probit", "dbarts", "bart()", r$seconds, e, colMeans(e), truth_b)

if (installed[["BART"]]) {
  r <- timer(BART::pbart(xtr, yb[d$train], xte, ntree = trees, nskip = draws,
                         ndpost = draws, printevery = draws + 1L))
  if (!is.null(r$value)) {
    record("probit", "BART", "pbart()", r$seconds, r$value$prob.test,
           colMeans(r$value$prob.test), truth_b)
  }
}
```

## Binary response, logit link

Only bartisan and `BART::lbart` offer this; dbarts is probit-only. The two use
different augmentations -- Pólya-Gamma here, the Holmes and Held (2006) scale
mixture in `lbart` -- so this is also a comparison of those.

```{r logit}
prob_l <- plogis(d$eta)
yl <- rbinom(2 * n, 1, prob_l)
dtr_l <- data.frame(xtr, y = yl[d$train])
truth_l <- prob_l[d$test]

for (aug in c(FALSE, TRUE)) {
  r <- timer(bartisan(y ~ ., data = dtr_l, family = binomial("logit"),
                     control = ctrl(augment = aug)))
  e <- predict(r$value, newdata = dte, type = "response", draws = TRUE)
  record("logit", "bartisan", paste0("soft, augment = ", aug), r$seconds, e,
         colMeans(e), truth_l)
}

if (installed[["BART"]]) {
  r <- timer(BART::lbart(xtr, yl[d$train], xte, ntree = trees, nskip = draws,
                         ndpost = draws, printevery = draws + 1L))
  if (!is.null(r$value)) {
    record("logit", "BART", "lbart()", r$seconds, r$value$prob.test,
           colMeans(r$value$prob.test), truth_l)
  }
}
```

## Families only bartisan has

These have no counterpart in the other packages, so there is nothing to compare
against on speed. What the column shows is the cost of the general machinery when
no Gaussian rewriting is available -- which is the price of the coverage.

```{r general}
# The gamma, Poisson and negative binomial families reach the *exponential* form
# rather than the quadratic one -- see ?bartisan_control -- which under hard rules
# collapses the leaf work to one pass just as the quadratic form does. The
# `exact_quadratic = FALSE` runs show what that is worth.
dtr_g <- data.frame(xtr, y = rgamma(n, shape = 4, rate = 4 / exp(d$eta[d$train])))
for (exact in c(TRUE, FALSE)) {
  r <- timer(bartisan(y ~ ., data = dtr_g, family = Gamma("log"),
                     control = ctrl(gate = "hard", exact_quadratic = exact)))
  e <- predict(r$value, newdata = dte, type = "link", draws = TRUE)
  record("gamma (log link)", "bartisan",
         if (exact) "hard rules" else "hard, no shortcut", r$seconds, e,
         colMeans(e), truth)
}

dtr_p <- data.frame(xtr, y = rpois(n, exp(d$eta[d$train] + 1)))
for (exact in c(TRUE, FALSE)) {
  r <- timer(bartisan(y ~ ., data = dtr_p, family = poisson(),
                     control = ctrl(gate = "hard", exact_quadratic = exact)))
  e <- predict(r$value, newdata = dte, type = "link", draws = TRUE)
  record("poisson", "bartisan",
         if (exact) "hard rules" else "hard, no shortcut", r$seconds, e,
         colMeans(e), truth + 1)
}

dtr_nb <- data.frame(xtr, y = rnbinom(n, mu = exp(d$eta[d$train] + 1), size = 2))
for (aug in c(TRUE, FALSE)) {
  r <- timer(bartisan(y ~ ., data = dtr_nb, family = negbin(),
                     control = ctrl(gate = "hard", augment = aug)))
  e <- predict(r$value, newdata = dte, type = "link", draws = TRUE)
  record("negative binomial", "bartisan",
         if (aug) "hard, augment = TRUE" else "hard, augment = FALSE",
         r$seconds, e, colMeans(e), truth + 1)
}

lt <- d$eta[d$train] + 0.7 * rlogis(n)
cens <- quantile(lt, 0.75)
dtr_s <- data.frame(xtr, time = exp(pmin(lt, cens)),
                    status = as.numeric(lt <= cens))
r <- timer(bartisan(survival::Surv(time, status) ~ ., data = dtr_s,
                   family = loglogistic_aft(), control = ctrl(gate = "hard")))
e <- predict(r$value, newdata = dte, type = "link", draws = TRUE)
record("log-logistic AFT", "bartisan", "hard rules", r$seconds, e, colMeans(e),
       truth)

cuts <- c(-1, 0.5, 2)
probs <- vapply(d$eta[d$train], function(z) diff(c(0, pnorm(cuts - z), 1)),
                numeric(4))
dtr_o <- data.frame(xtr, y = factor(apply(probs, 2L, function(q)
  sample.int(4, 1, prob = q)), levels = 1:4, ordered = TRUE))

# Both links have a latent-variable representation now, so both are timed with
# the rewriting on and off; the gap between the pair of rows for a link is what
# the augmentation is worth there.
for (setting in list(list(lab = "hard, probit, augment = TRUE",
                          fam = ordinal("probit"), gate = "hard",
                          augment = "ordinal"),
                     list(lab = "soft, probit, augment = TRUE",
                          fam = ordinal("probit"), gate = "smoothstep",
                          augment = "ordinal"),
                     list(lab = "hard, probit, augment = FALSE",
                          fam = ordinal("probit"), gate = "hard",
                          augment = FALSE),
                     list(lab = "hard, logit, augment = TRUE",
                          fam = ordinal("logit"), gate = "hard",
                          augment = "ordinal"),
                     list(lab = "soft, logit, augment = TRUE",
                          fam = ordinal("logit"), gate = "smoothstep",
                          augment = "ordinal"),
                     list(lab = "hard, cloglog, augment = TRUE",
                          fam = ordinal("cloglog"), gate = "hard",
                          augment = "ordinal"),
                     list(lab = "hard, logit, augment = FALSE",
                          fam = ordinal("logit"), gate = "hard",
                          augment = FALSE),
                     list(lab = "hard, cloglog, augment = FALSE",
                          fam = ordinal("cloglog"), gate = "hard",
                          augment = FALSE))) {
  r <- timer(bartisan(y ~ ., data = dtr_o, family = setting$fam,
                      control = ctrl(gate = setting$gate,
                                     augment = setting$augment)))
  e <- predict(r$value, newdata = dte, type = "link", draws = TRUE)
  record("ordinal", "bartisan", setting$lab, r$seconds, e, colMeans(e), truth)
}

# stochtree fits ordinal responses through a complementary log-log link, which
# bartisan now also offers, so the `cloglog` rows above are the like-for-like
# comparison. The RMSE column still is not comparable across links.
if (installed[["stochtree"]]) {
  r <- timer(stochtree::bart(
    X_train = xtr, y_train = as.integer(dtr_o$y), X_test = xte, num_gfr = 0L,
    num_burnin = draws, num_mcmc = draws,
    general_params = list(outcome_model = stochtree::OutcomeModel(
      outcome = "ordinal", link = "cloglog")),
    mean_forest_params = list(num_trees = trees)))
  if (!is.null(r$value)) {
    e <- t(r$value$y_hat_test)
    record("ordinal", "stochtree", "bart(ordinal/cloglog)", r$seconds, e,
           colMeans(e), truth)
  }
}
```

## How many trees, and how many per forest

Two sweeps that inform the defaults rather than comparing packages. The first is
the tree count under each kind of rule, on the same Gaussian task as above: soft
rules reach their best held-out error at far fewer trees than hard ones, and past
that point they get worse. The second is the per-forest vector, on
`location_scale()`, where the scale forest is where nearly all the time goes.

```{r tree-count}
tree_sweep <- do.call(rbind, lapply(c("smoothstep", "hard"), function(g) {
  do.call(rbind, lapply(c(10L, 20L, 50L, 100L, 200L), function(nt) {
    r <- timer(bartisan(y ~ ., data = dtr, family = gaussian(),
                       control = bartisan_control(num_trees = nt, gate = g,
                                                  num_burn = draws,
                                                 num_draws = draws,
                                                 verbose = FALSE)))
    e <- predict(r$value, newdata = dte, type = "link")
    data.frame(gate = g, trees = nt, seconds = round(r$seconds, 2),
               rmse = round(sqrt(mean((e - truth)^2)), 3))
  }))
}))

tree_sweep
```

```{r per-forest}
dtr_ls <- dtr
dtr_ls$y <- dtr$y + rnorm(nrow(dtr), 0, exp(-1 + dtr$x2))

per_forest <- do.call(rbind, lapply(list(c(trees, trees), c(trees, 10L),
                                         c(trees, 5L)), function(nt) {
  r <- timer(bartisan(y ~ ., data = dtr_ls, family = location_scale(),
                     control = bartisan_control(num_trees = nt,
                                                num_burn = draws,
                                               num_draws = draws,
                                               verbose = FALSE)))
  e <- predict(r$value, newdata = dte, type = "link")
  data.frame(num_trees = paste(nt, collapse = ","),
             seconds = round(r$seconds, 2),
             mean_rmse = round(sqrt(mean((e[, 1] - truth)^2)), 3))
}))

per_forest
```

## Several chains at once

The chain is the only parallel axis a sampler like this has, and it is what makes
a convergence diagnostic possible. Any `future` backend works, mirai's included.

```{r chains}
cores <- suppressWarnings(parallel::detectCores())
workers <- if (is.finite(cores)) max(1L, min(4L, cores - 1L)) else 2L

# A backend that cannot start -- no sockets, no forking, a locked-down
# environment -- should not take the document down with it.
started <- FALSE

if (have("future.apply") && have("future") && workers > 1L) {
  started <- tryCatch({
    future::plan(future::multisession, workers = workers)
    TRUE
  }, error = function(e) {
    message("multisession unavailable (", conditionMessage(e),
            "); trying multicore")
    tryCatch({
      future::plan(future::multicore, workers = workers)
      TRUE
    }, error = function(e2) {
      message("no parallel backend available; timing one chain at a time")
      FALSE
    })
  })
}

if (started) {
  chain_draws <- max(50L, draws %/% workers)

  r <- timer(bartisan(y ~ ., data = dtr, chains = workers,
                      control = bartisan_control(num_trees = trees,
                                                 num_burn = draws,
                                               num_draws = chain_draws,
                                               verbose = FALSE)))

  if (!is.null(r$value)) {
    cat(sprintf("%d chains of %d draws on %d workers: %.2f s\n", workers,
                chain_draws, workers, r$seconds))
    # Rank-normalized folded split R-hat, and the bulk and tail effective sample
    # sizes. R-hat above about 1.01, or an effective sample size below about 400,
    # says the run is too short for the quantity on that row -- and the tail
    # column is the one that governs an interval endpoint.
    print(r$value$rhat, digits = 4)
  }

  future::plan(future::sequential)
}
```

## Results

```{r table}
out <- do.call(rbind, results)
out <- out[order(out$task, out$seconds), ]
out$relative <- NA_real_

for (task in unique(out$task)) {
  rows <- out$task == task
  out$relative[rows] <- out$seconds[rows] / min(out$seconds[rows])
}

knitr::kable(out, digits = c(0, 0, 0, 3, 1, 1, 4, 3, 2), row.names = FALSE)
```

`seconds` is the best of `r reps` runs. `ess` is the median over test
observations of the effective sample size of the fitted value, out of
`r draws` draws; `ess_per_sec` divides one by the other and is the number to
compare. `rmse` is against the true regression function on the held-out half.
`relative` is the time divided by the fastest fit for that task.

Two things the `rmse` column will not bear:

* It is against the additive predictor on the scale the data were generated on,
  which for the ordinal task is the probit scale. The `ordinal` rows for the
  logit link and for stochtree's complementary log-log link are therefore
  measuring a scale difference, not a worse fit, and only the probit rows are
  comparable with each other. The `seconds` and `ess` columns are comparable
  across all of them.
* bartMachine reports no posterior draws of the fitted value through
  `bart_predict_for_test_data()`, so its `ess` is missing and its
  `ess_per_sec` with it. Its timing is comparable; its mixing is not measured.

Any fit that was attempted and failed is listed here rather than being left out
of the table above, so that "installed but not working" cannot be mistaken for
"not installed".

```{r skipped}
if (length(skipped)) {
  knitr::kable(do.call(rbind, skipped), row.names = FALSE)
} else {
  "none"
}
```

```{r sessioninfo}
sessionInfo()
```
